---
title: 'Supplementary material of "A graphical approach for multivariate modeling" by'
author: "Anna Freni-Sterrantino, Denis Rustand, Janet van Niekerk, Elias T Krainski, and H\\aa vard Rue"
date: "November, 2023"
output:
  pdf_document: 
    keep_tex: yes
  html_document:
    df_print: paged
header-includes:
- \usepackage{lscape}
- \usepackage{multicol}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(graphpcor)
```


In this material we detail the model derivation and properties of the 
DAG approach used to specify the relationship 
between the modeled variables through latent variables.
We introduce the model definition 
allowing the full range of the correlation values, from $-1$ to $1$.
The main purpose is to model the correlation matrix between $n$ 
target variables considering some structure made from 
model assumptions on the variables relationship. 
This gives an structured correlation matrix, where certain 
pairs of variables have the same (in absolute value) correlation. 
This reduces the reduced number of parameters, from
$n(n-1)/2$ (for unstructured correlation matrices)
to $p$, where $p$ is usually lower than $n$.

The assumptions for the structure of the correlations are represented
on an augmented direct acyclic graph - DAG.
This specific DAG include assumption on the relations
between the target variables being modeled through a set 
of $p$ latent variables.
The model, or the joint distribution for all the variables,
is defined through a set of conditional distributions,
which follows from the assumed DAG structure and two simple rules.
These conditional distributions defines the precision matrix 
which determines a covariance matrix for all the variables.

The $n$ variables whose dependency are being modeled can be 
observation variables or to consider the relationship 
in a set of model parameters in a complex model.
We start by establishing relations between the variables 
through a specific DAG and the the model specification.
This specifies the covariance structure which has 
properties listed at the end of this section.

## The DAG definition

Our approach does not consider directly a DAG for the $n$ variables.
We start by modeling considering $p$ latent variables and its 
relationship and how the $n$ target variables share these latent ones.
As usual in a DAG, the $n+p$ variables are nodes and 
the arrows represent the relation between them.
In our specific kind of DAG the target variables are 
children and the latent variables are parents.
A parent can have a parent but a children cannot have children.
The covariance between the children is derived from the
relationship between the parents and how they are shared by the variables. 
Let a set of $n$ children, $c_i$, $i$=1,...,$n$, 
and $p$ parents, $p_j$, $j=1,...,p$.
We have some examples in Figure 1.

```{r graphs, echo = FALSE, fig.width = 9, fig.height = 4, width="99%", fig.cap = "Examples of DAG: one latent variable (A), two latent varialbes (B), and three latent varialbes (C) and (D)."}
g1 <- list(p1 ~ c1 + c2 + c3)
g1p <- GraphPlot(g1)
g2 <- list(p1 ~ p2 + c3 + c4, p2 ~ c1 + c2)
g2p <- GraphPlot(g2)
g3a <- list(p1 ~ p2 + p3 + c1 + c2, p2 ~ c3 + c4, p3 ~ c5 + c6)
g3b <- list(p1 ~ p2 + c1 + c2, p2 ~ p3 + c3 + c4, p3 ~ c5 + c6)
g3ap <- GraphPlot(g3a)
g3bp <- GraphPlot(g3b)
par(mfrow = c(2, 2), mar = c(0, 0, 0, 0))
plot(g1p$gr, nodeAttrs = g1p$nAttrs)
legend("topleft", "(A)", bty = "n")
plot(g2p$gr, nodeAttrs = g2p$nAttrs)
legend("topleft", "(B)", bty = "n")
plot(g3ap$gr, nodeAttrs = g3ap$nAttrs)
legend("topleft", "(C)", bty = "n")
plot(g3bp$gr, nodeAttrs = g3bp$nAttrs)
legend("topleft", "(D)", bty = "n")
```

The model is now defined from a DAG and is used to impose
constraints in the correlation structure between all the variables.
In general, a graph with less parents will have less constraint.
The $n=3$ children variables in the graph (A) have 
the same (absolute) correlation between them
as they share the same parent.
We will point some properties from the graphs in Figure 1
and will list a set of properties later on.
Let C(a,b) be the correlation between the variables $a$ and $b$.
In the model defined with graph (A), 
$|$C($c_1$,$c_2$)$|$=$|$C($c_1$,$c_3$)$|$=$|$C($c_2$,$c_3$)$|$.
In the model defined with graph (B), 
$|$C($c_1$,$c_3$)$|$=$|$C($c_1$,$c_4$)$|$=
$|$C($c_2$,$c_3$)$|$=$|$C($c_2$,$c_4$)$|$,
$|$C($c_1$,$c_3$)$|\leq|$C($c_1$,$c_2$)$|$, and 
$|$C($c_1$,$c_3$)$|\leq|$C($c_3$,$c_4$)$|$.
Both the graphs (C) and (D) have three latent variables 
with $p_1$ as the main node. 
However, in the sense of a path between parent nodes, 
$p_1$ is a central parent in (C), but not in (D), where $p_2$ is.
This changes the pair correlation order, 
for example, there are three latent variables in the path between 
$c_1$ and $c_5$ in (C) but two in (D).

In the DAG represented in (A), if we add more children, 
the (absolute) correlation between it and any other children is the same. 
Thus with $p=1$ we have only one (absolute) correlation value.
In the DAG represented in (B), we can move one children to be children 
of the other parent changing the correlation model. 
This changes affects the correlation between this children 
and all the others, but still preserves the same number of 
(absolute) different correlation values, as we keep the number of parents.
We will list the model properties at the end of this section.

## Model definition

From the DAG defined for the $n$ children and $p$ parents we have
$n+p$ variables and a set of direct relations. 
We use it along with simple rules to define the joint model. 
This model definition starts assuming a distribution for the 
main parent and then a set of conditional distributions for 
the $n+p-1$ other variables. 
We assume Gaussian distributions.

* **1st rule**: For the main parent, $p_1$ the one without a parent, 
the mean is zero and the variance is $b_1^2$. 

The $p+n-1$ conditional distributions are defined following the 
cycles starting from the main parent. 
We define a weight parameter for each arrow in the DAG.
If we have to set the conditional distribution of $p_j$, 
$j>1$ as $p_1$ is already defined,
this weight parameter is named $b_j$.
If it is for a children variable $c_i$, 
this weight parameter is named as $a_i$.

The number of conditional distributions is equal 
the number of arrows in the DAG.
We start from each children of $p_1$, in any order.
Then to the next level, if any, and follows each cycle, 
defining the a conditional distribution for the variable after 
each arrow until the end, 
and until all and each variable conditional distribution is defined.

* **2nd rule**: The conditional mean is the weight times the parent name
and the conditional variance is the weight squared. 

As an illustration, consider the graph (A) from Figure 1.
The joint distribution for $\{c_1, c_2, c_3, p_1\}$ is 
\[
\pi(c_1, c_2, c_3, p_1) \propto \mathrm{exp} 
  \left\{-\frac{1}{2} \left[ \frac{p_1^2}{b_1^2} + \frac{(c_1-a_1p_1)^2}{a_1^2} + \frac{(c_2-a_2p_1)^2}{a_2^2} + \frac{(c_3-a_3p_1)^2}{a_3^2} \right] \right\} 
\]

The joint distribution for $\{c_1, c_2, c_3, c_4, p_1, p_2\}$, 
from the DAG (B) in Figure 1, is proportional to
\[ 
\mathrm{exp} 
  \left\{-\frac{1}{2} \left[ \frac{p_1^2}{b_1^2} + \frac{(p_2-b_2p_1)^2}{b_2^2} +
  \frac{(c_1-a_1p_1)^2}{a_1^2} + \frac{(c_2-a_2p_1)^2}{a_2^2} +
  \frac{(c_3-a_3p_2)^2}{a_3^2} + \frac{(c_4-a_4p_2)^2}{a_4^2} \right] \right\} 
\]

## Model properties

Our primarily interest is not in the correlation between 
a target variable (children) and a latent variable (parent), 
but between the children ones. 
The joint distribution defined from the DAG and the simple rules,
is used to define the precision matrix for the $n+p$ variables.
From this, we obtain the covariance and look at its first $n$ rows and columns
to learn the marginal properties and the desired correlation.
We will start with examples and them list the model properties. 

The model from the DAG (A) in Figure 1 for 
$\{c_1, c_2, c_3, p_1\}$ 
has precision and covariance matrices equals to 
\[ 
\left[ \begin{array}{cccc} 
 {a_1}^{-2}&0&0&-{a_1}^{-1} \\ 
 0&{a_2}^{-2}&0&-{a_2}^{-1}\\ 
 0&0&{a_3}^{-2}&-{a_3}^{-1} \\
 -{a_1}^{-1}&-{a_2}^{-1}&-{a_3}^{-1}&{b_1}^{-2}+3 
\end{array} \right] 
\textrm{and}\; 
\left[ \begin {array}{cccc}  
( 1+{b_1}^{2} ) {a_1}^{2}&a_1a_2{b_1}^{2}&a_1a_3{b_1}^{2}&a_3{b_1}^{2}\\ 
 a_1a_2{b_1}^{2}& ( 1+{b_1}^{2} ) {a_2}^{2} & a_2a_3{b_1}^{2} & a_2{b_1}^{2} \\ 
a_1a_3{b_1}^{2} & a_2a_3{b_1}^{2}& ( 1+{b_1}^{2} ) {a_3}^{2}&a_3{b_1}^{2} \\
a_1{b_1}^{2} & a_2{b_1}^{2} & a_3{b_1}^{2} & {b_1}^{2}\end {array} \right] 
\]
respectively. 
The correlation is $b_1^2/(1+b_1^2)$ times the 
sign of the product between the weights.

The model defined from the DAG (B) in Figure 1, 
has $n=4$ and $p=2$ for the variables $\{c_1, c_2, c_3, c_4, p_1, p_2\}$.
The key rows of the precision matrix are the last $p$ ones wich are
\[ 
\left[ \begin {array}{cccccc} 
%%{{\it a_1}}^{-2}&0&0&0&-{{\it a_1}}^{-1}&0\\ 
%%0&{{\it a_2}}^{-2}&0&0&-{{\it a_2}}^{-1}&0\\ 
%% 0&0&{{\it a_3}}^{-2}&0&0&-{{\it a_3}}^{-1}\\ 
%%0&0&0&{{\it a_4}}^{-2}&0&-{{\it a_4}}^{-1}\\ 
-{{\it a_1}}^{-1}&-{{\it a_2}}^{-1}&0&0&{{\it b_1}}^{-2}+3&-{{\it b_2}}^{-1}\\ 
0&0&-{{\it a_3}}^{-1}&-{{\it a_4}}^{-1}&-{{\it b_2}}^{-1}&{{\it b_2}}^{-2}+2 
\end {array} \right] 
\]
and the covariance matrix for the children variables is
\[ 
\left[ \begin {array}{cccc} 
( 1+{{\it b_1}}^{2} ) {{\it a_1}}^{2}&
{\it a_2}\,{{\it b_1}}^{2}{\it a_1}&
{\it a_3}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_1}&
{\it a_4}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_1}\\ 
{\it a_2}\,{{\it b_1}}^{2}{\it a_1} & 
( 1+{{\it b_1}}^{2} ) {{\it a_2}}^{2} & 
{\it a_3}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_2} & 
{\it a_4}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_2}\\
{\it a_3}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_1} & 
{\it a_3}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_2} & 
[ 1+{{\it b_2}}^{2}(1+{{\it b_1}}^{2}) ] {{\it a_3}}^{2} & 
{\it a_4}\,{\it a_3}\,{{\it b_2}}^{2} ( 1+{{\it b_1}}^{2} ) \\ 
{\it a_4}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_1} & 
{\it a_4}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_2} & 
{\it a_4}\,{\it a_3}\,{{\it b_2}}^{2} ( 1 + {{\it b_1}}^{2} ) & 
[ 1+{{\it b_2}}^{2}(1+{{\it b_1}}^{2})] {{\it a_4}}^{2}
\end {array} \right] .
\]
We can learn the properties from this matrix 
and will list some later.
The upper triangle of the correlation matrix 
between the observation variables is
\[
\left[ \begin {array}{ccc} 
\mathbf{C}_{1,2} = 
 \frac{\textrm{sign}(a_1a_2){b_2}^{2}(1+{b_1}^{2})}{1+{b_2}^{2}(1+{b_1}^{2})} &
\mathbf{C}_{1,3} = \frac{\textrm{sign}(a_1a_2b_2){b_1}^{2}}
 {\sqrt{1+{b_2}^{2}(1+{b_1}^{2})}\sqrt {1+{b_1}^{2}}} &
\mathbf{C}_{1,4} = \frac{\textrm{sign}(a_1a_4b_2){b_1}^{2}}
  {\sqrt{1+{b_2}^{2}(1+{b_1}^{2})}\sqrt{1+{b_1}^{2}}} \\ 
 & \mathbf{C}_{2,3} = \frac {\textrm{sign}(a_2a_3b_2){b_1}^{2}}
  {\sqrt{ 1+{b_2}^{2}(1+{b_1}^{2}) }\sqrt { 1+{b_1}^{2} }} &
 \mathbf{C}_{2,4} = \frac{\textrm{sign}(a_2a_4b_2){b_1}^{2}}
  {\sqrt{1+{b_2}^{2}(1+{b_1}^{2})} \sqrt{1+{b_1}^{2} }} \\ 
 && \mathbf{C}_{3,4} = \frac {\textrm{sign}(a_3a_4){b_1}^{2}}{ 1+{b_1}^{2} }\\ 
\end {array} \right].
\]

In the models given from graphs (C) and (D) in Figure 1, 
$n=6$ and $p=3$, each latent variable has two children.
The only difference is who is parent to $p_3$.

For the case when $p_1$ is parent to $p_3$ we have the last three rows of 
the precision matrix as 
\[
 \left[ \begin {array}{ccccccccc} -{{\it a_1}}^{-1}&-{{\it a_2}}^{-1}&0&0&0&0&{{\it b_1}}^{-2}+4&-{{\it b_2}}^{-1}&-{{\it b_3}}^{-1}
\\ 0&0&-{{\it a_3}}^{-1}&-{{\it a_4}}^{-1}&0&0&-{{\it b_2}}^{-1}&{{\it b_2}}^{-2}+2&0\\ 0&0&0&0&-{{\it a_5}}
^{-1}&-{{\it a_6}}^{-1}&-{{\it b_3}}^{-1}&0&{{\it b_3}}^{-2}+2\end {array} \right] 
\]
and the covariance for $\{c_1,...,c_6\}$ is
\[ \tiny
 \left[ \begin {array}{cccccc}  
 ( 1+{{\it b_1}}^{2} ) {{\it a_1}}^{2} & 
 {\it a_2}\,{{\it b_1}}^{2}{\it a_1} & 
 {\it a_3}\,{\it a_1}\,{{\it b_1}}^{2}{\it b_2} & 
 {\it a_4}\,{\it a_1}\,{{\it b_1}}^{2}{\it b_2} &
 {\it a_5}\,{{\it b_1}}^{2}{\it a_1}\,{\it b_3} &
 {\it a_6}\,{{\it b_1}}^{2}{\it a_1}\,{\it b_3}\\
 {\it a_2}\,{{\it b_1}}^{2}{\it a_1} &
 ( 1+{{\it b_1}}^{2} ) {{\it a_2}}^{2} &
 {\it a_3}\,{\it a_2}\,{{\it b_1}}^{2}{\it b_2}&
 {\it a_4}\,{\it a_2}\,{{\it b_1}}^{2}{\it b_2}&
 {\it a_5}\,{{\it b_1}}^{2}{\it a_2}\,{\it b_3}&
 {\it a_6}\,{{\it b_1}}^{2}{\it a_2}\,{\it b_3}\\ 
 {\it a_3}\,{\it a_1}\,{{\it b_1}}^{2}{\it b_2}&
 {\it a_3}\,{\it a_2}\,{{\it b_1}}^{2}{\it b_2}& 
 [ 1+{{\it b_2}}^{2}(1+{{\it b_1}}^{2})] {{\it a_3}}^{2}&
 {\it a_4}\, ( 1+{{\it b_1}}^{2}) {\it a_3}\,{{\it b_2}}^{2}&
 {\it a_5}\,{\it b_2}\,{{\it b_1}}^{2}{\it b_3}\,{\it a_3}&
 {\it a_6}\,{\it b_2}\,{{\it b_1}}^{2}{\it b_3}\,{\it a_3}\\ 
 {\it a_4}\,{\it a_1}\,{{\it b_1}}^{2}{\it b_2}&
 {\it a_4}\,{\it a_2}\,{{\it b_1}}^{2}{\it b_2}&
 {\it a_4}\, ( 1+{{\it b_1}}^{2} ) {\it a_3}\,{{\it b_2}}^{2}& 
 [ 1+{{\it b_2}}^{2}(1+{{\it b_1}}^{2})] {{\it a_4}}^{2}&
 {\it a_5}\,{\it b_2}\,{{\it b_1}}^{2}{\it b_3}\,{\it a_4}&
 {\it a_6}\,{\it b_2}\,{{\it b_1}}^{2}{\it b_3}\,{\it a_4}\\ 
{\it a_5}\,{{\it b_1}}^{2}{\it a_1}\,{\it b_3}&
{\it a_5}\,{{\it b_1}}^{2}{\it a_2}\,{\it b_3}&
{\it a_5}\,{\it b_2}\,{{\it b_1}}^{2}{\it b_3}\,{\it a_3}&
{\it a_5}\,{\it b_2}\,{{\it b_1}}^{2}{\it b_3}\,{\it a_4}& 
 [ 1+{{\it b_3}}^{2}(1+{{\it b_1}}^{2} )]{{\it a_5}}^{2}&
 {\it a_6}\,{\it a_5}\, ( 1+{{\it b_1}}^{2} ) {{\it b_3}}^{2}\\ 
{\it a_6}\,{{\it b_1}}^{2}{\it a_1}\,{\it b_3}&
{\it a_6}\,{{\it b_1}}^{2}{\it a_2}\,{\it b_3}&
{\it a_6}\,{\it b_2}\,{{\it b_1}}^{2}{\it b_3}\,{\it a_3}&
{\it a_6}\,{\it b_2}\,{{\it b_1}}^{2}{\it b_3}\,{\it a_4}&
{\it a_6}\,{\it a_5}\, ( 1+{{\it b_1}}^{2} ) {{\it b_3}}^{2}& 
 [ 1+{{\it b_3}}^{2}(1+{{\it b_1}}^{2}) ] {{\it a_6}}^{2}
\end {array} \right]
\]

For the case when $p_2$ is parent to $p_3$ we have the last three rows of 
the precision matrix as 
\[
\left[ \begin {array}{ccccccccc} 
-{{\it a_1}}^{-1}&-{{\it a_2}}^{-1}&0&0&0&0&{{\it b_1}}^{-2}+3&-{{\it b_2}}^{-1}&0\\ 
0&0&-{{\it a_3}}^{-1}&-{{\it a_4}}^{-1}&0&0&-{{\it b_2}}^{-1}&{{\it b_2}}^{-2}+3&-{{\it b_3}}^{-1}\\ 
0&0&0&0&-{{\it a_5}}^{-1}&-{{\it a_6}}^{-1}&0&-{{\it b_3}}^{-1}&{{\it b_3}}^{-2}+2
\end {array} \right]
 \]
and the covariance of $c_1$, ..., $c_6$ is 
\[ \tiny
\left[ \begin {array}{cccccc}  
( 1+{{\it b_1}}^{2} ) {{\it a_1}}^{2}&
{\it a_2}\,{{\it b_1}}^{2}{\it a_1}&
{\it a_3}\,{\it a_1}\,{{\it b_1}}^{2}{\it b_2}&
{\it a_4}\,{\it a_1}\,{{\it b_1}}^{2}{\it b_2}&
{\it a_5}\,{\it b_3}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_1}&
{\it a_6}\,{\it b_3}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_1}\\ 
{\it a_2}\,{{\it b_1}}^{2}{\it a_1}& 
( 1+{{\it b_1}}^{2} ) {{\it a_2}}^{2}&
{\it a_3}\,{\it a_2}\,{{\it b_1}}^{2}{\it b_2}&
{\it a_4}\,{\it a_2}\,{{\it b_1}}^{2}{\it b_2}&
{\it a_5}\,{\it b_3}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_2}&
{\it a_6}\,{\it b_3}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_2}\\ 
{\it a_3}\,{\it a_1}\,{{\it b_1}}^{2}{\it b_2}&
{\it a_3}\,{\it a_2}\,{{\it b_1}}^{2}{\it b_2}& 
[ 1+{{\it b_2}}^{2}(1+{{\it b_1}}^{2}) ] {{\it a_3}}^{2}&
{\it a_4}\, ( 1+{{\it b_1}}^{2}) {\it a_3}\,{{\it b_2}}^{2}&
{\it a_5}\,{\it b_3}\, ( 1+{{\it b_1}}^{2} ) {{\it b_2}}^{2}{\it a_3}&
{\it a_6}\,{\it b_3}\, ( 1+{{\it b_1}}^{2} ) {{\it b_2}}^{2}{\it a_3}\\ 
{\it a_4}\,{\it a_1}\,{{\it b_1}}^{2}{\it b_2}&
{\it a_4}\,{\it a_2}\,{{\it b_1}}^{2}{\it b_2}&
{\it a_4}\, ( 1+{{\it b_1}}^{2} ) {\it a_3}\,{{\it b_2}}^{2}& 
[ 1+{{\it b_2}}^{2}(1+{{\it b_1}}^{2})] {{\it a_4}}^{2}&
{\it a_5}\,{\it b_3}\, ( 1+{{\it b_1}}^{2} ) {{\it b_2}}^{2}{\it a_4}&
{\it a_6}\,{\it b_3}\, ( 1+{{\it b_1}}^{2} ) {{\it b_2}}^{2}{\it a_4} \\ 
{\it a_5}\,{\it b_3}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_1}&
{\it a_5}\,{\it b_3}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_2}&
{\it a_5}\,{\it b_3}\, ( 1+{{\it b_1}}^{2} ) {{\it b_2}}^{2}{\it a_3}&
{\it a_5}\,{\it b_3}\, ( 1+{{\it b_1}}^{2} ) {{\it b_2}}^{2}{\it a_4}&
 \{$1+${{\it b_3}}^{2}[$1+${{\it b_2}}^{2}(1+{{\it b_1}}^{2})] \} {{\it a_5}}^{2}&
 {\it a_6}\,{\it a_5}\,{{\it b_3}}^{2} [1+{{\it b_2}}^{2}(1+{{\it b_1}}^{2})] \\ 
 {\it a_6}\,{\it b_3}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_1}&
 {\it a_6}\,{\it b_3}\,{\it b_2}\,{{\it b_1}}^{2}{\it a_2}&
 {\it a_6}\,{\it b_3}\, ( 1+{{\it b_1}}^{2} ) {{\it b_2}}^{2}{\it a_3}&
 {\it a_6}\,{\it b_3}\, ( 1+{{\it b_1}}^{2} ) {{\it b_2}}^{2}{\it a_4}&
 {\it a_6}\,{\it a_5}\,{{\it b_3}}^{2} [ 1+{{\it b_2}}^{2}(1+{{\it b_1}}^{2})] & 
 \{ $1+${{\it b_3}}^{2}[$1+${{\it b_2}}^{2}(1+{{\it b_1}}^{2}) ]\} {{\it a_6}}^{2}
 \end {array} \right] 
\]

From these examples we can list a set of properties as follows.

* **Property 1**: The marginal variance is equal the conditional one times 
a factor greater tan one, which is equal one plus the variance of the immediate parent. 
* **Property 2**: The covariance between two variables is equal the product of 
the weights of the edges in the path linking them times the variance of the top indexed parent in the path.
* **Property 3**: The (absolute) correlation value is equal $v/(1+v)$, 
where $v$ is the variance of the top indexed parent in the path.
   * **Property 3a**: The correlation strength increases as $v$ increases. 
   * **Property 3b**: The correlation tends $-1$ or $1$ if $v \rightarrow \infty$.
   * **Property 3c**: For pairs with common parent $v$ is the variance of the parent.
* **Property 4**: The sign of the covariance/correlation is the sign the product of the 
weights of the edges in the path linking them.
* **Property 5**: The maximum number of distinct absolute correlation values is 
equal to $n+p(p-1)/2$, achieved when each latent variable has at least two (non-latent) variables as children.
