---
output:
  pdf_document: 
    keep_tex: yes
  html_document:
    df_print: paged
header-includes:
- \usepackage{lscape}
- \usepackage{multicol}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(graphpcor)
```

**Extending the DAG's approach**

- Correlation between parents:
  - model the correlation between children of different parents
  - correlation between children of same parent is as before

- Add links between children: 
  - full range of correlation between children with link between then
  - correlation between children of different parent as before

- Undirected links model:
   - between children variables, no need of parent nodes
   - is a more general approach
   
## Correlation between parents


In our specific DAG we have $n$ variables of interest, the
children, and $k$ latent variables. 
Specifically, we define let a set of $n$ children, $c_i$, $i$=1,...,$n$, 
and $k$ latents, $p_j$, $j$=1,...,$k$.


```{r graphs, echo = FALSE, fig.width = 9, fig.height = 3, out.width="99%", fig.align = 'center', fig.cap = "Graphs with parents and children."}
g2 <- list(p1 ~ p2 + c1 + c2, p2 ~ c3 + c4)
g2p <- GraphPlot(g2)
g3a <- list(p1 ~ p2 + p3 + c1 + c2, p2 ~ c3 + c4, p3 ~ c5 + c6)
g3b <- list(p1 ~ p2 + c1 + c2, p2 ~ p3 + c3 + c4, p3 ~ c5 + c6)
g3ap <- GraphPlot(g3a)
g3bp <- GraphPlot(g3b)
par(mfrow = c(1, 3), mar = c(0, 0, 0, 0))
plot(g2p$gr, nodeAttrs = g2p$nAttrs)
legend("topleft", "(A)", bty = "n")
plot(g3ap$gr, nodeAttrs = g3ap$nAttrs)
legend("topleft", "(B)", bty = "n")
plot(g3bp$gr, nodeAttrs = g3bp$nAttrs)
legend("topleft", "(C)", bty = "n")
```


* **1st rule**: Joint distribution for parents,
$p_1$, ..., $p_k$, with unit diagonal precision matrix and
off diagonal elements equals zero if there is no link between 
$p_i$ and $p_j$, or the conditional correlation parameter.

* **2nd rule**: For $c_i$, $i=1$, ..., $n$, we set
N($p_{c_i}$, 1), where $p_{c_i}$ is the parent of $c_i$.
**(as before)**

The joint distribution for all the $n+k$ variables is
\[ 
p(c_1,\ldots,c_n,p_1,\ldots,p_k) \propto 
 \mathrm{exp} \left\{ \frac{-1}{2} \left[ 
 \sum_{i=1}^n (c_j - p_{c_j})^2 + 
 \mathbf{p}^{T}\mathbf{R}\mathbf{p}
 \right] \right\} 
\]


For the DAGs (A), (B) and (C) in Figure 1 we have
\[ 
\mathbf{R}^{(A)} = \left[ 
\begin{array}{cc} 
a^2 & -r \\ -r & b^2 
\end{array} \right],\;
\mathbf{R}^{(B)} = \left[ 
\begin{array}{ccc} 
a^2 & -r & -s \\ 
-r & b^2 &  \\
-s &  & c^2
\end{array} \right]\; \textrm{ and }\;
\mathbf{R}^{(C)} = \left[ 
\begin{array}{ccc} 
a^2 & -r &  \\ 
-r & b^2 & -s \\
 & -s  & c^2
\end{array} \right]
\]

For the model from (A), the joint distribution has precision
\[
\left[ \begin {array}{cccccc} 
1&0&0&0&-1&0\\ 
0&1&0&0&-1&0\\ 
0&0&1&0&0&-1\\ 
0&0&0&1&0&-1\\ 
-1&-1&0&0&{a}^{2}+2&-r\\ 
0&0&-1&-1&-r&{b}^{2}+2
\end {array} \right] 
\] 
and the covariance for $c_1$, ..., $c_4$ is
\[ \frac{1}{1-a^2b^2r^2}
 \left[ \begin {array}{cccccc} 1-{r}^{2}{a}^{2}{b}^{2}+{a}^{2}&{a}^{2}&r{a}^{2}{b}^{2}&r{a}^{2}{b}^{2}&{a}^{2}&r{a}^{2}{b}^{2}\\ 
{a}^{2}&1-{r}^{2}{a}^{2}{b}^{2}+{a}^{2}&r{a}^{2}{b}^{2}&r{a}^{2}{b}^{2}&{a}^{2}&r{a}^{2}{b}^{2} \\ 
r{a}^{2}{b}^{2}&r{a}^{2}{b}^{2}&1-{r}^{2}{a}^{2}{b}^{2}+{b}^{2}&{b}^{2}&r{a}^{2}{b}^{2}&{b}^{2}
\\ r{a}^{2}{b}^{2}&r{a}^{2}{b}^{2}&{b}^{2}&1-{r}^{2}{a}^{2}{b}^{2}+{b}^{2}&r{a}^{2}{b}^{2}&{b}^{2} \\ 
{a}^{2}&{a}^{2}&r{a}^{2}{b}^{2}&r{a}^{2}{b}^{2}&{a}^{2}&r{a}^{2}{b}^{2}\\ 
r{a}^{2}{b}^{2}&r{a}^{2}{b}^{2}&{b}^{2}&{b}^{2}&r{a}^{2}{b}^{2}&{b}^{2}\end {array} \right] .
\]
If $r=0$ we have no correlation between children of different parents.

This gives 
C($c_1$, $c_2$) = $\frac{a^2}{1+a^2(1-r^2b^2)}$, 
C($c_1$,$c_3$) = 
C($c_1$,$c_4$) = 
C($c_2$,$c_3$) = 
C($c_2$,$c_4$) = 
$\frac{ra^2b^2}{\sqrt{1+a^2(1-r^2b^2)}\sqrt{1+b^2(1-a^2r^2)}}$
and C($c_3$,$c_4$)= $\frac{b^2}{b^2(1+a^2r^2)}$


```{r correls1, echo = FALSE}
c12 <- function(r,a=1,b=a) {
  a^2/(1+a^2*(1-b^2*r^2))
}
c12 <- function(r,a=1,b=1) {
  x <- -1 + r^2*a^2*b^2
  n <- a^2/x
  d <- (-1 + r^2*a^2*b^2 -a^2)/x
  -n/d
}
c13 <- function(r,a=1,b=1) {
  n <- r*a^2*b^2
  d <- sqrt((a^2*(1-b^2*r^2)+1) * (b^2*(1-a^2*r^2))+1)
  return(n/d)
}
c34 <- function(r,a=1,b=1){
  n <- b^2
  d <- b^2*(1-a^2*r^2)+1
  return(n/d)
}
rr <- c(-0.99, -0.9, -0.5, 0, 0.5, 0.9, 0.99) 
par(mfrow = c(2, 2), mar = c(3, 3, 1, 1), mgp = c(2, 1, 0), las = 1)
plot(function(x) c12(0,x,1), 0.1, 10, 
     type = "n", bty = "n", log = "x", ylim = c(-1,1), 
     xlab = "a", ylab = expression(C(c[1],c[2])))
for(r in rr) 
  plot(function(x) c12(r,x,1), 0.01, 100, 
       add = TRUE, n = 1001, 
       lty = which(rr==r))
aa <- c(0.1, 1, 10)
bb <- c(0.1, 1, 10)
plot(function(x) c13(x,1,1), -1, 1,
     type = "n", bty = "n", ylim = c(-1,1), 
     xlab = "r", ylab = expression(C(c[1],c[3])))
for(a in aa) 
  for(b in bb)
  plot(function(x) c13(x,a,b), -1, 1, 
       add = TRUE, n = 1001, 
       col = which(a==aa), lty = which(b==bb))
plot(function(x) c34(0,x,1), 0.1, 10, 
     type = "n", bty = "n", log = "x", ylim = c(-1,1), 
     xlab = "b", ylab = expression(C(c[3],c[4])))
for(r in rr)
  for(a in aa)
    plot(function(x) c34(r,a,x), 0.01, 100, 
         add = TRUE, n = 10001, 
         col = which(a==aa), lty = which(rr==r))
```

Notice that if $a\rightarrow 0$ C($c_1$,$c_2$) $\rightarrow 0$ but
if both $a \rightarrow 0$ and $b\rightarrow 0$ we have C($c_3$,$c_4$) = 1/2. 

## Add links between children

**as before** but add undirected links between children,
having a joint distribution for children as
\[
p(\mathbf{c}) \propto \exp \left\{
-\frac{1}{2}(\mathbf{c}-\mathbf{p}_\mathbf{c})^T\mathbf{K}
(\mathbf{c}-\mathbf{p}_\mathbf{c})
\right\}. 
\]

### Example: 
Four children with some links and a common parent.

\begin{figure}
\centering
\includegraphics[angle=90,width=0.5\textwidth]{p1c4.jpeg}
\caption{Four children with some links and a common parent.}
\end{figure}

The joint distribution can be written as
\[ p(\mathbf{c},\mathbf{p}) \propto \exp \left\{ -\frac{1}{2} \left(
\left[ \begin{array}{c} c_1$-$p_1 \\ c_2$-$p_1 \end{array} \right]^T
\left[ \begin{array}{cc}
1 & $-$r \\ $-$r & 1
\end{array} \right] 
\left[ \begin{array}{c} c_1$-$p_1 \\ c_2$-$p_1 \end{array} \right] + 
\left[ \begin{array}{c} c_3$-$p_2 \\ c_4$-$p_2 \end{array} \right]^T
\left[ \begin{array}{cc}
1 & $-$s \\ $-$s & 1
\end{array} \right]
\left[ \begin{array}{c} c_3$-$p_2 \\ c_4$-$p_2 \end{array} \right]
+\frac{p_1^2}{q^2}
\right) \right\} .
\]

This gives the joint precision as 
\[
\left[ \begin {array}{ccccc} 
1&-r&0&0&1-r\\ 
-r&1&0&0&1-r\\ 
0&0&1&-s&1-s\\ 
0&0&-s&1&1-s\\ 
1-r&1-r&1-s&1-s&{q}^{-2}+2(2-r-s)
\end {array} \right] 
\]
and the covariance for the children as
\[
\left[ \begin {array}{cccc} 
{\frac {-1-{q}^{2}+{q}^{2}{r}^{2}}{{r}^{2}-1}}&
{\frac {-r-{q}^{2}+{q}^{2}{r}^{2}}{{r}^{2}-1}}&
{q}^{2}&{q}^{2}
\\ {\frac {-r-{q}^{2}+{q}^{2}{r}^{2}}{{r}^{2}-1}}&{\frac {-1-{q}^{2}+{q}^{2}{r}^{2}}{{r}^{2}-1}}&{q}^{2}&{q}^{2}
\\ {q}^{2}&{q}^{2}&{\frac {-1-{q}^{2}+{q}^{2}{s}^{2}}{{s}^{2}-1}}&{\frac {-s+{q}^{2}{s}^{2}-{q}^{2}}{{s}^{2}-1}}
\\ {q}^{2}&{q}^{2}&{\frac {-s+{q}^{2}{s}^{2}-{q}^{2}}{{s}^{2}-1}}&{\frac {-1-{q}^{2}+{q}^{2}{s}^{2}}{{s}^{2}-1}}
\end {array} \right]
\]

The correlations are 
C($c_1$, $c_2$) = $\frac {-r-{q}^{2}+{q}^{2}{r}^{2}}{-1-{q}^{2}+{q}^{2}{r}^{2}}$, 
C($c_1$,$c_3$) = 
C($c_1$,$c_4$) = 
C($c_2$,$c_3$) = 
C($c_2$,$c_4$) = 
$\frac{{q}^{2}}{\sqrt {{\frac {-1-{q}^{2}+{q}^{2}{r}^{2}}{{r}^{2}-1}}}\sqrt {{\frac {-1-{q}^{2}+{q}^{2}{s}^{2}}{{s}^{2}-1}}}}$
and 
C($c_3$,$c_4$) = 
$\frac {-s-{q}^{2}+{q}^{2}{s}^{2}}{-1-{q}^{2}+{q}^{2}{s}^{2}}$.

Correlation instead 

\includegraphics[width=0.5\textwidth]{prpc4}

For the above graph, we consider the joint distribution for the 
parents as 
\[ \left[ \begin{array}{cc} 
1 & -a \\ -a & 1 
\end{array} \right]
\] 
and the ones for the children as before. 

This gives the joint precision as 

