---
title: "General framework for graph derived correlation matrices"
author: "H. Rue et al"
date: "Dec, 2023"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(INLA)
options(digits = 2)
hrexample <- TRUE
```

## The fill-in elements

Let $\mathbf{Q}$ be a $n \times n$ sparse symmetric positive definite matrix 
which can be written as $\mathbf{Q} = \mathbf{L}\mathbf{L}^T$, 
where $\mathbf{L}$ is a lower triangular matrix 
called the Cholesky triangle/factor of $\mathbf{Q}$.
The elements $\mathbf{Q}_{ij}$ are given from $\mathbf{L}$ as
\[
\mathbf{Q}_{ij} = \sum_{k=1}^j \mathbf{L}_{ik}\mathbf{L}_{jk},
\textrm{ for } i\geq j,
\]
which depends on previously computed elements of $\mathbf{L}$ at 
lines $i$ and $j$ from column $1$ up to column $j$.

Let $\mathbf{R}$ be the lower triangle of $\mathbf{Q}$
including its diagonal.
For certain models the number of non-zero elements in $\mathbf{L}$
is equal the number of non-zero elements in $\mathbf{R}$.
However, in general $\mathbf{L}$ contains more non-zero elements than $\mathbf{R}$.
The non-zero elements of $\mathbf{L}$ at the corresponding elements in $\mathbf{R}$
which are equal to zero are called the fill-in elements.
Let $n_R$ be the number of non-zero elements in $\mathbf{R}$ and $n_L$ be the 
number of non-zero elements in $\mathbf{L}$.

We parametrize $\mathbf{L}$ having its diagonal elements 
$\mathbf{L}_{ii} = \exp(\theta_i) > 0$ for $1<i<n$ and, 
the $n_R-n$ off-diagonal elements in $\mathbf{L}$ whose corresponding 
non-zero elements of $\mathbf{R}$, for $i>j$, as
$\mathbf{L}_{ij} = \theta_k \in \mathbb{R}$, $k\in\{n+1,\ldots,n_R\}$.
The remaining $n_L-n_R>0$ fill-in elements in $\mathbf{L}$ are computed with
\[
\mathbf{L}_{ij} = -\frac{\sum_{k=1}^{j-1} \mathbf{L}_{ik}\mathbf{L}_{jk}}
{\mathbf{L}_{jj}}
\]
starting from the top-left fill-in element,
line by line from left to right and from top to bottom.

# Example

Let a $n \times n$ SPD matrix $\mathbf{Q}$.

```{r graph, include = hrexample}
## Example (from "idea.pdf")
Q <- Matrix(cbind(
  c(3,0,-1,0,0,0,0,-1,0,-1),
  c(0,2,-1,0,0,0,-1,0,0,0),
  c(-1,-1,2,0,0,0,0,0,0,0),
  c(0,0,0,2,-1,0,0,-1,0,0),
  c(0,0,0,-1,3,-1,0,-1,0,0),
  c(0,0,0,0,-1,2,-1,0,0,0),
  c(0,-1,0,0,0,-1,2,0,0,0),
  c(-1,0,0,-1,-1,0,0,5,-1,-1),
  c(0,0,0,0,0,0,0,-1,1,0),
  c(-1,0,0,0,0,0,0,-1,0,2)
) + diag(10))
n <- nrow(Q)
```

```{r anotherR, eval=!hrexample, include = !hrexample}
rQ <- function(n) {
  Q <- cov(matrix(rnorm(10*n*n), ncol = n)) 
  p <- log(n)/n
  Q[lower.tri(Q)] <- Q[lower.tri(Q)] * rbinom(n*(n-1)/2, 1, p)
  Q[upper.tri(Q)] <- t(Q)[upper.tri(Q)]
 return(Matrix(Q))
}
n <- 10
Q <- rQ(n)
```

Check it 
```{r Qcheck}  
isSymmetric(Q)
q <- Q * lower.tri(Q, diag = TRUE)
q
nnzq <- sum(q!=0) ## number of non-zeros in the lower triangle of Q
nnzq
g <- inla.read.graph(Q)
plot(g)
```

The Cholesky decomposition: $\mathbf{Q} = \mathbf{L}\mathbf{L}^T$ 
```{r l1}
l <- t(chol(Q))
l
nnzl <- sum(l!=0) ### number of nonzeros in L
c(nnzq, nnzl)
```

Ordering
```{r og}
og <- inla.qreordering(Q)
og$ireordering
Q.o <- Q[og$i, og$i]
q.o <- Q.o*lower.tri(Q.o, diag = TRUE)
l.o <- t(chol(Q.o))
l.o
nnzl.o <- sum(l.o!=0) ### non-zeros of the Cholesky factor after reordering Q
c(nnzq, nnzl, nnzl.o)
lfi <- ((l.o!=0) & (q.o==0))*1L
lfi 
nlfi <- sum(lfi)
```

Visualize the ordered graph (right), the fill-in edges (center) and the 
graph from the $\mathbf{L}+\mathbf{L}^T$ pattern (right):
```{r l2plot, fig.width=12, fig.height=6, include = nlfi>0}
par(mfrow = c(1, 3))
plot(inla.read.graph(Q.o))
plot(inla.read.graph(lfi + t(lfi)))
plot(inla.read.graph(l.o+t(l.o)))
```

The $i$ and $j$ index set of the fill-in elements

```{r ij, include=nlfi>0}
ii <- row(lfi)[as.matrix(lfi)!=0]
ii
jj <- col(lfi)[as.matrix(lfi)!=0]
jj
```


Start by the matrix with the fill-in positions as NA

```{r lres, include = nlfi>0}
### starting L (from the Q_ij pattern)
l.res <- as.matrix(l.o)
l.res[which(lfi==1)] <- NA
Matrix(l.res) ## this is what we have to start with
```

Fill this matrix computing the fill-in elements, 
from top-left, at each line from left to right.

```{r lfill, include = nlfi>0}
### NOTE: ii and jj index vectors are ordered (topleft to bottomright)
for(v in 1:length(ii)) {
  i <- ii[v]
  j <- jj[v]
  if(j==1) {
    cat("j = 1\n")
    l.res[i,1] <- 0.0
  } else {
    stopifnot((i>1) & (j>1)) ## L_{11} not allowed
    stopifnot(j>1) ## j=1 is not allowed
    k <- 1:(j-1)
    l.res[i, j] <- -sum(l.res[i, k] * l.res[j, k]) / l.res[j, j]
  }
}
Matrix(l.res)
all.equal(Matrix(l.res), l.o)
```
