---
title: 'Supplement for "A graphical approach for multivariate modeling" by'
author: "Anna Freni-Sterrantino, Denis Rustand, Janet van Niekerk, Elias T Krainski, and H\\aa vard Rue"
date: "November, 2023"
output:
  pdf_document: 
    keep_tex: yes
  html_document:
    df_print: paged
header-includes:
- \usepackage{lscape}
- \usepackage{multicol}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(corGraphs)
```


This material summarizes the model derivation and its properties.
We first define a specific kind of direct acyclic graph - DAG which is 
used to define the model and provides an intuitive visualization. 
Then we define three simple rules for writing the joint distribution
for all the variables in the model.
The properties are follows from the covariance and correlation matrices derived.
Illustrative examples are considered all the way along.

## The DAG definition

As usual in a DAG, the variables are represented as nodes and 
the arrows represents the relation between them.
In our specific DAG we have $n$ variables of interest, the
children, and $k$ latent variables, the parents. 
Specifically, we define let a set of $n$ children, $c_i$, $i$=1,...,$n$, 
and $k$ parents, $p_j$, $j$=1,...,$k$.
A parent can have a parent but a children cannot have children.
Also, we have a main parent, with no parent, 
and a unique arrow arriving at each of the other variables.
We have some examples in Figure 1 bellow.

```{r graphs, echo = FALSE, fig.width = 9, fig.height = 4, out.width="79%", fig.align = 'center', fig.cap = "Graphs representing different models. A model with one parent, (A), a model with two parents (B), and two different models with three parents (C) and (D)."}
g1 <- list(p1 ~ c1 + c2 + c3)
g1p <- GraphPlot(g1)
g2 <- list(p1 ~ p2 + c3 + c4, p2 ~ c1 + c2)
g2p <- GraphPlot(g2)
g3a <- list(p1 ~ p2 + p3 + c1 + c2, p2 ~ c3 + c4, p3 ~ c5 + c6)
g3b <- list(p1 ~ p2 + c1 + c2, p2 ~ p3 + c3 + c4, p3 ~ c5 + c6)
g3ap <- GraphPlot(g3a)
g3bp <- GraphPlot(g3b)
par(mfrow = c(2, 2), mar = c(0, 0, 0, 0))
plot(g1p$gr, nodeAttrs = g1p$nAttrs)
legend("topleft", "(A)", bty = "n")
plot(g2p$gr, nodeAttrs = g2p$nAttrs)
legend("topleft", "(B)", bty = "n")
plot(g3ap$gr, nodeAttrs = g3ap$nAttrs)
legend("topleft", "(C)", bty = "n")
plot(g3bp$gr, nodeAttrs = g3bp$nAttrs)
legend("topleft", "(D)", bty = "n")
```


The covariance between the children is derived from the
relationship between the parents and how they are shared by the variables. 
To build a DAG for a problem we have to consider that, as we will see later, 
usually stronger correlation are induced between variables linked by a shorter path. 
Let C(a,b) be the correlation between the variables $a$ and $b$.
In the model defined with graph (A), 
$|$C($c_1$,$c_2$)$|$=$|$C($c_1$,$c_3$)$|$=$|$C($c_2$,$c_3$)$|$.
In the model defined with graph (B), 
$|$C($c_1$,$c_3$)$|$=$|$C($c_1$,$c_4$)$|$=
$|$C($c_2$,$c_3$)$|$=$|$C($c_2$,$c_4$)$|$,
$|$C($c_1$,$c_3$)$|\leq|$C($c_1$,$c_2$)$|$, and 
$|$C($c_1$,$c_3$)$|\leq|$C($c_3$,$c_4$)$|$.

Both the graphs (C) and (D) have three latent variables 
with $p_1$ as the main node. 
However, in the sense of a path between parent nodes, 
$p_1$ is a central parent in (C), but not in (D), where $p_2$ is.
This changes the pair correlation order, 
for example, there are three latent variables in the path between 
$c_1$ and $c_5$ in (C) but two in (D). 
This changes the marginal properties for the variables involved 
as we will see later on.

## Model definition

The specific DAG given is now used along with simple rules to define the model. 
This model definition starts assuming a distribution for the 
main parent and then a set of conditional distributions for 
the $n+p-1$ other variables. 
The number of conditional distributions is equal the number of arrows in the DAG.
We assume Gaussian distributions and so we just need to define 
the (conditional) mean and variance.

* **1st rule**: For the main parent $p_1$, the one with no parent, 
the mean is zero and the variance is $v_1^2$. 

* **2nd rule**: The conditional mean for $p_j$, $j=2$,...$k$, 
is its parent and the conditional variance is $v_j^2$.

* **3nd rule**: The conditional mean of $c_i$, $i=1$, ..., $n$, 
is $s_i$ times its parent and the conditional variance is $s_i^2$.

In this model we have $n+k$ parameters. 
As $k<n$, this represents a reduction, from $n(n+1)/2$.

As an illustration, consider the graph (A) from Figure 1.
The joint distribution for $\{c_1, c_2, c_3, p_1\}$ is 
\[
\pi(c_1, c_2, c_3, p_1) \propto \mathrm{exp} 
  \left\{-\frac{1}{2} \left[ \frac{p_1^2}{v_1^2} + \frac{(c_1-s_1p_1)^2}{s_1^2} + \frac{(c_2-s_2p_1)^2}{s_2^2} + \frac{(c_3-s_3p_1)^2}{s_3^2} \right] \right\}.
\]
From this set of conditional distributions we obtain the precision matrix as 
\[
\left[ \begin {array}{cccc} 
{{\it s_1}}^{-2}&0&0&-{{\it s_1}}^{-1}\\ 
0&{{\it s_2}}^{-2}&0&-{{\it s_2}}^{-1}\\ 
0&0&{{\it s_3}}^{-2}&-{{\it s_3}}^{-1}\\ 
-{{\it s_1}}^{-1}&-{{\it s_2}}^{-1}&-{{\it s_3}}^{-1}&{{\it v_1}}^{-2}+3
\end {array} \right].
\]

The joint distribution for $\{c_1, c_2, c_3, c_4, p_1, p_2\}$, 
from the DAG (B) in Figure 1, is proportional to
\[ 
\mathrm{exp} 
  \left\{-\frac{1}{2} \left[ \frac{p_1^2}{v_1^2} + \frac{(p_2-p_1)^2}{v_2^2} +
  \frac{(c_1-s_1p_1)^2}{s_1^2} + \frac{(c_2-s_2p_1)^2}{s_2^2} +
  \frac{(c_3-s_3p_2)^2}{s_3^2} + \frac{(c_4-s_4p_2)^2}{s_4^2} \right] \right\}.
\]
The precision matrix is 
\[
\left[ \begin {array}{cccccc} 
{{\it s_1}}^{-2}&0&0&0&-{{\it s_1}}^{-1}&0\\ 
0&{{\it s_2}}^{-2}&0&0&-{{\it s_2}}^{-1}&0\\ 
0&0&{{\it s_3}}^{-2}&0&0&-{{\it s_3}}^{-1}\\ 
0&0&0&{{\it s_4}}^{-2}&0&-{{\it s_4}}^{-1}\\ 
-{{\it s_1}}^{-1}&-{{\it s_2}}^{-1}&0&0&
{{\it v_1}}^{-2}+{{\it v_2}}^{-2}+2&-{{\it v_2}}^{-2}\\ 
0&0&-{{\it s_3}}^{-1}&-{{\it s_4}}^{-1}&-{{\it v_2}}^{-2}&{{\it v_2}}^{-2}+2
\end {array} \right] 
\]

The models represented by graphs (C) and (D) in Figure 1 have both the same
number of children and parents. 
The only difference is that parent $p_3$ is children to parent $p_1$ in (C) 
whereas in (D) $p_3$ is children to parent $p_2$.
For the case when $p_1$ is parent to $p_3$ the last three rows of the precision matrix are 
\[
\left[ \begin {array}{ccccccccc} 
%% {{\it s_1}}^{-2}&0&0&0&0&0&-{{\it s_1}}^{-1}&0&0\\ 
%% 0&{{\it s_2}}^{-2}&0&0&0&0&-{{\it s_2}}^{-1}&0&0\\ 
%% 0&0&{{\it s_3}}^{-2}&0&0&0&0&-{{\it s_3}}^{-1}&0\\ 
%% 0&0&0&{{\it s_4}}^{-2}&0&0&0&-{{\it s_4}}^{-1}&0\\ 
%% 0&0&0&0&{{\it s_5}}^{-2}&0&0&0&-{{\it s_5}}^{-1} \\ 
%% 0&0&0&0&0&{{\it s_6}}^{-2}&0&0&-{{\it s_6}}^{-1}\\ 
-{{\it s_1}}^{-1}&-{{\it s_2}}^{-1}&0&0&0&0&
 {{\it v_1}}^{-2}+{{\it v_2}}^{-2}+{{\it v_3}}^{-2} +2 &
-{{\it v_2}}^{-2}&-{{\it v_3}}^{-2}\\ 
0&0&-{{\it s_3}}^{-1}&-{{\it s_4}}^{-1}&0&0&
 -{{\it v_2}}^{-2}&{{\it v_2}}^{-2}+2&0\\ 
 0&0&0&0&-{{\it s_5}}^{-1}&-{{\it s_6}}^{-1}&
 -{{\it v_3}}^{-2}&0&{{\it v_3}}^{-2}+2
 \end {array} \right]
\]
and the case when $p_2$ is parent to $p_3$ the last three rows of the precision matrix are
\[ 
\left[ \begin {array}{ccccccccc} 
-{{\it s_1}}^{-1}&-{{\it s_2}}^{-1}&0&0&0&0&{{\it v_1}}^{-2}+{{\it v_2}}^{-2}+2&-{{\it v_2}}^{-2}
&0\\ 0&0&-{{\it s_3}}^{-1}&-{{\it s_4}}^{-1}&0&0&-{{\it v_2}}^{-2}&{{\it v_2}}^{-2}+{{\it v_3}}^{-2}+2&-{{\it v_3}}^{-2}\\ 0&0&0&0&-{{\it s_5}}^{-1}&-{
{\it s_6}}^{-1}&0&-{{\it v_3}}^{-2}&{{\it v_3}}^{-2}+2\end {array} \right] 
 \]
and the first $6$ rows are equal the ones from the previous case.

## Model properties

Our primarily interest is in the correlation between the children variables. 
The joint distribution defined from the DAG and the simple rules,
also defines the precision matrix for the $n+k$ variables.
We solve it and consider the covariance, first $n$ rows and columns
to learn the marginal properties, and correlations as well.

The model defined from the DAG (A) in Figure 1 for 
$\{c_1, c_2, c_3, p_1\}$ induces a covariance matrix equals to
\[ 
\left[ \begin {array}{cccc}  
( 1+{{\it v_1}}^{2} ) {{\it s_1}}^{2} & {\it s_2}{\it s_1}{{\it v_1}}^{2}&
{\it s_3}{\it s_1}{{\it v_1}}^{2} & {\it s_1}{{\it v_1}}^{2} \\ 
{\it s_2}{\it s_1}{{\it v_1}}^{2} & ( 1+{{\it v_1}}^{2} ) {{\it s_2}}^{2}& 
{\it s_3}{\it s_2}{{\it v_1}}^{2} & {\it s_2}{{\it v_1}}^{2} \\ 
{\it s_3}{\it s_1}{{\it v_1}}^{2} &{\it s_3}{\it s_2}{{\it v_1}}^{2}& 
( 1+{{\it v_1}}^{2} ) {{\it s_3}}^{2} & {\it s_3}{{\it v_1}}^{2} \\ 
{\it s_1}{{\it v_1}}^{2} & {\it s_2}{{\it v_1}}^{2} &
{\it s_3}{{\it v_1}}^{2}&{{\it v_1}}^{2}
\end {array} \right] 
\]
and the C($c_i,c_j$) = sign($s_is_j$)$v_1^2/(1+v_1^2)$, for $i,j\in\{1,2,3\}$ and $i\neq j$.

The model defined from the DAG (B) in Figure 1 gives the covariance matrix 
for $\{c_1, c_2, c_3, c_4, p_1, p_2\}$ as
\[ 
\left[ \begin {array}{cccccc}  \left( 1+{{\it v_1}}^{2} \right) {{\it s_1}}^{2}&{\it s_2}\,{{\it v_1}}^{2}{\it s_1}&{\it s_3}\,{{\it v_1}}^{2}{\it s_1}&{\it s_4}\,{{\it v_1}}^{2}{\it 
s_1}&{{\it v_1}}^{2}{\it s_1}&{{\it v_1}}^{2}{\it s_1}\\ {\it s_2}\,{{\it v_1}}^{2}{\it s_1}& \left( 1+{{\it v_1}}^{2} \right) {{\it s_2}}^{2}&{\it s_3}\,{{\it v_1}
}^{2}{\it s_2}&{\it s_4}\,{{\it v_1}}^{2}{\it s_2}&{{\it v_1}}^{2}{\it s_2}&{{\it v_1}}^{2}{\it s_2}\\ {\it s_3}\,{{\it v_1}}^{2}{\it s_1}&{\it s_3}\,{{\it v_1}}^{
2}{\it s_2}& \left( 1+{{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {{\it s_3}}^{2}&{\it s_4}\, \left( {{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {\it s_3}&{{\it v_1}}^{2}{\it s_3}& \left( {{
\it v_2}}^{2}+{{\it v_1}}^{2} \right) {\it s_3}\\ {\it s_4}\,{{\it v_1}}^{2}{\it s_1}&{\it s_4}\,{{\it v_1}}^{2}{\it s_2}&{\it s_4}\, \left( {{\it v_2}}^{2}+{{\it 
v_1}}^{2} \right) {\it s_3}& \left( 1+{{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {{\it s_4}}^{2}&{{\it v_1}}^{2}{\it s_4}& \left( {{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {\it s_4}
\\ {{\it v_1}}^{2}{\it s_1}&{{\it v_1}}^{2}{\it s_2}&{{\it v_1}}^{2}{\it s_3}&{{\it v_1}}^{2}{\it s_4}&{{\it v_1}}^{2}&{{\it v_1}}^{2}\\ {{\it v_1}
}^{2}{\it s_1}&{{\it v_1}}^{2}{\it s_2}& \left( {{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {\it s_3}& \left( {{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {\it s_4}&{{\it v_1}}^{2}&{{\it v_2}}
^{2}+{{\it v_1}}^{2}\end {array} \right] 
\]
from what the marginal properties can be learn. 
The upper triangle of the children correlation matrix is
\[
\left[ \begin {array}{ccc} 
\frac{\textrm{sign}(s_1s_2){{\it v_1}}^{2}}{{( {{\it v_1}}^{2}+1 )}} & 
\frac{\textrm{sign}(s_1 s_3){{\it v_1}}^{2}} {\sqrt{( {{\it v_1}}^{2}+1 )} \sqrt { ( 1+{{\it v_2}}^{2}+{{\it v_1}}^{2} )}} & 
\frac{\textrm{sign}(s_1 s_4){{\it v_1}}^{2}} {\sqrt{( {{\it v_1}}^{2}+1 )} \sqrt { ( 1+{{\it v_2}}^{2}+{{\it v_1}}^{2} )}}\\ 
& \frac{\textrm{sign}(s_2 s_3){{\it v_1}}^{2}} {\sqrt{( {{\it v_1}}^{2}+1 )} \sqrt { ( 1+{{\it v_2}}^{2}+{{\it v_1}}^{2} )}} & 
\frac{\textrm{sign}(s_2 s_4){{\it v_1}}^{2}} {\sqrt{( {{\it v_1}}^{2}+1 )} \sqrt { ( 1+{{\it v_2}}^{2}+{{\it v_1}}^{2} )}}\\
 &  & \frac{{\textrm{sign}(s_3 s_4) ( {{\it v_2}}^{2}+{{\it v_1}}^{2}) }}{
{( 1+{{\it v_2}}^{2}+{{\it v_1}}^{2})}}
\end{array} \right] 
\]
where C($s_i,s_j$) = sign($s_is_j$)$\frac{v_1^2}{(v_1^2+1)(1+v_1^2+v_2^2)}$ 
for $i\in\{1,2\}$ and $j\in\{3,4\}$, which defines a block with the same absolute value 
and is the main structure that this model induced for the correlation matrix.

In the model given from graph (C), when $p_1$ is parent to $p_3$,
the covariance matrix of $\{c_1,...,c_6\}$ is 
\[ \scriptsize 
\left[ \begin {array}{cccccc}  \left( 1+{{\it v_1}}^{2} \right) {{\it s_1}}^{2}&{\it s_2}\,{{\it v_1}}^{2}{\it s_1}&{\it s_3}\,{{\it v_1}}^{2}{\it s_1}&{\it s_4}\,{{\it v_1}}^{2}{\it 
s_1}&{\it s_5}\,{{\it v_1}}^{2}{\it s_1}&{\it s_6}\,{{\it v_1}}^{2}{\it s_1}\\ {\it s_2}\,{{\it v_1}}^{2}{\it s_1}& \left( 1+{{\it v_1}}^{2} \right) {{\it s_2}}^{2}
&{\it s_3}\,{{\it v_1}}^{2}{\it s_2}&{\it s_4}\,{{\it v_1}}^{2}{\it s_2}&{\it s_5}\,{{\it v_1}}^{2}{\it s_2}&{\it s_6}\,{{\it v_1}}^{2}{\it s_2}\\ {\it s_3}\,{{
\it v_1}}^{2}{\it s_1}&{\it s_3}\,{{\it v_1}}^{2}{\it s_2}& \left( 1+{{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {{\it s_3}}^{2}&{\it s_4}\, \left( {{\it v_2}}^{2}+{{\it v_1}}^{2}
 \right) {\it s_3}&{\it s_5}\,{{\it v_1}}^{2}{\it s_3}&{\it s_6}\,{{\it v_1}}^{2}{\it s_3}\\ {\it s_4}\,{{\it v_1}}^{2}{\it s_1}&{\it s_4}\,{{\it v_1}}^{2}{\it s_2
}&{\it s_4}\, \left( {{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {\it s_3}& \left( 1+{{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {{\it s_4}}^{2}&{\it s_5}\,{{\it v_1}}^{2}{\it s_4}&{\it s_6}
\,{{\it v_1}}^{2}{\it s_4}\\ {\it s_5}\,{{\it v_1}}^{2}{\it s_1}&{\it s_5}\,{{\it v_1}}^{2}{\it s_2}&{\it s_5}\,{{\it v_1}}^{2}{\it s_3}&{\it s_5}\,{{\it v_1}}^{2}
{\it s_4}& \left( 1+{{\it v_3}}^{2}+{{\it v_1}}^{2} \right) {{\it s_5}}^{2}&{\it s_6}\,{\it s_5}\, \left( {{\it v_3}}^{2}+{{\it v_1}}^{2} \right) \\ {\it s_6}\,{{\it 
v_1}}^{2}{\it s_1}&{\it s_6}\,{{\it v_1}}^{2}{\it s_2}&{\it s_6}\,{{\it v_1}}^{2}{\it s_3}&{\it s_6}\,{{\it v_1}}^{2}{\it s_4}&{\it s_6}\,{\it s_5}\, \left( {{\it v_3}}^{2}+{{\it 
v_1}}^{2} \right) & \left( 1+{{\it v_3}}^{2}+{{\it v_1}}^{2} \right) {{\it s_6}}^{2}\end {array} \right] 
\]
where the variances and covariance for $c_5$ and $c_6$ does not depends on $v_2^2$.

For the case when $p_2$ is parent to $p_3$ the covariance of $c_1$, ..., $c_6$ is 
\[ \tiny
\left[ \begin {array}{cccccc}  \left( 1+{{\it v_1}}^{2} \right) {{\it s_1}}^{2}&{\it s_2}\,{{\it v_1}}^{2}{\it s_1}&{\it s_3}\,{{\it v_1}}^{2}{\it s_1}&{\it s_4}\,{{\it v_1}}^{2}{\it 
s_1}&{\it s_5}\,{{\it v_1}}^{2}{\it s_1}&{\it s_6}\,{{\it v_1}}^{2}{\it s_1}\\ {\it s_2}\,{{\it v_1}}^{2}{\it s_1}& \left( 1+{{\it v_1}}^{2} \right) {{\it s_2}}^{2}
&{\it s_3}\,{{\it v_1}}^{2}{\it s_2}&{\it s_4}\,{{\it v_1}}^{2}{\it s_2}&{\it s_5}\,{{\it v_1}}^{2}{\it s_2}&{\it s_6}\,{{\it v_1}}^{2}{\it s_2}\\ {\it s_3}\,{{
\it v_1}}^{2}{\it s_1}&{\it s_3}\,{{\it v_1}}^{2}{\it s_2}& \left( 1+{{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {{\it s_3}}^{2}&{\it s_4}\, \left( {{\it v_2}}^{2}+{{\it v_1}}^{2}
 \right) {\it s_3}&{\it s_5}\, \left( {{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {\it s_3}&{\it s_6}\, \left( {{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {\it s_3}\\ {\it 
s_4}\,{{\it v_1}}^{2}{\it s_1}&{\it s_4}\,{{\it v_1}}^{2}{\it s_2}&{\it s_4}\, \left( {{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {\it s_3}& \left( 1+{{\it v_2}}^{2}+{{\it v_1}}^{2}
 \right) {{\it s_4}}^{2}&{\it s_5}\, \left( {{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {\it s_4}&{\it s_6}\, \left( {{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {\it s_4}\\ {
\it s_5}\,{{\it v_1}}^{2}{\it s_1}&{\it s_5}\,{{\it v_1}}^{2}{\it s_2}&{\it s_5}\, \left( {{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {\it s_3}&{\it s_5}\, \left( {{\it v_2}}^{2}+{{\it 
v_1}}^{2} \right) {\it s_4}& \left( 1+{{\it v_3}}^{2}+{{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {{\it s_5}}^{2}&{\it s_6}\,{\it s_5}\, \left( {{\it v_3}}^{2}+{{\it v_2}}^{2}+{{\it v_1}}
^{2} \right) \\ {\it s_6}\,{{\it v_1}}^{2}{\it s_1}&{\it s_6}\,{{\it v_1}}^{2}{\it s_2}&{\it s_6}\, \left( {{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {\it s_3}&{\it s_6
}\, \left( {{\it v_2}}^{2}+{{\it v_1}}^{2} \right) {\it s_4}&{\it s_6}\,{\it s_5}\, \left( {{\it v_3}}^{2}+{{\it v_2}}^{2}+{{\it v_1}}^{2} \right) & \left( 1+{{\it v_3}}^{2}+{{\it v_2}
}^{2}+{{\it v_1}}^{2} \right) {{\it s_6}}^{2}\end {array} \right] 
\]
where the variances and covariance for $c_5$ and $c_6$ now depends on $v_2^2$. 

For the correlations, we now compare C($c_i,c_j$), for $i\neq j$ and $i,j\in\{3,4,5,6\}$, 
with these two different models. 
The upper triangle of the correlation matrix for $c_3$, ..., $c_6$
obtained with the model where $p_1$ is parent to $p_3$ is 
\[
\left[ \begin {array}{ccc} 
\textrm{sign}(s_3s_4)\frac{v_1^2+v_2^2}{1+v_1^2+v_2^2} & 
\textrm{sign}(s_3s_5)\frac{v_1^2}{\sqrt{(1+v_1^2)(1+v_1^2+v_3^2)}} &
\textrm{sign}(s_3s_6)\frac{v_1^2}{\sqrt{(1+v_1^2)(1+v_1^2+v_3^2)}} \\
 & \textrm{sign}(s_4s_5)\frac{v_1^2}{\sqrt{(1+v_1^2)(1+v_1^2+v_3^2)}} &
\textrm{sign}(s_4s_6)\frac{v_1^2}{\sqrt{(1+v_1^2)(1+v_1^2+v_3^2)}} \\
 & & \textrm{sign}(s_5s_6)\frac{v_1^2 + v_3^2}{\sqrt{(1+v_1^2+v_3^2)}} \\
\end{array} \right],
\]
and with the model when $p_2$ is parent to $p_3$ it is
\[
\left[ \begin {array}{ccc} 
\textrm{sign}(s_3s_4)\frac{v_1^2+v_2^2}{1+v_1^2+v_2^2} & 
\textrm{sign}(s_3s_5)\frac{v_1^2+v_2^2}{\sqrt{(1+v_1^2+v_2^2)(1+v_1^2+v_3^2)}} &
\textrm{sign}(s_3s_6)\frac{v_1^2+v_2^2}{\sqrt{(1+v_1^2+v_2^2)(1+v_1^2+v_3^2)}} \\
 & \textrm{sign}(s_4s_5)\frac{v_1^2+v_2^2}{\sqrt{(1+v_1^2+v_2^2)(1+v_1^2+v_3^2)}} &
\textrm{sign}(s_4s_6)\frac{v_1^2+v_2^2}{\sqrt{(1+v_1^2+v_2^2)(1+v_1^2+v_3^2)}} \\
 & & \textrm{sign}(s_5s_6)\frac{v_1^2+v_2^2+v_3^2}{1+v_1^2+v_2^2+v_3^2} \\
\end{array} \right].
\]
Comparing these results, when $p_1$ is parent to $p_3$ there is no contribution 
from $v_2^2$ in C($c_5,c_6$) and neither in C($c_i,c_j$), for $i\in\{3,4\}$,
whereas when $p_2$ is parent to $p_3$ there is.

From these examples we can list a set of properties as follows.

* **Property 1**: The marginal variance of $p_i$ is $v_i^2$ plus the variance of its parent.
   * **Property 1a**: For $p_1$ it is equal its conditional variance, $v_1^2$.
   * **Property 1b**: The marginal variance of $p_i$ is $v_i^2$ plus 
   the conditional variance of each one of its ancestors.
* **Property 2**: The marginal variance of $c_i$ is equal $s_i^2(1+V)$,
where $V$ is the variance of its parent.
* **Property 3**: The covariance between $c_i$ and $c_j$ is equal $s_is_jR$, 
where $R$ is the variance of the top indexed parent in the path linking them. 
   * **Property 3a**: If $c_i$ and $c_j$ have a common parent, 
   $R$ is the variance of their parent.
* **Property 4**: C($c_i,c_j$)=sign($s_is_j$)R/$\sqrt{V_{c_i}V_{c_j}}$, 
where $V_{c_i}$ and $V_{c_j}$ are the marginal variances of $c_i$ and $c_j$, respectively. 
   * **Property 4a**: If $c_i$ and $c_j$ have a common parent, 
   the correlation is sign($s_is_j$)$V/(1+V)$ where $V$ is the variance of their parent.
* **Property 5**: Let $k_0$ be the number of parents with no children variable as children, 
   $k_1$ be the number of parents with one children variable as children,
   and $k_2$ be the number of parents with at least two children variables as children.
   The number of distinct absolute correlation values is $k_2+(k_1-k_0)(k_1-k_0-1)/2$.
   * **Property 5a**: If each parent variable has only one children variable as children,
   this number is $k(k-1)/2$.
   * **Property 5b**: The maximum number of distinct absolute correlation values is 
$k(k+1)/2$, achieved if each parent variable is parent to at least two children variables.

Notice that we have defined a model with $n+k$ parameters.
The correlation matrix became structured, as $p<n$, because the maximum number 
of distinct absolute correlation values is $k(k+1)/2$.
The model definition allows the full range of the correlation values, from $-1$ to $1$.
Prior distributions can be considered for the marginal variances 
and for the correlation matrix separately.

## A further example

We now consider more variables and use two models, as shown in Figure 2 bellow.
The difference in on who is the parent to $p_4$. 
We will use it to further illustrate the **Property 2a**.

```{r g4, echo = FALSE,  fig.width = 9, fig.height = 4, out.width="69%", fig.align = 'center', fig.cap = "Two different models for the same number of children and parents."}
g4a <- list(p1 ~ p2 + p3 + c1 + c2, 
            p2 ~ p4 + c3 + c4, p3 ~ p5 + c5 + c6, 
            p4 ~ c7 + c8, p5 ~ c9 + c10)
g4b <- list(p1 ~ p2 + p3 + c1 + c2, 
            p2 ~ c3 + c4, p3 ~ p4 + p5 + c5 + c6, 
            p4 ~ c7 + c8, p5 ~ c9 + c10)
g4ap <- GraphPlot(g4a)
g4bp <- GraphPlot(g4b)
par(mfrow = c(1, 2), mar = c(0, 0, 0, 0))
plot(g4ap$gr, nodeAttrs = g4ap$nAttrs)
legend("topleft", "(E)", bty = "n")
plot(g4bp$gr, nodeAttrs = g4bp$nAttrs)
legend("topleft", "(F)", bty = "n")
```

The model for the DAG in (E) gives a precision matrix whose last five columns are
\[ \scriptsize
\left[ \begin {array}{ccccc} -{{\it s_1}}^{-1}&0&0&0&0\\ -{{\it s_2}}^{-1}&0&0&0&0\\ 0&-{{\it s_3}}^{-1}&0&0&0\\ 0&-{{\it s_4}}^{-1}&0&0
&0\\ 0&0&-{{\it s_5}}^{-1}&0&0\\ 0&0&-{{\it s_6}}^{-1}&0&0\\ 0&0&0&-{{\it s_7}}^{-1}&0\\ 0&0&0&-{{\it s_8}}^{-1}&0
\\ 0&0&0&0&-{{\it s_9}}^{-1}\\ 0&0&0&0&-{{\it s_{10}}}^{-1}\\ {{\it v_1}}^{-2}+2+{{\it v_2}}^{-2}+{{\it v_3}}^{-2}&-{{\it v_2}}^{-2}&-{{
\it v_3}}^{-2}&0&0\\ -{{\it v_2}}^{-2}&{{\it v_2}}^{-2}+2+{{\it v_4}}^{-2}&0&-{{\it v_4}}^{-2}&0\\ -{{\it v_3}}^{-2}&0&{{\it v_3}}^{-2}+2+{{\it v_5}}^
{-2}&0&-{{\it v_5}}^{-2}\\ 0&-{{\it v_4}}^{-2}&0&{{\it v_4}}^{-2}+2&0\\ 0&0&-{{\it v_5}}^{-2}&0&{{\it v_5}}^{-2}+2\end {array} \right] 
\]
while the last five columns for the precision matrix for the model for DAG in (F) are
\[ \scriptsize  
\left[ \begin {array}{ccccc} 
-{{\it s_1}}^{-1}&0&0&0&0\\ -{{\it s_2}}^{-1}&0&0&0&0\\ 0&-{{\it s_3}}^{-1}&0&0&0\\ 0&-{{\it s_4}}^{-1}&0&0
&0\\ 0&0&-{{\it s_5}}^{-1}&0&0\\ 0&0&-{{\it s_6}}^{-1}&0&0\\ 0&0&0&-{{\it s_7}}^{-1}&0\\ 0&0&0&-{{\it s_8}}^{-1}&0
\\ 0&0&0&0&-{{\it s_9}}^{-1}\\ 0&0&0&0&-{{\it s_{10}}}^{-1}\\ {{\it v_1}}^{-2}+2+{{\it v_2}}^{-2}+{{\it v_3}}^{-2}&-{{\it v_2}}^{-2}&-{{
\it v_3}}^{-2}&0&0\\ -{{\it v_2}}^{-2}&{{\it v_2}}^{-2}+2&0&0&0\\ -{{\it v_3}}^{-2}&0&{{\it v_3}}^{-2}+2+{{\it v_4}}^{-2}+{{\it v_5}}^{-2}&-{{\it v_4}
}^{-2}&-{{\it v_5}}^{-2}\\ 0&0&-{{\it v_4}}^{-2}&{{\it v_4}}^{-2}+2&0\\ 0&0&-{{\it v_5}}^{-2}&0&{{\it v_5}}^{-2}+2
\end {array} \right]
\]

To illustrate property about the top parent on a path, we consider 
the upper triangle of the correlation matrix for the last four children variables, $c_7,...,c_{10}$.
For the model defined from (E) we have it as
\[ 
\left[ \begin {array}{ccc} 
 {\frac { \textrm{sign}({\it s_7 s_8}) ( {{\it v_1}}^{2}+{{\it v_2}}^{2}+{{\it v_4}}^{2} )}{( 1+{{\it v_1}}^{2}+{{\it v_2}}^{2}+{{\it v_4}}^{2} )}} &
\frac {\textrm{sign}({\it s_7 s_9}) {{\it v_1}}^{2}}{\sqrt { ( 1+{{\it v_1}}^{2}+{{\it v_2}}^{2}+{{\it v_4}}^{2} ) ( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_5}}^{2} )}} & 
\frac {\textrm{sign}({\it s_7 s_{10}}) {{\it v_1}}^{2}}{\sqrt { ( 1+{{\it v_1}}^{2}+{{\it v_2}}^{2}+{{\it v_4}}^{2} ) ( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_5}}^{2} )}} \\ 
 & \frac {\textrm{sign}({\it s_8 s_9}) {{\it v_1}}^{2}}{\sqrt { ( 1+{{\it v_1}}^{2}+{{\it v_2}}^{2}+{{\it v_4}}^{2} ) ( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_5}}^{2} )}} &
\frac {\textrm{sign}({\it s_8 s_{10}}) {{\it v_1}}^{2}}{\sqrt { ( 1+{{\it v_1}}^{2}+{{\it v_2}}^{2}+{{\it v_4}}^{2} ) ( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_5}}^{2} )}}\\ 
&  &  \frac { \textrm{sign}({\it s_{9} s_{10}}) ( {{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_5}}^{2} ) }{\sqrt { ( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_5}}^{2} ) }} \\
\end {array} \right] 
\]
where  C($c_i,c_j$), for $i\in\{7,8\}$ and $j\in\{9,10\}$ 
increase depends only on $v_1^2$ and 
depends on all $v_j^2$ for $j=\{1,\ldots,5\}$.
With the model defined from (F) we have 
\[ \
\left[ \begin {array}{ccc} 
\frac {\textrm{sign}({\it s_7s_8}) ( {{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_4}}^{2} ) }{( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_4}}^{2} ) } & 
\frac { \textrm{sign}({\it s_7s_9}) ( {{\it v_1}}^{2}+{{\it v_3}}^{2} ) }{\sqrt { ( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_4}}^{2} ) ( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_5}}^{2} ) }} &
\frac { \textrm{sign}({\it s_7s_{10}}) ( {{\it v_1}}^{2}+{{\it v_3}}^{2} ) }{\sqrt { ( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_4}}^{2} ) ( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_5}}^{2} ) }} \\ 
& \frac { \textrm{sign}({\it s_9s_9}) ( {{\it v_1}}^{2}+{{\it v_3}}^{2} ) }{\sqrt { ( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_4}}^{2} ) ( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_5}}^{2} ) }}  &
\frac { \textrm{sign}({\it s_8s_{10}}) ( {{\it v_1}}^{2}+{{\it v_3}}^{2} ) }{\sqrt { ( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_4}}^{2} ) ( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_5}}^{2} ) }} \\
& & \frac {\textrm{sign}({\it s_9s_{10}}) ( {{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_5}}^{2} ) }{ ( 1+{{\it v_1}}^{2}+{{\it v_3}}^{2}+{{\it v_5}}^{2} )}
\end {array} \right] 
\]
where now C($c_i,c_j$), for $i\in\{7,8\}$ and $j\in\{9,10\}$, 
increases if $v_1^2+v_3^2$ (the variance of $p_3$) increases 
and no longer depends on $v_2^2$.
