---
output:
  pdf_document: 
    keep_tex: yes
  html_document:
    df_print: paged
header-includes:
- \usepackage{lscape}
- \usepackage{multicol}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(corGraphs)
```

**Extending the DAG's approach**

- Correlation between parents:
  - correlation between children of different parent is directly modeled
  - correlation between children of same parent is as before

- Add links between children: 
  - full range of correlation between children with link between then
  - correlation between children of different parent as before

- Undirected links model:
   - between children variables, no need of parent nodes
   - is a more general approach
   
## Correlation between parents


In our specific DAG we have $n$ variables of interest, the
children, and $k$ latent variables. 
Specifically, we define let a set of $n$ children, $c_i$, $i$=1,...,$n$, 
and $k$ latents, $p_j$, $j$=1,...,$k$.


```{r graphs, echo = FALSE, fig.width = 12, fig.height = 5, out.width="99%", fig.align = 'center', fig.cap = "Graphs with parents and children."}
g2 <- list(p1 ~ p2 + c1 + c2, p2 ~ c3 + c4)
g2p <- GraphPlot(g2)
g3a <- list(p1 ~ p2 + p3 + c1 + c2, p2 ~ c3 + c4, p3 ~ c5 + c6)
g3b <- list(p1 ~ p2 + c1 + c2, p2 ~ p3 + c3 + c4, p3 ~ c5 + c6)
g3ap <- GraphPlot(g3a)
g3bp <- GraphPlot(g3b)
par(mfrow = c(1, 3), mar = c(0, 0, 0, 0))
plot(g2p$gr, nodeAttrs = g2p$nAttrs)
legend("topleft", "(B)", bty = "n")
plot(g3ap$gr, nodeAttrs = g3ap$nAttrs)
legend("topleft", "(C)", bty = "n")
plot(g3bp$gr, nodeAttrs = g3bp$nAttrs)
legend("topleft", "(D)", bty = "n")
```


* **1st rule**: Joint distribution for parents,
$p_1$, ..., $p_k$, with unit diagonal precision matrix and
off diagonal elements, $i,j$, equal zero if there is no link between 
parents $i$ and $j$ or the conditional correlation parameter.

* **2nd rule**: For $c_i$, $i=1$, ..., $n$, we set
N($p_{c_i}$, 1), where $p_{c_i}$ is the parent of $c_i$.
**(as before)**

The joint distribution for all the $n+k$ variables is
\[ 
p(c_1,\ldots,c_n,p_1,\ldots,p_k) \propto 
 \mathrm{exp} \left\{ \frac{-1}{2} \left[ 
 \sum_{i=1}^n (c_j - p_{c_j})^2 + 
 \mathbf{p}^{T}\mathbf{R}\mathbf{p}
 \right] \right\} 
\]


For the DAGs (B), (C) and (D) in Figure 1 we have
\[ 
\mathbf{R}^{(B)} = \left[ 
\begin{array}{cc} 
a^2 & -r \\ -r & b^2 
\end{array} \right],\;
\mathbf{R}^{(C)} = \left[ 
\begin{array}{ccc} 
a^2 & -r & -s \\ 
-r & b^2 &  \\
-s &  & c^2
\end{array} \right]\; \textrm{ and }\;
\mathbf{R}^{(D)} = \left[ 
\begin{array}{ccc} 
a^2 & -r &  \\ 
-r & b^2 & -s \\
 & -s  & c^2
\end{array} \right]
\]

For the model from (B), the joint distribution has precision
\[
\left[ \begin {array}{cccccc} 
1&0&0&0&-1&0\\ 
0&1&0&0&-1&0\\ 
0&0&1&0&0&-1\\ 
0&0&0&1&0&-1\\ 
-1&-1&0&0&{a}^{-2}+3&-r\\ 
0&0&-1&-1&-r&{b}^{-2}+2
\end {array} \right] 
\]


The covariance for $c_1$, ..., $c_4$ in (B) is
\[ \Sigma =  \frac{1}{b^2a^2+b^2-r^2}
\left[ \begin {array}{cccc} 
{b}^{2}{a}^{2}+2\,{b}^{2}-{r}^{2}&{b}^{2}&r&r\\ {b}^{2}&{b}^{2}{a}^{2}+2\,{b}^{2}-{r}^{2}&r&r\\ r&r&{b}^{2}{a}^{2}+{b}^{2}-{r}^{2}+{a}^{2}+1&{a}^{2}+1\\ r&r&{a}^{2}+1&{b}^{2}{a}^{2}+{b}^{2}-{r}^{2}+{a}^{2}+1\end {array} \right]
\]

For the correlations,
C($c_1$, $c_2$) =
$\frac{{b}^{2}}{{b}^{2}({a}^{2}+2)-{r}^{2}}$,
C($c_1$,$c_3$) = 
C($c_1$,$c_4$) = 
C($c_2$,$c_3$) = 
C($c_2$,$c_4$) = 
$\frac{r}{\sqrt{[{b}^{2}({a}^{2}+2)-{r}^{2}]
[{b}^{2}({a}^{2}+1)-{r}^{2}+{a}^{2}+1]}}$
and 
C($c_3$,$c_4$)=
$\frac {{a}^{2}+1}{{b}^{2}({a}^{2}+1)-{r}^{2}+{a}^{2}+1}$.

If $r=0$ we have
\[ \Sigma = 
\left[ \begin {array}{cccc} 
\frac{{a}^{2}+2}{{a}^{2}+1} & ( {a}^{2}+1 )^{-1} & 0 & 0\\
( {a}^{2}+1 )^{-1}& \frac{{a}^{2}+2}{{a}^{2}+1} & 0 & 0\\ 
 0 & 0 & \frac{{b}^{2}+1}{{b}^{2}}& {b}^{-2} \\ 
 0 & 0 & {b}^{-2} & \frac{{b}^{2}+1}{{b}^{2}}
 \end {array} \right] 
\]

## Add links between children

**as before** but add undirected links between children,
having a joint distribution for children as
\[
p(\mathbf{c}) \propto \exp \left\{
-\frac{1}{2}(\mathbf{c}-\mathbf{p}_\mathbf{c})^T\mathbf{K}
(\mathbf{c}-\mathbf{p}_\mathbf{c})
\right\}. 
\]

### Example: 
Four children with some links and a common parent.

\begin{figure}
\centering
\includegraphics[angle=90,width=0.5\textwidth]{p1c4.jpeg}
\caption{Four children with some links and a common parent.}
\end{figure}

The joint distribution can be written as
\[ p(\mathbf{c},\mathbf{p}) \propto \exp \left\{ -\frac{1}{2} \left(
\left[ \begin{array}{c} c_1$-$p_1 \\ c_2$-$p_1 \end{array} \right]^T
\left[ \begin{array}{cc}
1 & $-$r \\ $-$r & 1
\end{array} \right] 
\left[ \begin{array}{c} c_1$-$p_1 \\ c_2$-$p_1 \end{array} \right] + 
\left[ \begin{array}{c} c_3$-$p_2 \\ c_4$-$p_2 \end{array} \right]^T
\left[ \begin{array}{cc}
1 & $-$s \\ $-$s & 1
\end{array} \right]
\left[ \begin{array}{c} c_3$-$p_2 \\ c_4$-$p_2 \end{array} \right]
+\frac{p_1^2}{q^2}
\right) \right\} .
\]

This gives the joint precision as 
\[
\left[ \begin {array}{ccccc} 
1&-r&0&0&1-r\\ 
-r&1&0&0&1-r\\ 
0&0&1&-s&1-s\\ 
0&0&-s&1&1-s\\ 
1-r&1-r&1-s&1-s&{q}^{-2}+2(2-r-s)
\end {array} \right] 
\]
and the covariance for the children as
\[
\left[ \begin {array}{cccc} 
{\frac {-1-{q}^{2}+{q}^{2}{r}^{2}}{{r}^{2}-1}}&
{\frac {-r-{q}^{2}+{q}^{2}{r}^{2}}{{r}^{2}-1}}&
{q}^{2}&{q}^{2}
\\ {\frac {-r-{q}^{2}+{q}^{2}{r}^{2}}{{r}^{2}-1}}&{\frac {-1-{q}^{2}+{q}^{2}{r}^{2}}{{r}^{2}-1}}&{q}^{2}&{q}^{2}
\\ {q}^{2}&{q}^{2}&{\frac {-1-{q}^{2}+{q}^{2}{s}^{2}}{{s}^{2}-1}}&{\frac {-s+{q}^{2}{s}^{2}-{q}^{2}}{{s}^{2}-1}}
\\ {q}^{2}&{q}^{2}&{\frac {-s+{q}^{2}{s}^{2}-{q}^{2}}{{s}^{2}-1}}&{\frac {-1-{q}^{2}+{q}^{2}{s}^{2}}{{s}^{2}-1}}
\end {array} \right]
\]

The correlations are 
C($c_1$, $c_2$) = $\frac {-r-{q}^{2}+{q}^{2}{r}^{2}}{-1-{q}^{2}+{q}^{2}{r}^{2}}$, 
C($c_1$,$c_3$) = 
C($c_1$,$c_4$) = 
C($c_2$,$c_3$) = 
C($c_2$,$c_4$) = 
$\frac{{q}^{2}}{\sqrt {{\frac {-1-{q}^{2}+{q}^{2}{r}^{2}}{{r}^{2}-1}}}\sqrt {{\frac {-1-{q}^{2}+{q}^{2}{s}^{2}}{{s}^{2}-1}}}}$
and 
C($c_3$,$c_4$) = 
$\frac {-s-{q}^{2}+{q}^{2}{s}^{2}}{-1-{q}^{2}+{q}^{2}{s}^{2}}$.

Correlation instead 

\includegraphics[width=0.5\textwidth]{prpc4}

For the above graph, we consider the joint distribution for the 
parents as 
\[ \left[ \begin{array}{cc} 
1 & -a \\ -a & 1 
\end{array} \right]
\] 
and the ones for the children as before. 

This gives the joint precision as 

