---
title: "Directed tree based correlation model"
author: "Elias T Krainski, Anna Freni-Sterrantino, Denis Rustand, Janet van Niekerk,
  and Håvard Rue"
date: "Started in November 2023, updated in `r format(Sys.time(), '%B, %Y')`"
output:
  pdf_document: 
    keep_tex: yes
  html_document:
    df_print: paged
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = !FALSE)
chinclude <- TRUE
library(corGraphs)
```

# Abstract

In this vignette we show how to work with a particular class of graphs used to introduce a model class for correlation matrices in 
@dtg2024model.

# The model definition

A directed acyclic graph is defined from a set of nodes and 
directed edges linking them. 
Because the edges are directed, there is no closed loop on it.
A tree is a graph where there is only one possible path 
between a pair of nodes.
We consider such two definitions to define a 
directed tree graph - DTG. 
A DTG can be simple view as a tree with directed edges.
This also impose a hierarchy on the nodes.

We classify the nodes of the DTG in parent or children.
The $n$ variables of interest, 
those for we want to model their correlation,
will be classified as children variables.
They are labeled as $c_i$, $i\in\{1,\ldots,n\}$. 
The nodes in the DTG to represent children variables are
leafs of the DTG, which are nodes with an ancestor (parent) 
but with no children.
Each children variable has a directed edge from its parent.
The path between each $c_i$ goes through a set of $m$ parent variables, 
labeled as $p_j$, $j\in\{1,\ldots,k\}$.
The $m$ parent variables are nodes with children nodes.
and some may have parent but will still be classified as parent 
because they have children.

In the \textbf{\textsf{R}} environment we represent the parent 
variables with the letter ``p`` along with an integer number
(``p1``, $\ldots$, ``pn``) and the children variables with 
the letter ``c`` along with an integer number (``c1``, $\ldots$, ``ck``).
We adopt a simple way to specify the parent children representation. 
We consider the ``~`` (tilde) to represent the directed link and 
``+`` (plus) or ``-`` (minus) to append the decendants to a parent.
E.g.: ``p1 ~ p2 + c1 + c2, p2 ~ c3 - c4`` .

## Intial example

Let us consider a correlation model for three variables, 
with one parameter, 
that is the same absolute correlation between each pair 
but the sign may differ.
This can be the case when these variables share the
same latent factor. 
The parent is represented as ``p1``, 
and the children variables as ``c1``, ``c2`` and ``c3``. 
We consider that ``c3`` will be negatively correlated 
with ``c1`` and ``c2``.
The DTG for this case is defined as
```{r graph1, include = chinclude}
g1 <- dtg(p1 ~ c1 + c2 - c3)
g1
summary(g1)
```
where the summary shows their relationship.
The number of children and parent variables 
are obtainted with
```{r dim1}
dim(g1)
```
and we will collect the number of children:
```{r n1}
n1 <- dim(g1)[1]
```

This DTG can be visualized with 
```{r visgraph1, include = chinclude}
plot(g1)
```

From this model definition we will use the methods 
to build the correlation matrix. 
First, we build the precision matrix structure
(that is not yet a precision matrix):
```{r qs1, include = chinclude}
precision(g1)
```
and we can use inform the log of the precision 
for the model parameter with
```{r q, include = chinclude}
q1 <- precision(g1, theta = 0)
q1
```

We can obtain the correlation matrix, which is our 
primarily interest, from the precision matrix. 
However, also have a covariance method to be 
directly applied with
```{r c1, include = chinclude}
cov1a <- variance(q1) 
cov1a[1:n1, 1:n1]
c1 <- cov2cor(cov1a[1:n1, 1:n1])
round(c1, 3)
```
where the first lines/columns
are for the observation level variables.

# Correlation matrix with two parameters

In this example, we model the correlation between four 
variables using two parameters.
We consider ``c1`` and ``c2`` having the same parent, ``p1``
and ``c3`` and ``c4`` having the second parent as parent.
We want to have the correlation between ``c3`` and ``c4``
higher than the correlation between ``c1`` and ``c3``. 
This requires ``p2`` to be children of ``p1``.
The DTG for this is set by
```{r graph2}
g2 <- dtg(p1 ~ p2 + c1 + c2, 
          p2 ~ c3 - c4)
dim(g2)
g2
summary(g2)
```
which can be visualized by
```{r visgraph2}
plot(g2)
```

We can drop the last parent with
```{r drop1}
drop(g2)
```

We now have two parameters and the precision matrix 
can be obtained with:
```{r q2}
q2 <- precision(g2, theta = c(0, 0))
q2
```

The correlation matrix is then obtained with 
```{r c2}
cov2 <- solve(q2)
n2 <- dim(g2)[1]
cov2[1:n2, 1:n2]
c2 <- cov2cor(cov2[1:n2, 1:n2])
round(c2, 3)
```

## Playing with sign

We can change the sign at any edge of the graph. 
The change in the edge of parent to children is 
simpler to interpret. 
We have already seen this case in the one-parameter graph example. 
Let us consider this second example and change the sign 
between the parents and swap it in the second equation:
```{r graph2b}
g2b <- dtg(p1 ~ -p2 + c1 + c2, 
           p2 ~ -c3 + c4)
g2b
```

Now let us compare the covariance between children
from this with the previous one
```{r cov2b}
n2 <- dim(g2b)[1]
q2b <- precision(g2b, theta = c(0, 0))
q2b
cov2b <- solve(q2b)
cov2b[1:n2, 1:n2]
c2b <- cov2cor(cov2b)
round(c2b[1:n2, 1:n2], 3)
```

Therefore, allowing flexibility in an edge of parent to another parent is not useful and will only imply more complexity.
Therefore we will not consider it in the 
`covariance` method. 
However we will consider it in the 
`precision`.
So, please use it with care.

# References

