---
title: "Generalized Kronecker product models"
author: "Elias et. al."
date: "Jan 2024"
output: 
  pdf_document:
classoption: twocolumn
header-includes: 
  \usepackage{hyperref,comment}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We review the usual Kronecker product between two covariances to
model within and between groups correlation, 
and the generalized Kronecker product as an extension to consider 
possible different covariance models within groups. 

The Kronecker product of a order $n\times m$ matrix $\mathbf{A}=(a_{ij})$ 
and a order $r \times s$ matrix $\mathbf{B}=(b_{ij})$ is defined as
\[\left( \begin{array}{cccc}
a_{11}\mathbf{B} & a_{12}\mathbf{B} & \ldots & a_{1m}\mathbf{B} \\
a_{21}\mathbf{B} & a_{22}\mathbf{B} & \ldots & a_{2m}\mathbf{B} \\
\vdots & \vdots  & \ddots  & \vdots \\
a_{n1}\mathbf{B} & a_{n2}\mathbf{B} & \ldots & a_{nm}\mathbf{B} 
\end{array} \right) . \] 
giving the block matrix of order $nr\times ms$.

## Kronecker product model

A Kronecker covariance product model can be used for a random 
vector that can be organized as a matrix with elements 
$\mathbf{x}_{ij}$, $i \in \{1, \ldots, n\}$ and $j \in \{1, \ldots, k\}$. 
Let vec($\mathbf{x}$) a vector of length equal $n\times k$ defined as 
\[\textrm{vec}(\mathbf{x}) = (\mathbf{x}_{.1}^T,\ldots,\mathbf{x}_{.k}^T)^{T}\]
where $\mathbf{x}_{i.}$ is the $i$th row of $\mathbf{x}$ and
$\mathbf{x}_{.j}$ is the $j$th column of $\mathbf{x}$.
Assuming the same e covariance for each column vector $\mathbf{x}_{.j}$,
the covariance between rows $\Sigma_r$, 
and the same covariance for each row vector $\mathbf{x}_{i.}$, 
which is the covariance between the columns $\Sigma_c$,
the covariance of vec($\mathbf{x}$) is given as
\begin{equation}\label{eq:kronecker}
\Sigma_\mathbf{x} = \Sigma_c \otimes \Sigma_r .
\end{equation} 

Usually $\Sigma_c$ or $\Sigma_r$ is set as a correlation matrix.
If this is assumed for $\Sigma_r$,
the variance of the columns are the diagonal elements of $\Sigma_c$.
The computational aspect is appealing as the factorization of 
$\Sigma_{\mathbf{x}}$ is reduced to the factorization of 
$\Sigma_r$ and $\Sigma_c$.

Simulations can be drawn with with 
\begin{equation}\label{eq:rkronecker}
\mathbf{x} = \mathbf{L}_r\mathbf{z}\mathbf{L}_c^T 
\end{equation}
where $\mathbf{z}$ is a $n\times k$ matrix with each entry
$\mathbf{z}_{ij}$ is a independent drawn from 
a random variable with unit variance, 
$\mathbf{L}_c\mathbf{L}_c^T = \Sigma_c$ and 
$\mathbf{L}_r\mathbf{L}_r^T = \Sigma_r$. 
$\mathbf{L}_r$ and $\mathbf{L}_c$ are, respectively,
transformations applied to the columns and rows of $\mathbf{z}$. 

For suitable matrices A, B and C,
\[\textrm{vec}(ABC) = (C^T\otimes A)\textrm{vec}(B),\]
and we can optionally do
\[ 
\textrm{vec}(\mathbf{x}) = (\mathbf{L}_c \otimes \mathbf{L}_r)\textrm{vec}(\mathbf{z}) .
\]
which gives Cov(vec($\mathbf{x}$)) = 
$(\mathbf{L}_c\otimes\mathbf{L}_r) (\mathbf{L}_c\otimes\mathbf{L}_r)^T$.
Notice that the order of the two matrix vector multiplications 
in (\ref{eq:rkronecker}) does not matter.

### Example

A between columns correlation matrix
```{r cc}
cc <- matrix(
  c(1,.9,.7, .9,1,0.8, .7,0.8,1), 
  nrow = 3)
```

A between lines covariance matrix
```{r rr1}
rr1 <- matrix(c(1, 0.5, 0.5, 1), 2)
```

The covariance matrix 
```{r vv1}
vv1 <- kronecker(cc, rr1)
```

Simulations
```{r sim1}
rmvn <- function(n, u) {
    matrix(rnorm(n * ncol(u)), n) %*% u
}
xx <- rmvn(1000, chol(vv1))
cor(cbind(c(xx[, 1], xx[, 3], xx[, 5]),
          c(xx[, 2], xx[, 4], xx[, 6])))
cor(cbind(c(xx[, 1], xx[, 2]), 
          c(xx[, 3], xx[, 4]), 
          c(xx[, 5], xx[, 6])))
```


## Generalized kronecker product model

A generalization of this model is to assume a different covariance 
for each column of $\mathbf{x}$, $\Sigma_{1}, \ldots, \Sigma_{k}$,
withheld the covariance between columns as $\Sigma_c$. 
The covariance for vec($\mathbf{x}$) now depends on how the 
dependence is introduced, from the order of 
the transformations applied to the columns and rows of 
$\mathbf{z}$ to drawn $\mathbf{x}$.

One can first apply the transformation to each column,
introducing correlation between rows first,
and then between columns
\[
\mathbf{x}  = [\mathbf{L}_1\mathbf{z_{.1}} : \ldots : \mathbf{L}_k\mathbf{z}_{.k}] \mathbf{L}_c^T
\]
where $\mathbf{L}_j\mathbf{L}_j^T = \Sigma_j$, $j\in \{1,\ldots,k\}$.
The covariance of vec($\mathbf{x}$) can be computed from 
\[ \begin{array}{rcl}
\textrm{vec}(\mathbf{x}) & = & \textrm{vec}([\mathbf{L}_1\mathbf{z_{.1}} : \ldots : \mathbf{L}_k\mathbf{z}_{.k}] \mathbf{L}_c^T) \\
 & = &  (\mathbf{L}_c\otimes I_n)\textrm{vec}([\mathbf{L}_1\mathbf{z_{.1}} : \ldots : \mathbf{L}_k\mathbf{z}_{.k}]) \\
 & = & (\mathbf{L}_c\otimes\mathbf{I}_n)\textrm{bdiag}(\mathbf{L}_1, \ldots, \mathbf{L}_k)\textrm{vec}(\mathbf{z})
\end{array} \]
which is equal to
\begin{equation}\label{eq:gk1} 
(\mathbf{L}_c\otimes \mathbf{I}_n)
\textrm{bdiag}(\Sigma_1, \ldots, \Sigma_k)(\mathbf{L}_c^T\otimes\mathbf{I}_n)
\end{equation}
where bdiag() binds the matrices diagonally as 
\[\textrm{bdiag}(M_1,\ldots,M_k) = \left(\begin{array}{cccc}
M_1 &  &  &   \\
 & M_2 &  & \\
& & \ddots &  \\
 &  &  & M_k  
\end{array} \right) .
\]

Another way is to first apply the transformation to introduce correlation between columns 
and then between rows of the already cor^2elated columns
\[ \begin{array}{rcl}
\textrm{vec}(\mathbf{x}) & = & \mathrm{vec}([\mathbf{L}_1(\mathbf{z}\mathbf{L}_c^T)_{.1} : \ldots : \mathbf{L}_k(\mathbf{z}\mathbf{L}_c^T)_{.k}]  \\
 & = & \mathrm{bdiag}(\mathbf{L}_1,\ldots,\mathbf{L}_k)\mathrm{vec}(\mathbf{z}\mathbf{L}_c^T) \\
 & = & \mathrm{bdiag}(\mathbf{L}_1,\ldots,\mathbf{L}_k)(\mathbf{L}_c^T\otimes\mathbf{I}_n)\mathrm{vec}(\mathbf{z})
 \end{array} \]
giving the covariance matrix as
\begin{equation}\label{eq:gk2}
\mathrm{bdiag}(\mathbf{L}_1,\ldots,\mathbf{L}_k)(\Sigma_c\otimes\mathbf{I}_n)
\mathrm{bdiag}(\mathbf{L}_1^T,\ldots,\mathbf{L}_k^T)
\end{equation}
which is different from (\ref{eq:gk1}).

### Example 

Let the between columns correlation be
\[ \Sigma_c = \left( \begin{array}{cc}
 1 & r \\
 r & 1
\end{array} \right),
\] 
and define two between lines correlation matrices as
\[
\Sigma_1 = \left( \begin{array}{cc}
 1 & a \\
 a & 1
\end{array} \right)\;\;\; \textrm{and}\;\;\; 
\Sigma_2 = \left( \begin{array}{cc}
 1 & b \\
 b & 1
\end{array} \right) \]

First applying the correlation between rows gives
the (upper triangle) of the correlation matrix as
\[ \left(
\begin {array}{ccc} 
a&r&ra\\ &ra&r \\ 
  & & b+r^2(a-b) 
\end{array} \right) \]
that preserves the between columns correlation, 
because the elements $1,3$ and $2,4$ are equal $r$,
but not the between rows correlations, 
because the element $3,4$ is not $b$.

First applying the correlation between columns gives
the (upper triangle) of the correlation matrix as
\[  \left(
\begin{array}{ccc} 
a&r&br\\ 
&ar&r[ab+\sqrt {1-a^2}\sqrt {1-b^2}] \\ 
& &b
\end{array} \right) \]
that preservers the between rows correlations,
because element $1,2$ is $a$ and element $3,4$ is $b$,
but not the between columns correlation, 
because element $3,4$ is not $r$.

### Numerical example

A function to compute the covariance
```{r fn1, echo=!FALSE}
gkpc <- function(A, B) {
  if(is.matrix(A) & is.matrix(B)) {
    nr <- nrow(A)
    nc <- ncol(B)
    r <- kronecker(A, B)
  }
  if(is.matrix(A) & is.list(B)) {
    stopifnot(is.list(B))
    nr <- nrow(A)
    nc <- nrow(B[[1]])
    kd <- diag(nc)
    l <- chol(A)
    r <- kronecker(t(l), kd) %*%
        Matrix::bdiag(B) %*%
        kronecker(l, kd)
  } 
  if(is.list(A) & is.matrix(B)) {
    nr <- nrow(B)
    A <- lapply(A, chol)
    nc <- nrow(A[[1]])
    dd <- diag(nc)
    r <- Matrix::bdiag(lapply(A, t)) %*%
        kronecker(B, dd)  %*%
        Matrix::bdiag(A)
  }
 rownames(r) <- colnames(r) <- paste0(
  "r", rep(1:nr, nc), 
  "c", rep(1:nc, each = nr))
  return(as.matrix(r))
}
```

Define a list of matrices with covariance between lines for each column
```{r rr2}
rr2 <- list(matrix(c(1, 0.5, 0.5, 1), 2), 
           matrix(c(1, -0.5, -0.5, 1), 2),
           matrix(c(1, 0.2, 0.2, 1), 2))
```

The covariance when first applying the transformation to each column is
```{r V1}
V1 <- gkpc(cc, rr2) ## between rows first
round(V1 * 100)
```

The covariance when first applying the transformation between columns
```{r V2}
V2 <- gkpc(rr2, cc) ## between columns first
round(V2 * 100)
```

Simulations
```{r sim2}
set.seed(1)
xx1 <- rmvn(1000, chol(V1))
set.seed(1)
xx2 <- rmvn(1000, chol(V2))
cor(cbind(c(xx1[, 1], xx1[, 2]),
          c(xx1[, 3], xx1[, 4]),
          c(xx1[, 5], xx1[, 6])))
cor(cbind(c(xx2[, 1], xx2[, 2]),
          c(xx2[, 3], xx2[, 4]),
          c(xx2[, 5], xx2[, 6])))
list(cov(xx1[, 1:2]), 
     cov(xx1[, 3:4]), 
     cov(xx1[, 5:6]))
list(cov(xx2[, 1:2]), 
     cov(xx2[, 3:4]), 
     cov(xx2[, 5:6]))
```

