---
title: "Graph based correlation models"
author: "Anna Freni-Sterrantino, Denis Rustand, Janet van Niekerk, Elias T Krainski,
  and Håvard Rue"
date: "November, 2023"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(corGraphs)
```

# Abstract

In this vignette we introduce the models for correlation matrices 
based on a graph from a Direct Acyclic Graph - DAG.

# The model definition

The definition is through a graph which is basically nodes and edges, 
where there is two kind of nodes, one to represent observation level 
variables and other to represent latent level variables. 
The particularity of the graphs we consider start by defining that
observation level variables are represented by nodes 
which are always at the end of the graph, or are always leaves.
The latent variables are represented by nodes which are always 
parent either latent variables or to observation level variables.
Therefore, we consider the latent variables identified as parents and the 
observation level variables as children, even though a parent may be 
children to another parent as well. 

We should mention that the "children" are not restricted to be 
representing observations but can also represent
parameters in the linear predictor of a model.

Our convention is to identify the parents 
by the letter "p" and an integer number 
from one to the number of parents,
and the children by the letter "c" 
and and an integer number from one to the total number of children.
Thus $p1$ is parent one and $p2$ is parent two,
representing two latent variables, 
and $c1$, $c2$ and $c3$ 
are children one, children two and children three,
respectively, representing three observation level variables. 

## Intial example

In order to set the notation, let us consider three observation variables 
and only one common correlation parameter. 
The graph to represent this assumption has only one parent.
We specify this by
```{r graph1}
g1 <- list(p1 ~ c1 + c2 + c3)
nc <- 3
```
which means that $p1$ is parent to children $c1$, $c2$ and $c3$.

We can visualize it by 
```{r visgraph1}
g1p <- GraphPlot(g1)
plot(g1p$gr, nodeAttrs = g1p$nAttrs)
```

From this model definition we will use the methods 
IN THE PAPER 
to build the correlation matrix. 
First, we build the precision matrix, 
assuming $q_1$ = 1:
```{r q1}
q1 <- dag_precision(g1, theta = c(0))
q1
```

The correlation matrix is then obtained with 
```{r c1}
cov1a <- solve(q1)
cov1a[1:nc, 1:nc]
c1 <- cov2cor(cov1a)
round(c1[1:nc, 1:nc], 3)
```
where the first lines/columns
are for the observation level variables.

## Allowing negative correlations

The signal of the correlation is defined in the formula. 
If we want the correlation between $c1$ or $c2$ and $c3$ 
equal minus the correlation between $c1$ and $c2$ we define
```{r graph1b}
g1b <- list(p1 ~ c1 + c2 - c3)
```

The correlation matrix is then obtained with 
```{r c1b}
q1b <- dag_precision(g1b, theta = 0)
q1b
cov1b <- solve(q1b)
cov1b[1:nc, 1:nc]
c1b <- cov2cor(cov1b)
round(c1b[1:nc, 1:nc], 3)
```

If we want to have the correlation between 
$c1$ and $c2$ and $c1$ and $c3$ be minus the 
correlation between $c2$ and $c3$ we do
```{r graph1c}
g1c <- list(p1 ~ -c1 + c2 + c3)
```

The correlation matrix is then obtained with 
```{r c1c}
q1c <- dag_precision(g1c, theta = 0)
q1c
cov1c <- solve(q1c)
cov1c[1:nc, 1:nc]
c1c <- cov2cor(cov1c)
round(c1c[1:nc, 1:nc], 3)
```

As we have three edges, either one can have different 
signal to the other two, but same absolute value.

# Correlation matrix with two parameters

We now implement a correlation matrix model with two parameters, 
considering two latent factors.
We consider the example in section 3.2 of the paper.

We have three variables, observations, which are correlated. 
Following the previous definition, we will represent it by $c1$, $c2$ and $c3$. 
Suppose that the correlation between $c1$ and $c3$ is equal to the 
correlation between $c2$ and $c3$.
This corresponds to having $c1$ and $c2$ under the same branch.
The graph to represent this model assumption, based in our convention, 
can be represented by the following set of two equations
```{r graph2}
g2 <- list(p1 ~ p2 + c3, 
           p2 ~ c1 + c2)
```
which means that $p1$ is parent to $c1$ and $c2$ and that
$p2$ is parent to $p1$ and $c3$. 

We can visualize it by 
```{r visgraph2}
g2p <- GraphPlot(g2)
plot(g2p$gr, nodeAttrs = g2p$nAttrs)
```

From this model definition we will use the methods 
IN THE PAPER 
to build the correlation matrix. 
First, we build the precision matrix, 
assuming $q_1$ = $q_2$ = 1:
```{r q2}
q2 <- dag_precision(g2, theta = c(0, 0))
q2
```

The correlation matrix is then obtained with 
```{r c2}
cov2 <- solve(q2)
cov2[1:nc, 1:nc]
c2 <- cov2cor(cov2)
round(c2[1:nc, 1:nc], 3)
```
where the first lines/columns 
are for the observation level variables.

Notice that we can have more children under either parent 
keeping the same correlation model, with two parameters. 

## Negative correlation

We can change the signal at any edge of the graph. 
The change in the edge of parent to children is 
simpler to interpret. 
We have already seen this case in the one-parameter graph example. 

In this second graph example, the change in the 
signal of edge between $p1$ and $p2$ or the
edge between $p1$ and $c3$ imply the same change 
in the correlation matrix, as we can see next
```{r graph2b}
g2b <- list(p1 ~ -p2 + c3, 
            p2 ~ c1 + c2)
q2b <- dag_precision(g2b, theta = c(0, 0))
q2b
cov2b <- solve(q2b)
cov2b[1:nc, 1:nc]
c2b <- cov2cor(cov2b)
round(c2b[1:nc, 1:nc], 3)
```

```{r graph2c}
g2c <- list(p1 ~ p2 - c3, 
            p2 ~ c1 + c2)
q2c <- dag_precision(g2c, theta = c(0, 0))
q2c
cov2c <- solve(q2c)
cov2c[1:nc, 1:nc]
c2c <- cov2cor(cov2c)
round(c2c[1:nc, 1:nc], 3)
```

However, the change in an edge of parent to another parent
could have more complex implications if these
parents have parents as children. 
We will see this in the next example.

# Correlation matrix with three parameters

Here we consider five correlated observation variables
represented as $c1$, $c2$, $c3$, $c4$ and $c5$ in the graph.
We assume that the correlation between $c1$ and $c2$ 
is equal to the correlation between $c1$ and $c3$, 
and, of course, equal the correlation between $c2$ and $c3$.
This restriction also implies that the correlation between 
any of these three first variables to $c4$ is the same. 
A further assumption is if we assume the correlation between
$c1$ and $c4$ is the same as between $c1$ and $c5$.

## Three different correlation values

If we assume that the correlation between $c1$
(or any of the first variables) and $c4$ to be 
the same as between $c1$ and $c5$.
This implies that $c4$ and $c5$ are children to the same parent.

The graph to represent this model assumption
can be represented by the following set of equations
```{r graph3}
g3 <- list(p1 ~ p2 + p3, 
           p2 ~ c1 + c2 + c3,
           p3 ~ c4 + c5)
nc <- 5
```
that is, $p1$ is parent to $p2$ and $p3$, 
$p2$ is parent to $c1$, $c2$ and $c3$, 
and $p3$ is parent to $c4$ and $c5$.

We can visualize it by 
```{r visgraph3}
g3p <- GraphPlot(g3)
plot(g3p$gr, nodeAttrs = g3p$nAttrs)
```

From this model definition we will use the methods 
IN THE PAPER 
to build the correlation matrix. 
First, we build the precision matrix, 
assuming $q_1$ = 1, $q_2$ = 2 and $q_3$ = 3.
```{r q3}
q3 <- dag_precision(g3, theta = log(1:3))
q3
```

The correlation matrix is then obtained with 
```{r c3}
cov3 <- solve(q3)
cov3[1:nc, 1:nc]
c3 <- cov2cor(cov3)
round(c3[1:nc, 1:nc], 3)
```
where the first lines/columns represents 
are for the observation level variables.

## Negative correlations

Here we explore negative correlations by changing 
the signal of edge between $p1$ and $p2$.

```{r graph3b}
g3b <- list(p1 ~ -p2 + p3, 
            p2 ~ c1 + c2 + c3,
            p3 ~ c4 + c5)
q3b <- dag_precision(g3b, theta = log(c(1,2,3)))
q3b
cov3b <- solve(q3b)
cov3b[1:nc, 1:nc]
c3b <- cov2cor(cov3b)
round(c3b[1:nc, 1:nc], 3)
```

Now, at two different edges simultaneously, 
by also changing the signal of edge between $p2$ and $c5$

```{r graph3c}
g3c <- list(p1 ~ -p2 + p3, 
            p2 ~ c1 + c2 + c3,
            p3 ~ c4 - c5)
q3c <- dag_precision(g3c, theta = c(0, 0, 0))
q3c
cov3c <- solve(q3c)
cov3c[1:nc, 1:nc]
c3c <- cov2cor(cov3c)
round(c3c[1:nc, 1:nc], 3)
```

## Four different correlation values

With the same number of latent variables
but now a different graph.
If we make $p1$, the main parent, also to have 
observation variables as children, 
there will be then three possible correlation 
values between observation variables.

We can make $c3$ as children of $p1$.
```{r graph4}
g4 <- list(p1 ~ p2 + p3 + c3, 
           p2 ~ c1 + c2,
           p3 ~ c4 + c5)
g4p <- GraphPlot(g4)
plot(g4p$gr, nodeAttrs = g4p$nAttrs)
```

```{r c4}
q4 <- dag_precision(g4, theta = log(1:3))
q4
cov4 <- solve(q4)
cov4[1:nc, 1:nc]
c4 <- cov2cor(cov4)
round(c4[1:nc, 1:nc], 3)
```

We can also allow negative correlations. 
For example, change the signal of the edge
between $p3$ and $p2$ and
between $p2$ and $c5$. 

```{r graph4b}
g4b <- list(p1 ~ p2 - p3 + c3, 
            p2 ~ c1 + c2,
            p3 ~ c4 - c5)
q4b <- dag_precision(g4b, theta = log(1:3))
q4b
cov4b <- solve(q4b)
cov4b[1:nc, 1:nc]
c4b <- cov2cor(cov4b)
round(c4b[1:nc, 1:nc], 3)
```

## Five different correlations values

Changing further, we can make $p1$ children to $p2$.

```{r graph5}
g5 <- list(p1 ~ p2 + c4 + c5, 
           p2 ~ p3 + c3,
           p3 ~ c1 + c2)
g5p <- GraphPlot(g5)
plot(g5p$gr, nodeAttrs = g5p$nAttrs)
```

```{r c5}
q5 <- dag_precision(g5, theta = log(1:3))
q5
cov5 <- solve(q5)
cov5[1:nc, 1:nc]
c5 <- cov2cor(cov5)
round(c5[1:nc, 1:nc], 3)
```

Changes in signals now can make interesting behavior.
Let us change the signal of the edges 
$p1$ to $c1$, $p3$ to $p2$ and $p2$ to $p5$.

```{r graph5b}
g5b <- list(p1 ~ -p2 + c4 + c5, 
            p2 ~ p3 + c3,
            p3 ~ -c1 + c2)
```

```{r c5b}
q5b <- dag_precision(g5b, theta = log(1:3))
q5b
cov5b <- solve(q5b)
cov5b[1:nc, 1:nc]
c5b <- cov2cor(cov5b)
round(c5b[1:nc, 1:nc], 3)
```

