---
title: "An informative prior for correlation matrix"
output:
  - rmarkdown::pdf_document
  - rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An informative prior for correlation matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amsfonts}
  - \usepackage{amssymb}
  - \usepackage{graphicx}
  - \usepackage{mathtools}
  - \usepackage{calc}
  - \newcommand{\bx}{\textbf{x}}
  - \newcommand{\bmu}{\boldsymbol{\mu}}
  - \newcommand{\bgamma}{\boldsymbol{\gamma}}
  - \newcommand{\btheta}{\boldsymbol{\theta}}
  - \newcommand{\bxi}{\boldsymbol{\xi}}
  - \newcommand{\brho}{\boldsymbol{\rho}}
  - \newcommand{\bvphi}{\boldsymbol{\varphi}}
  - \newcommand{\bA}{\mathbf{A}}
  - \newcommand{\bB}{\mathbf{B}}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align='center', out.width = "99%",
  fig.width = 10, fig.height = 7)
library(graphpcor)
library(INLA)
set.seed(1)
```

# Abstract

We illustrate the use of the informative prior for 
correlation matrix. 
This prior is constructed by penalizing deviations 
from a base correlation matrix, 
applying the idea from @simpsonetal2017.
This prior may be **un**informative for small data size, 
as shown.

# Introduction

We assume that a $p$-dimensional random vector $\mathbf{X}$
is Normally distributed and want to consider its covariance
matrix $\mathbf{V}$.
This covariance matrix is written as
\[ \Sigma = \mathbf{S} \mathbf{C}(\btheta) \mathbf{S} \]
where the matrix $\mathbf{S}$ is a diagonal matrix whose
elements $\mathbf{S}_{jj} = \sigma_j$ is the
standard deviation of $\mathbf{X}_j$.

## Parametrization

We consider that $\mathbf{C}(\btheta)$ is parametrized
considering the canonical partial correlation - CPC,
where the $\btheta$ is a vector of dimension $m=n(n-1)/2$.

The CPC parametrization, @lkj2009,
we define $r[k] = \textrm{tanh}(\theta[k])$, 
for $k=\{1, \ldots,m\}$, and the two 
$p\times p$ matrices
\[ \bA = \left[
     \begin{array}{ccccc}
       1 & & & & \\
       r_1 & 1 & & & \\
       r_2 & r_p & 1 & & \\
       \vdots & \vdots & \ddots & \ddots & \\
       r_{p-1} & r_{2p-3} & ... & r_m & 1
     \end{array} \right]
     \textrm{ and } \bB = \left[
     \begin{array}{ccccc}
       1 & & & & \\
       \sqrt{1-r_1^2} & 1 & & & \\
       \sqrt{1-r_2^2} & \sqrt{1-r_p^2} & 1 & & \\
       \vdots & \vdots & \ddots & \ddots & \\
       \sqrt{1-r_{p-1}^2} & \sqrt{r_{2p-3}^2} & ... & \sqrt{1-r_m^2} & 1
     \end{array} \right]  \] 
The matrices $\bA$ and $\bB$ are then used to build the Cholesky factor of the correlation matrix, given as
\[ \mathbf{L} = \left[
     \begin{array}{ccccc}
       1 & 0 & 0 & ... & 0\\
       \bA_{2,1} & \bB_{2,1} & 0 & ... & 0\\
       \bA_{3,1} & \bA_{3,2}\bB_{3,1} & \bB_{3,1}\bB_{3,2} & & \vdots \\
       \vdots & \vdots & \ddots & \ddots & 0\\
       \bA_{p,1} & \bA_{p,2}\bB_{p,1} & ... &
       \bA_{p,p-1}\Pi_{k=1}^{p-1}\bB_{p,k} & \Pi_{k=1}^{p-1}\bB_{p,k}
     \end{array} \right]  \]
     
The determinant of the correlation matrix is
\[  \Pi_{i=2}^p \Pi_{j=1}^{i-1}\bB_{i,j} = \Pi_{i=2}^p \mathbf{L}_{i,i}. \]

The prior distribution for $\btheta$ is 
\[ 
\pi_{\btheta}(\btheta|\lambda) = 
\pi_{\bxi}(\bxi|\lambda)|\textrm{det}(\textbf{I}(\btheta_0)^{1/2})|,
\]
where $\lambda$ is a (scalar) prior parameter, 
$\textbf{I}(\btheta_0)$ is the Hessian of the KLD($\btheta,\btheta_0$) around $\btheta_0$,
and $\bxi = \textbf{I}(\btheta_0)^{1/2}(\btheta-\btheta_0)$.
See the PC-multivariate vignette for details.

We will consider some values for $\lambda$

```{r lambdas}
lambdas <- c(0.1, 1, 10, 100, 1000)
nlambs <- length(lambdas)
```

# Example with p = 2, m = 1

```{r p2m1}
p <- 2
(m <- p*(p-1)/2)
```

For each value of the parameter we define the `cgeneric` model,
implemented with model (name) as `pc_correl`.
Notice that $n=p$ and that we did set the base model 
as $\theta=0$ (no correlation).

## The prior

We evaluate the integral of the density (for $\theta$)
by applying `prior()` to the `generic` model as
```{r iprior, eval = FALSE}
plot(function(th) 
  exp(prior(Cmodel, theta = th)), 
  from = a, to = b)
```
where `a` and `b` are the limits.
We do it with different lambdas
```{r inttheta}
for(i in 1:nlambs) {
    Cmodel <- cgeneric(
        model = "pc_correl",
        n = p,
        lambda = lambdas[i],
        base = 0,
        useINLAprecomp = FALSE)
    print(integrate(
      function(th) exp(prior(Cmodel, theta = th)), 
      lower = -5, upper = 5)) 
}
```

Note: For small $\lambda$ it did not get to one
as the prior has relevant mass 
outside the domain considered.

Similarly, we can also visualize 
the density, with respect to $\theta$.
We also consider the posterior from 
fitting the model to an empty data, 
thus it should be equal the prior, with INLA.

Set up a empty data set and fix the precision of a 
Gaussian likelihood to a high precision with
```{r d1p}
fi <- y ~ 0 + f(i, model = Cmodel)
d1p <- data.frame(y = NA, i = 1:p)
cfam <- list(hyper = list(prec = list(
  initial = 10, fixed = TRUE)))
```

We visualize both, the prior directly computed with `prior()` and the one from `inla()`, 
for different prior parameter. 

```{r vizthn2m1}
par(mfrow = c(2, 3), mar = c(3,3,0.1,0.1), mgp = c(1.5, 0.5, 0), bty = "n")
for(i in 1:nlambs) {
    Cmodel <- cgeneric(
        model = "pc_correl",
        n = p,
        lambda = lambdas[i],
        useINLAprecomp = FALSE)
    ifit <- inla(
        formula = fi, data = d1p, control.family = cfam
    )
    pm <- ifit$marginals.hyperpar[[1]]
    plot(function(th) exp(prior(Cmodel, theta = th)),
         from = min(pm[, 1])*2, to = max(pm[, 1]) * 2, 
         xlab = expression(theta), 
         ylab = expression(pi[theta]))
    lines(pm, col = 2, lty = 3)
}
```

The differences for small $\lambda$ are due to 
lack of numerical refinements for such cases. 

## The $\theta$ and correlation map

We can use the function `basecor` evaluate $\theta$ with
```{r basecor1}
basecor(-1, p)
basecor(0, p)
basecor(1, p)
```
which include more than what we need.
Therefore we can just do
```{r cholcor1}
tcrossprod(cholcor(-1, p))
```
or invert the `prec` evaluated over a `cgeneric` object for the
`pc_correl` with
```{r prec1}
solve(prec(Cmodel, theta = -1))
```

We can use the `inla.hyperpar.sample` to drawn samples from
the $\theta$, compute the correlation and visualize it
as the prior for the correlation next.

```{r vizCn2m1}
par(mfrow = c(2, 3), mar = c(3,3,0.1,0.1), mgp = c(1.5, 0.5, 0), bty = "n")
for(i in 1:nlambs) {
    Cmodel <- cgeneric(
        model = "pc_correl",
        n = p,
        lambda = lambdas[i],
        useINLAprecomp = FALSE)
    ifit <- inla(
        formula = fi, data = d1p, control.family = cfam
    )
    thsim <- inla.hyperpar.sample(n = 5000, result = ifit)
    csim <- sapply(1:nrow(thsim), function(i)
      tcrossprod(cholcor(thsim[i,], p = p))[2,1])
    hist(csim, 100, freq = FALSE, 
         main = "", xlab = expression(rho))
}
```

The histograms for the correlation drawn with different 
$\lambda$ are narrowed toward 0 (set as base) as 
$\lambda$ increases. 
See also the PC-multivariate vignette for the exact posterior.


# Example with p = 3, m = 3

Consider that we have a correlation matrix $\mathbf{C}_0$: 
```{r c3}
c0 <- matrix(c(1,     .8,  -.5,
               0.8,    1,  -.4,
               -.5,  -.4,   1), 3)
c0
(p <- ncol(c0))
(m <- p*(p-1)/2)
```

We now want to compute the $\btheta$, under the CPC
parametrization.
The `basecor` function does this and
also compute the Hessian around
(see PC-multivariate vignette for details) with

```{r base3}
b0 <- basecor(c0)
b0
```

Collect the parameter for this base correlation matrix
```{r theta3}
(theta3 <- b0$theta)
```

Let us visualize the (3 dimensional) density.
We first define the `cgeneric` model where
the base model is $\mathbf{C}_0$ with

```{r cg3}
cmodel <- cgeneric(
    model = "pc_correl",
    base = c0,
    lambda = 5,
    useINLAprecomp = FALSE)

cmodel

graph(cmodel)

initial(cmodel)

prec(cmodel, theta = theta3)

all.equal(as.matrix(solve(prec(cmodel, theta = theta3))),
          c0)

```

The prior can be evaluated at one $\btheta$ (vector) value with
```{r prior3_1}
prior(cmodel, theta = theta3)
```
or at several (one at each column)
```{r prior3_2}
prior(cmodel, theta = cbind(theta3, rnorm(m)))
```


We now set a 3d grid for $\btheta$ with
```{r th3dgrid}
hth <- 0.02; hth2 <- hth^2
str(sths <- lapply(1:3, function(i)
    seq(-2+hth/2, 2-hth/2, hth) + theta3[i]))
(nsth <- length(sths[[1]]))
ncol(th3 <- t(expand.grid(sths)))/1e6
```
evaluate the prior and organize it into an array with
```{r p3}
p3 <- array(prior(cmodel, theta = th3), rep(nsth, 3))
str(p3)
```

Checks:
```{r p3check}
sum(exp(p3)*(hth^3))

hth2 <- hth * hth
sum(apply(exp(p3)*hth2, 2, sum) * hth)
sum(apply(exp(p3)*hth2, 2, sum) * hth)
sum(apply(exp(p3)*hth2, 3, sum) * hth)

```

We will visualize every 2 dimensions
(at the mid value of the other) and
each marginal with

```{r visp3}
par(mfrow = c(2, 2), mar = c(3, 3, 0.5, 0.5),
    mgp = c(2, 0.5, 0), bty = "n")
image(sths[[1]], sths[[2]], p3[,,nsth/2])
contour(sths[[1]], sths[[2]], p3[,,nsth/2], add = TRUE, nlevels = 5)
image(sths[[1]], sths[[3]], p3[,nsth/2,])
contour(sths[[1]], sths[[3]], p3[,nsth/2,], add = TRUE, nlevels = 5)
image(sths[[2]], sths[[3]], p3[nsth/2,,])
contour(sths[[2]], sths[[3]], p3[nsth/2,,], add = TRUE, nlevels = 5)


plot(sths[[1]], apply(exp(p3)*hth2, 1, sum),
     type = "l", xlab = "", ylab = "density",
     xlim = range(unlist(sths)))
lines(sths[[2]], apply(exp(p3)*hth2, 2, sum), col = 2)
lines(sths[[3]], apply(exp(p3)*hth2, 3, sum), col = 3)
legend("topleft", bty = "n", lty = 1, col = 1:3,
       c(expression(theta[1]), expression(theta[2]), expression(theta[3])))
```

We now consider different prior parameters
fit with no data, to get back the prior 
in the `inla` output. 
This will be then used to sample from the joint 
posterior (which is equal the prior in this case) 
for $\theta$. 
For each sample we can compute the correlation matrix
in order to have a joint sample for the correlations.
To do that, we use the `cholcor()` function as 
in the following example:
```{r cholcor}
lR <- cholcor(theta3, p)
lR
all.equal(c0, tcrossprod(lR))
```

Labels for the plots
```{r prep3}
pthlabs <- c(
    expression(pi[theta[1]](theta[1]~"|"~lambda)),
    expression(pi[theta[2]](theta[2]~"|"~lambda)),
    expression(pi[theta[3]](theta[3]~"|"~lambda))
)
rholabs <- c(
    expression(rho[2~","~1]),
    expression(rho[3~""~1]),
    expression(rho[3~""~2])
    )
prholabs <- c(
    expression(pi[rho[2~","~1]](rho[2~","~1]~"|"~lambda)),
    expression(pi[rho[3~","~1]](rho[3~","~1]~"|"~lambda)),
    expression(pi[rho[3~","~2]](rho[3~","~2]~"|"~lambda))
)

il3 <- which(lower.tri(diag(p)))
il3
```

Visualize the prior of $\btheta$ and of the correlations,
for each $\lambda$ case, and having the base model as 
$\mathbf{C}_0$. 

```{r p3cor, fig.width = 18, fig.height = 10}
par(mfrow = c(nlambs, 6), mar = c(4,4,0,0), mgp = c(2,0.5,0))
for(il in 1:nlambs) {
    lamb <- lambdas[il]
    Cmodel <- cgeneric(
        model = "pc_correl", n = p, lambda = lamb,
        base = theta3, useINLAprecomp = FALSE)
    itest <- inla(
        formula = fi, control.family = cfam,
        data = d1p
    )
    for(i in 1:3) {
        plot(itest$marginals.hyperpar[[i]], pch = 19, type = 'b',
             xlab = as.expression(bquote(theta[.(i)])),
             ylab = pthlabs[[i]], main = "")
        if(i==1)
            legend("topleft", bty = "n",
                   as.expression(bquote(lamda == .(lamb))))
    }
    hprior3 <- inla.hyperpar.sample(10000, itest)
    cprior3 <- apply(hprior3, 1, function(x)
        tcrossprod(cholcor(x, p=p))[il3])
    for(i in 1:3) {
        hist(cprior3[i, ], seq(-.1, .1, 0.005)+c0[il3[i]], freq = FALSE,
             main = "",
             xlab = rholabs[[i]],
             ylab = prholabs[[i]])
        abline(v = c0[il3[i]], col = 2, lty = 2, lwd = 2)
    }
}
```

As expected, by increasing $\lambda$, 
the prior gets narrower around the values from 
$\mathbf{C}_0$.

## Fit to data

We now want to see the effect of the prior on the posterior
considering different data size.
We drawn samples with $\mathbf{C})$ 
for two different sample size with
```{r data3}
nsims <- c(n50=50, n500=500)
xxs <- lapply(nsims, function(ns)
    tcrossprod(matrix(rnorm(ns * p), ns), lR))
```

Compute the sample correlations (to compare later)
```{r c3sample}
Rs <- lapply(xxs, cor)
Rs
```

Evaluate the $\btheta$ at the sample correlation,
to overlay in the plots as "observed" $\btheta$:
```{r otheta3}
Bs <- lapply(Rs, basecor, p = p)
(oths <- sapply(Bs, function(b) b$theta))
```

Some setup to organize the data for INLA and
the formula considering each sample as replicate.
```{r setup3}
datf3 <- lapply(1:2, function(i)
    data.frame(
        y = as.vector(xxs[[i]]),
        i = rep(1:p, each = nsims[i]),
        r = rep(1:nsims[i], p)))
names(datf3) <- names(nsims)
fr <- y ~ 0 + f(i, model = cmodel, replicate = r)
```

For each $\lambda$, we have 2 
(one for each of the 2 samples) times 6 plots
(one for each $\btheta$ and one for each correlation), 
each one of these plots shown at each row of the figure.

Now we consider that the base model is no correlation!
We set this as we want to see how the sample size is 
able to move the posterior away from the prior 
correlation value (which is zero) for 
different $\lambda$, one at each each row of the figure.

```{r vis3d2, fig.width = 12, fig.height = 10}
par(mfrow = c(nlambs, 6), mar = c(4,4,0,0), mgp = c(2,0.5,0))
for(lamb in lambdas) {
    cmodel <- cgeneric("pc_correl", n = p, lambda = lamb,
                       useINLAprecomp = FALSE)
    p3a <- array(exp(prior(cmodel, theta = th3)), rep(nsth, 3))
    itests <- lapply(datf3, function(ddf)
        inla(formula = fr,
             data = ddf,
             control.family = cfam
             )
        )
    for(i in 1:3) {
        pm1 <- itests[[1]]$marginals.hyperpar[[i]]
        pm2 <- itests[[2]]$marginals.hyperpar[[i]]
        xlm <- range(theta3[i],
                     Bs[[1]]$theta[i],
                     Bs[[2]]$theta[i],
                     pm1[, 1], pm2[, 1])
        ylm <- c(0, max(pm1[, 2], pm2[, 2]))
        plot(pm1, type = "l", col = 2, lwd = 2,
             xlim = xlm, ylim = ylm,
             xlab = as.expression(bquote(theta[.(i)])),
             ylab = pthlabs[[i]], main = "")
        lines(pm2, col = 3, lwd = 2)
        lines(sths[[i]], apply(p3a*hth2, i, sum), lty = 3, lwd = 3)
        abline(v = theta3[i], col = 1, lty = 1)
        abline(v = Bs[[1]]$theta[i], lty = 2, col = 2, lwd = 2)
        abline(v = Bs[[2]]$theta[i], lty = 2, col = 3, lwd = 2)
        if(i==1)
            legend("topleft", bty = "n",
                   c("true", "prior",
                     as.expression(
                         lapply(nsims, function(ns)
                             bquote("Obs."~theta~", n="~.(ns)))),
                    as.expression(
                         lapply(nsims, function(ns)
                             bquote("Post."~theta~", n="~.(ns))))),
                   lty = c(1, 3, 2, 2, 1, 1),
                   col = c(1, 1, 2:3, 2:3),
                   lwd = c(1, 3, 2, 2, 2, 2),
                   title = as.expression(bquote(lambda == .(lamb))))
    }
    hpriors3 <- lapply(itests, function(r)
                       inla.hyperpar.sample(10000, r))
    cpriors3 <- lapply(hpriors3, apply, 1, function(x)
        tcrossprod(cholcor(x, p))[il3])
    hh <- 0.01
    for(i in 1:3) {
        h1 <- hist(cpriors3[[1]][i, ], seq(-1, 1, hh), plot = FALSE)
        h2 <- hist(cpriors3[[2]][i, ], seq(-1, 1, hh), plot = FALSE)
        ih <- which((h1$counts>0) | (h2$counts>0))
        plot(h1, freq = FALSE, xlab = rholabs[[i]],
             ylab = prholabs[[i]],
             xlim = range(h1$mids[ih], h2$mids[ih]) + c(-1,1)*hh,
             ylim = range(h1$dens[ih], h2$dens[ih]),
             main = "", col = rgb(1,0,0,0.5))
        plot(h2, add = TRUE, freq = FALSE, col = rgb(0,1,0,0.5))
        abline(v = c0[il3[i]], lty = 1, col = 1)
        rug(c0[il3[i]], col = 1, lwd = 1)
        rug(Rs[[1]][il3[i]], col = 2, lwd = 5, lty = 2)
        rug(Rs[[2]][il3[i]], col = 3, lwd = 5, lty = 2)
        if(i==1)
            legend("topleft", bty = "n",
                   paste("n = ", nsims),
                   fill = c(2,3), border = 2:3,
                   title = as.expression(bquote(lambda == .(lamb))))
    }
}
```

We see that for $\lambda=0.1$ and $\lambda=1$ 
the posterior distributions are around the "true" values.
Also, for the bigger sample size, the posterior is
narrower. 
However, with $n=50$, 
for the $\lambda=10$ and $\lambda = 1000$ cases, 
each posterior is around 0 (zero), which is the value 
at the base model.
Therefore, with a sample size of $n=50$ it was not enough 
to move the posterior away from these priors, 
and is narrower than the posterior with $n=500$. 
With $\lambda = 1000$, 
the biggest value considered for $\lambda$ here, 
both sample sizes were not able to move away from the prior.

\textbf{NOTE}: this was just for one simulated data. 
The results may differ for different data simulated in 
the same way but different seed.
But, **most important** is that it depend on how far 
$\theta_j$ is from the base model.
Let us compute the mean of squared differences
```{r d3}
c(sd=sd(theta3))
```

# Example with $p=4$

We now consider $p=4$ that give $m=6$. 
We simulate $\btheta$ that not too close to the 
base model (vector of zeros).
Therefore, the following results have to be taken 
considering this. 
The posterior with $n=50$ is away from the prior, 
and around the true values, for $\lambda=0.1$ and $\lambda=1$. 
With $n=500$, only for the case with $\lambda=1000$ 
there is not enouth information in the data to move 
the posterior away from the prior. 

```{r p4}
p <- 4
(m <- p*(p-1)/2)

repeat{
 theta0 <- rnorm(m, 0, 0.5)
 if(sd(theta0)>1) break
}
print(theta0)
print(c(sd=sd(theta0)))

lR <- cholcor(theta0, p = p)
R <- tcrossprod(lR)
iil <- which(lower.tri(R))
R
```

We also simulate two data sets for $p=4$ with
```{r samples4}
xxs <- lapply(nsims, function(ns)
    tcrossprod(matrix(rnorm(ns * p), ns), lR))
```
compute the sample correlation with
```{r Rs4}
Rs <- lapply(xxs, cor)
lapply(Rs, round, 2)
```
and the base model at the sample correlation (to get observed theta)
```{r obsth4}
Bs <- lapply(Rs, basecor, p = p)
round(t(oths <- sapply(Bs, function(b) b$theta)), 2)
```

Organize the data
```{r data4}
datfs <- lapply(c(s=1:2), function(i)
    data.frame(
        y = as.vector(xxs[[i]]),
        i = rep(1:p, each = nsims[i]),
        r = rep(1:nsims[i], p)))
names(datfs) <- names(nsims)
```

Prepare cosmetic labels
```{r labs4}
rholabs4 <- c(
    expression(rho[2~","~1]),
    expression(rho[3~""~1]),
    expression(rho[4~""~1]),
    expression(rho[3~""~2]),
    expression(rho[4~""~2]),
    expression(rho[4~""~3])
)
prholabs4 <- c(
    expression(pi[rho[2~","~1]](rho[2~","~1]~"|"~lambda)),
    expression(pi[rho[3~","~1]](rho[3~","~1]~"|"~lambda)),
    expression(pi[rho[4~","~1]](rho[4~","~1]~"|"~lambda)),
    expression(pi[rho[3~","~2]](rho[3~","~2]~"|"~lambda)),
    expression(pi[rho[4~","~2]](rho[4~","~2]~"|"~lambda)),
    expression(pi[rho[4~","~3]](rho[4~","~3]~"|"~lambda))
)
```

We will now visualize the posterior for each correlation
(each column of plots)
considering these two data (different colors),
for different $\lambda$ (each row in the figure).


```{r c4vis, fig.width = 12, fig.height = 10}
par(mfrow = c(nlambs, m), mar = c(3,3,0,0), mgp = c(1.5,0.5,0))
for(il in 1:nlambs) {
    lamb <- lambdas[il]
    cmodel <- cgeneric("pc_correl", n = p, lambda = lamb,
                       useINLAprecomp = FALSE)
    itests <- lapply(datfs, function(ddf)
        inla(formula = fr,
             data = ddf,
             control.family = cfam
             )
        )
    hpriors <- lapply(itests, function(r)
        inla.hyperpar.sample(10000, r))
    cpriors <- lapply(hpriors, apply, 1, function(x)
        tcrossprod(cholcor(x, p=p))[iil])
    hh <- 0.01
    for(i in 1:m) {
        h1 <- hist(cpriors[[1]][i, ], seq(-1, 1, hh), plot = FALSE)
        h2 <- hist(cpriors[[2]][i, ], seq(-1, 1, hh), plot = FALSE)
        ih <- which((h1$counts>0) | (h2$counts>0))
        plot(h1, freq = FALSE, xlab = rholabs4[[i]],
             ylab = prholabs4[[i]],
             xlim = range(h1$mids[ih], h2$mids[ih]) + c(-1,1)*hh,
             ylim = range(h1$dens[ih], h2$dens[ih]),
             main = "", col = rgb(1,0,0,0.5))
        plot(h2, add = TRUE, freq = FALSE, col = rgb(0,1,0,0.5))
        abline(v = c0[iil[i]], lty = 1, col = 1)
        rug(c0[iil[i]], col = 1, lwd = 1)
        rug(Rs[[1]][iil[i]], col = 2, lwd = 5, lty = 2)
        rug(Rs[[2]][iil[i]], col = 3, lwd = 5, lty = 2)
        if(i==1)
            legend("topleft", bty = "n",
                   paste("n = ", nsims),
                   fill = c(2,3), border = 2:3,
                   title = as.expression(bquote(lambda == .(lamb))))
    }
}
```

# Example with $p=5$

With $p=5$ we have now $m=10$. 
We have similar results, attributable to the fact
that the $\btheta$ simulated reasonable far from 
the base model.


```{r p5, echo = FALSE, fig.width = 15, fig.height = 10}
p <- 5
(m <- p*(p-1)/2)

repeat{
 theta0 <- rnorm(m, 0, 0.5)
 if(sd(theta0)>1) break
}
print(theta0)
print(c(sd=sd(theta0)))

lR <- cholcor(theta0, p = p)
round(R <- tcrossprod(lR), 2)
iil <- which(lower.tri(R))
xxs <- lapply(nsims, function(ns)
    tcrossprod(matrix(rnorm(ns * p), ns), lR))
Rs <- lapply(xxs, cor)
lapply(Rs, round, 2)
Bs <- lapply(Rs, basecor, p = p)
round(t(oths <- sapply(Bs, function(b) b$theta)), 2)
datfs <- lapply(c(s=1:2), function(i)
    data.frame(
        y = as.vector(xxs[[i]]),
        i = rep(1:p, each = nsims[i]),
        r = rep(1:nsims[i], p)))
names(datfs) <- names(nsims)

## Prepare cosmetic labels
rholabsp <- c(
    expression(rho[2~","~1]),
    expression(rho[3~""~1]),
    expression(rho[4~""~1]),
    expression(rho[5~""~1]),
    expression(rho[3~""~2]),
    expression(rho[4~""~2]),
    expression(rho[5~""~2]),
    expression(rho[4~""~3]),
    expression(rho[5~""~3]),
    expression(rho[5~""~4])
)
prholabsp <- c(
    expression(pi[rho[2~","~1]](rho[2~","~1]~"|"~lambda)),
    expression(pi[rho[3~","~1]](rho[3~","~1]~"|"~lambda)),
    expression(pi[rho[4~","~1]](rho[4~","~1]~"|"~lambda)),
    expression(pi[rho[5~","~1]](rho[4~","~1]~"|"~lambda)),
    expression(pi[rho[3~","~2]](rho[3~","~2]~"|"~lambda)),
    expression(pi[rho[4~","~2]](rho[3~","~2]~"|"~lambda)),
    expression(pi[rho[5~","~2]](rho[4~","~2]~"|"~lambda)),
    expression(pi[rho[4~","~3]](rho[4~","~2]~"|"~lambda)),
    expression(pi[rho[5~","~3]](rho[4~","~2]~"|"~lambda)),
    expression(pi[rho[5~","~4]](rho[4~","~3]~"|"~lambda))
)

par(mfrow = c(nlambs, m), mar = c(3,3,0,0), mgp = c(1.5,0.5,0))
for(il in 1:nlambs) {
    lamb <- lambdas[il]
    cmodel <- cgeneric("pc_correl", n = p, lambda = lamb,
                       useINLAprecomp = FALSE)
    itests <- lapply(datfs, function(ddf)
        inla(formula = fr,
             data = ddf,
             control.family = cfam
             )
        )
    hpriors <- lapply(itests, function(r)
        inla.hyperpar.sample(10000, r))
    cpriors <- lapply(hpriors, apply, 1, function(x)
        tcrossprod(cholcor(x, p=p))[iil])
    hh <- 0.01
    for(i in 1:m) {
        h1 <- hist(cpriors[[1]][i, ], seq(-1, 1, hh), plot = FALSE)
        h2 <- hist(cpriors[[2]][i, ], seq(-1, 1, hh), plot = FALSE)
        ih <- which((h1$counts>0) | (h2$counts>0))
        plot(h1, freq = FALSE, xlab = rholabsp[[i]],
             ylab = prholabsp[[i]],
             xlim = range(h1$mids[ih], h2$mids[ih]) + c(-1,1)*hh,
             ylim = range(h1$dens[ih], h2$dens[ih]),
             main = "", col = rgb(1,0,0,0.5))
        plot(h2, add = TRUE, freq = FALSE, col = rgb(0,1,0,0.5))
        abline(v = c0[iil[i]], lty = 1, col = 1)
        rug(c0[iil[i]], col = 1, lwd = 1)
        rug(Rs[[1]][iil[i]], col = 2, lwd = 5, lty = 2)
        rug(Rs[[2]][iil[i]], col = 3, lwd = 5, lty = 2)
        if(i==1)
            legend("topleft", bty = "n",
                   paste("n = ", nsims),
                   fill = c(2,3), border = 2:3,
                   title = as.expression(bquote(lambda == .(lamb))))
    }
}
```

# References
