---
title: "Multivariate Penalized Complexity Prior"
output:
  pdf_document: 
    keep_tex: yes
  html_document:
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Multivariate Penalized Complexity Prior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amsfonts}
  - \usepackage{amssymb}
  - \usepackage{graphicx}
  - \usepackage{mathtools}
  - \usepackage{calc}
  - \newcommand{\bx}{\textbf{x}}
  - \newcommand{\bmu}{\boldsymbol{\mu}}
  - \newcommand{\bgamma}{\boldsymbol{\gamma}}
  - \newcommand{\btheta}{\boldsymbol{\theta}}
  - \newcommand{\bxi}{\boldsymbol{\xi}}
  - \newcommand{\brho}{\boldsymbol{\rho}}
  - \newcommand{\bvphi}{\boldsymbol{\varphi}}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  include = TRUE, 
  message = FALSE,
  warning = FALSE,
  fig.align='center',
  fig.width = 7, 
  fig.height = 4, 
  out.width = '0.59\\textwidth')
library(graphpcor)
library(plot3D)

```

# Abstract

The derivation of the 
multivariate penalized complexity prior 
is detailed and some illustrative examples given.

# Introduction

A key ingredient in statistical models is the distribution 
assumed for a random variable in the model.
A probability density function, (in short) distribution, is 
a non-negative mathematical function whose domain is 
the random variable and integrates to one.
For a random vector $\bx$, where each 
element is a random variable,  
a chosen distribution to model $\bx$,
$\pi_{\bx}(\bxi|\bgamma)$, is indexed by 
a $m$ dimentional vector $\bgamma$, 
the model parameters.
When learning about the model parameters through Bayesian 
inference, prior information about it is considered.
The prior information is represented by a distribution itself, 
which has parameters as well.

The aim to construct a prior distribution for $\bgamma$, 
$\pi_{\bgamma}(\bgamma|\lambda)$, 
where $\lambda$ is the prior parameter,
assumed here to be a scalar.
@simpsonetal2017 proposed a new concept. 
Instead of searching for a distribution for $\bgamma$, 
that better represent the prior beliefs, 
it was proposed to start with a base model, 
which is understood as a simpler model when 
$\bgamma=\bgamma_0$ in $\pi_{\bx}(\bx|\bgamma_0)$. 
Then, define the prior from a distribution to
penalize deviations from the base model, 
the penalized complexity prior, or PC-prior.

The domain of $\bgamma$ may be bounded. 
This is the case of variance parameters that are positive and
correlations which are between $-1$ and $1$.
In such case, a practical approach is to consider a mapping 
$\bgamma = \mathbf{h}(\btheta)$ such that
$\mathbf{h}^{-1}(\bgamma)$ = 
$\btheta \in \mathbf{R}^{m}$. 
By considering this mapping to be defined by
$m$ univariate transformations
$\btheta=
\{\theta_1=h_1(\gamma_1), \ldots, \theta_m = h_m(\gamma_m)\}$, 
where $h_j()$ is the identity function if $\gamma_j$ is unbounded,
we have 
\[
\pi_{\bgamma}(\bgamma|\lambda,\bgamma_0) =
\pi_{\btheta}(\mathbf{h}^{-1}(\bgamma)|\lambda,\mathbf{h}^{-1}(\bgamma_0)) 
\prod_{k=1}^{m}
|\frac{\partial h_k^{-1}(\gamma_k)}{\partial \gamma_k}|
\]
and in a more general form of $\mathbf{h}(\btheta)$ we have
\[
\pi_{\bgamma}(\bgamma|\lambda,\bgamma_0) =
\pi_{\btheta}(\mathbf{h}^{-1}(\bgamma)|\lambda,\mathbf{h}^{-1}(\bgamma_0)) 
|\textrm{det}(\textbf{G})|.
\]
where $\mathbf{G}_{ij}$ is 
$\frac{\partial \textbf{h}^{-1}_i(\bgamma)}{\partial \bgamma_j}$.

The proposal is to penalize a Kullback-Leibler divergence (KLD).
The KLD is not a metric but a divergence, a statistical distance, 
used to measure how much a model is different from the true model.
The approach is to replace the true model by the base model.
If $\mathbf{x}$ follows a multivariate Gaussian distribution, 
$\mathbf{x}\sim N(\bmu, \Sigma)$, the KLD of model
$\pi_\mathbf{x}(\mathbf{x}|\btheta)$
to the base model, $\pi_{\mathbf{x}}(\mathbf{x}|\btheta_0)$, is
\[\textrm{KLD}(\btheta,\btheta_0) = 
\frac{1}{2}\left\{ \textrm{trace}(\Sigma_0^{-1}\Sigma) +
(\bmu-\bmu_0)^T\Sigma_0(\bmu-\bmu_0)-m
-\log{|\Sigma|} + \log{|\Sigma_0|} \right\} \]
where $\bmu$ and $\Sigma$ are functions of $\btheta$.

For $\theta_0$ in the interior of the parameter space,
the KLD behaves quadratically, which leads to 
\[\textrm{KLD}(\btheta,\btheta_0) \approx 
\frac{1}{2}(\btheta-\btheta_0)^T\mathbf{I}(\btheta_0)
(\btheta-\btheta_0) \]
where $\mathbf{I}(\btheta_0)$ is the Hessian of
KLD($\btheta,\btheta_0$) around $\btheta_0$.
Actually, $\mathbf{I}(\btheta_0)$ is the
Fisher information matrix, around $\btheta_0$, 
because the two models are in the same family and 
have a common parametrization. 
However, the goodness of such approximation depends on the chosen
model's parametrization for $\pi_{\bx}(\mathbf{x}|\btheta)$.

We take the second order approximation for KLD($\btheta,\btheta_0$) 
to define a location scale transformation 
$\bxi = \mathbf{I}(\btheta_0)^{1/2}(\btheta-\btheta_0)$, 
along with $\btheta = \mathbf{I}(\btheta_0)^{-1/2}\bxi + \btheta_0$.
This scaling makes $\bxi$ unitless, scale-invariant, and takes into 
account the correlations between $\btheta$, 
in the information geometric sense 
because the natural gradient through 
$\mathbf{I}(\btheta_0)$ was used for the scaling.
The prior for $\btheta$ is then computed as
\[\pi_{\btheta}(\btheta|\lambda) = 
\pi_{\bxi}(\bxi|\lambda)
|\textrm{det}(\mathbf{I}(\btheta_0)^{1/2})| \]
where $\pi_{\bxi}(\bxi|\lambda)$ is the PC-prior for $\bxi$, 
and, as the determinant of $\mathbf{I}(\btheta_0)^{1/2}$
does not depend on $\btheta$, it need to be computed only once.
If $\theta_0$ does not represent a simpler model, 
the approach can still be considered. 
Such cases may lead the assignment of an informative prior, 
and such approximation makes $\theta_0$ its mode.

# The univariate PC-prior

From the aforementioned assumptions and definitions we consider
$\sqrt{2\textrm{KLD}(\btheta,\btheta_0)} \approx \sqrt{\bxi^T\bxi} = d(\bxi)$. 
We have that $d(\btheta_0)=0$ and 
$d(\btheta)>0$ if $\btheta \neq \btheta_0$.
The proposal in @simpsonetal2017 assume 
$d(\bxi)\sim\textrm{Exponential}(\lambda)$.
Considering
$\xi \in \mathbb{R}$ and $r$ = $d(\xi)$ = $\sqrt{\xi^2}$ = $|\xi|>0$ gives $\pi_{\xi}(\xi|\lambda)$ = 
$\frac{1}{2}\pi_{d(\xi)}(d(\xi)|\lambda)$ =
$\frac{\lambda}{2}\textrm{e}^{-\lambda|\xi|}$
which is the Laplace distribution 
with zero mean and precision $\lambda$.
Applying $m=1$ in the formula for the multivariate PC-prior 
in the next section gives the same result.
In Figure 1 we show this density 
for some values of $\lambda$.

\begin{figure}
\label{fdm1}
```{r m1}
lseq <- c(0.3, 1, 5)
llab <- lapply(lseq, function(l) 
  bquote(lambda == .(l)))
nl <- length(lseq)
lcol <- c(1,1,1)#2, 4)
b <- 3; a <- -b
h  <- 0.05
x0 <- seq(a+h/2, b-h/2, h)
nx0 <- length(x0)

par(mfrow = c(1,1), mar = c(3,3,0,0), mgp = c(2,0.5,0), 
    las = 1, bty = "n")
plot(function(x) 0.5*lseq[nl]*exp(-abs(x)*lseq[nl]), 
     a, b, n = 2001, type = 'n',
     xlab = expression(xi), 
     ylab = expression(pi[xi](xi~"|"~lambda)),
     main = '')
for(i in 1:nl) {
  plot(function(x) 0.5*lseq[i]*exp(-abs(x)*lseq[i]), 
       a, b, n = 2001, add = TRUE, 
       col = lcol[i], lty = i, lwd = 2)
}
legend("topleft", bty = "n", 
       as.expression(llab), 
       col = lcol, lty = 1:nl, lwd = 2) 
```
\caption{The Laplace density function for some values of $\lambda$.}
\end{figure}

## Example 

Suppose we want to model a bivariate random vector 
$\mathbf{x}=\{x_1, x_2\}$ and assume that it follows a
bivariate Gaussian distribution.
We assume that the mean of both variables is zero and 
unit marginal variances. 
In this case we have the correlation parameter to infer, $\rho$.
We define
\[ \mathbf{C}(\rho) = \left[ \begin{array}{cc} 
1 & \rho \\
\rho & 1 
\end{array} \right] \]
so we have
$\mathbf{x}|\rho \sim \textrm{N}(0, \mathbf{C}(\rho))$. 

The first step is to consider the map from the 
(bounded) model parameter, $\rho\in(-1,1)$ to $\theta$. 
We choose to define 
$g(\rho)=\textrm{tanh}(\rho)=\theta$ and 
$g^{-1}(\theta) = \textrm{tanh}^{-1}(\rho)=\textrm{atanh}(\rho)$,
which gives $\partial g^{-1}(\rho)=1/(1-\rho^2)$
leading to 
\[
\pi_\rho(\rho|\lambda) = \pi_\theta(\theta|\lambda)/(1-\rho^2).
\]

The next step is to define a base model as a model 
when the parameter assume a particular value in 
the interior of the parametric domain.
We assume a $\textrm{N}(0, \mathbf{C}(\theta_0))$. 
This gives the Kullback-Leibler divergence as
\[
\textrm{KLD}(\theta,\theta_0) = \frac{1}{2} \left(
\frac{1-\textrm{tanh}(\theta)\textrm{tanh}(\theta_0)}
{1-\textrm{tanh}(\theta_0)^2} - 2
-\textrm{log}(1-\textrm{tanh}(\theta)^2) +
\textrm{log}(1-\textrm{tanh}(\theta_0)^2) \right) .\]
and
\[ \begin{array}{rcl}
\mathbf{I}(\theta_0) & = &  
\frac{\partial^2}{\partial\theta^2}\frac{1}{2}
\left(
  \frac{1-\textrm{tanh}(\theta_0)\tanh{\theta}}{1-\textrm{tanh}(\theta_0)^2}-2
-\textrm{log}(1-\textrm{tanh}^2(\theta))+\textrm{log}(1-\textrm{tanh}(\theta_0)^2) \right)\\
 & = & \frac{\partial}{\partial\theta} \left( \frac{(1-\textrm{tanh}^2(\theta))\textrm{tanh}(\theta_0)}{1-\textrm{tanh}(\theta_0)^2}+\textrm{tanh}(\theta) \right) \\
 & = & \frac{2\textrm{tanh}(\theta)(1-\textrm{tanh}^2(\theta))\textrm{tanh}(\theta_0)}{1-\textrm{tanh}(\theta_0)^2} + 
 1-\textrm{tanh}^2(\theta)
 \overset{\theta=\theta_0}{=}
1 - \textrm{tanh}^2(\theta_0) 
 \end{array} \]
This gives
\[
\pi_\theta(\theta|\lambda) = \pi_\xi(\xi|\lambda)\sqrt{1-\textrm{tanh}^2(\theta_0)}. 
\]

Following the steps we have 
$\bxi = \mathbf{I}(\theta_0)^{1/2}(\theta - \theta_0)$ = 
$\sqrt{(1-\textrm{tanh}^2(\theta_0))}(\theta-\theta_0)$, 
which gives 
\[ \pi_\rho(\rho|\lambda) = 
\frac{\lambda\sqrt{(1-\rho_0^2)}}{2(1-\rho^2)}\textrm{e}^{-\lambda
|\sqrt{(1-\rho_0^2)}(\textrm{atanh}(\rho)-\textrm{atanh}(\rho_0))|}.
\]
We can see this density computed at different base models, 
for some values of $\lambda$ in Figure 3.


\begin{figure}
\label{fdrho}
```{r rho0, out.width='0.99\\textwidth', fig.width=9, fig.height=3}
drho <- function(rho, rho0, l) {
  x0 <- atanh(rho0)
  I0 <- 1-rho0^2
  x <- atanh(rho)
  sI0 <- sqrt(I0)
  xi <- sI0 * (x - x0)
  ld <- log(l*sI0/2) -l*abs(xi) -log(1-rho^2)
  exp(ld) 
}
rho0s <- c(-0.5, 0, 0.7)

if(FALSE)
  sapply(rho0s, function(r)
    sapply(lseq, function(l)
        integrate(function(x) drho(x,r,l),-1,1)$val))

par(mfrow = c(1, 3), mar = c(4,4,1,1), mgp = c(3,1,0), las = 1)
for(k in 1:3) {
  plot(function(x) drho(x,rho0s[k],lseq[nl]), 
       -1, 1, n=1001, type = "n", bty = "n", 
       xlab = expression(rho), 
       ylab = expression(pi[rho](rho~"|"~lambda~","~rho[0])))
  for(i in 1:nl) {
    plot(function(x) drho(x, rho0s[k], lseq[i]), 
         -1, 1, n=1001, add = TRUE, col = lcol[i], lty = i, lwd = 2)
  }
}
legend("top", bty = "n", 
       as.expression(llab), 
       ##title = expression(lambda), 
       col = lcol, lty = 1:nl, lwd = 2) 
```
\caption{Penalized complexity prior for $\rho$, 
for different base models: $\rho_0=-0.5$ (left), 
$\rho_0=0$ (center), and $\rho_0=0.9$ (rigth),
for some $\lambda$ values.}
\end{figure}


# The multivariate penalized complexity prior

The multivariate penalized complexity prior (MPCP) for $\bxi$ 
is an extension of the univariate penalized complexity prior. 
For $m>1$, let a manifold $\mathcal{M}$, 
where $\bxi\in\mathcal{M}$, 
be defined so that for each $d(\bxi) = r \geq 0$, the level sets
$\bxi \in S_r=\{\bxi\in\mathcal{M}:d(\bxi)=r\}$,
are embedded sub-manifolds decomposing
$\mathcal{M}$ = $\mathbb{R}^+(\cup_{r\geq 0}S_r)$.
The MPCP definition is completed by assuming that
for each $d(\bxi) = r>0$, the coordinates $\bxi$ 
are uniformly distributed in $S_r$.

A convenient option is to consider 
$\mathbf{g}(\bxi) = (r,\bvphi)$ to be
a map from Cartesian to spherical coordinates,
where $r$ is the radius and $\bvphi$ = 
$\varphi_1,\ldots,\varphi_{m-1}$ the $m-1$ angles.
The interpretation is that, for $r>0$, $\bxi$ lay on an $m-1$
dimensional sphere surface with radius $r$. 
With $m=2$, for each $r$, $S_r$ is a circle and 
$\{r,S_r\}$ make up a cone like those shown in Figure 3.

Under the uniform distribution assumption for the coordinates, 
which lay on the ($m-1$)-surface area with radius $r=d(\bxi)$,
in the Euclidean space, we have 
\[
\pi_{\bxi}(\bxi|\lambda) = 
\frac{\lambda \textrm{e}^{-\lambda r}}{|S_r|} =
\frac{\Gamma(m/2)}{2\pi^{m/2}r^{m-1}}
\lambda \textrm{e}^{-\lambda r} 
\]
and, in the spherical space which is 
useful for analytical computations, we have
\[
\pi_{r,\bvphi}(r,\bvphi|\lambda) = 
\frac{\Gamma(m/2)}{2\pi^{m/2}r^{m-1}}
\lambda \textrm{e}^{-\lambda r} 
|\textrm{det}(\mathbf{J})|
\]
where $\mathbf{J}$
is the Jacobian of $\mathbf{g}$, 
$\mathbf{J}_{ij} = \frac{\partial g_i}{\partial \varepsilon_j}$,
that gives
$\textrm{det}(\mathbf{J})$ =
$r^{m-1}$ 
$\sin^{m-2}(\varphi_1)$ 
$\sin^{m-3}(\varphi_2)$ ... 
$\sin(\varphi_{m-2})$.

For $m=2$ we have $|S_{d(\bxi)}|$ = $2\pi r$, 
giving
\[
\pi_{\bxi}(\xi_1,\xi_2|\lambda) =
\frac{\lambda}{2\pi r}\textrm{e}^{-\lambda r}
\]
and we can use the polar coordinates case as
\[\begin{array}{ccl|ccl}
r & = & \sqrt{\xi_1^2+\xi_2^2} &
\xi_1 & = & r\textrm{cos}(\varphi_1) \\
\varphi_1 & = & 
\textrm{atan2}(\xi_2,\xi_1) & 
\xi_2 & = & r\textrm{sin}(\varphi_1) 
\end{array}\]
which gives $|\mathbf{J}|$ = $r$, so that
\[ \begin{array}{rcl}
\int_{\mathbb{R}} \int_\mathbb{R} 
\frac{\lambda}{2\pi r}\textrm{e}^{-\lambda r}
\partial \xi_1 \partial \xi_2  & = &
\int_{0}^{2\pi}\int_{0}^{\infty}
\frac{\lambda}{2\pi r}\textrm{e}^{-\lambda \sqrt{r^2\textrm{cos}^2(\theta)+r^2\textrm{sin}^2(\theta)}} 
r \partial r \partial \theta \\
 & = & \int_{0}^{2\pi}\int_{0}^{\infty}
\frac{\lambda}{2\pi}
\textrm{e}^{-\lambda r} 
\partial r \partial \theta 
\;\; = \;\; \int_{0}^{2\pi} 
\frac{-1}{2\pi}
\textrm{e}^{-\lambda r} {|}_{r=0}^{r=\infty} \partial \theta  \\
 & = & \int_{0}^{2\pi} 
\frac{1}{2\pi} \partial \theta 
\;\; = \;\; \frac{\theta}{2\pi} {|}_{\theta=0}^{\theta=2\pi} = 1
\end{array} \]
We can visualize it in Figure 3, 
for some values of $\lambda$.

\begin{figure}
\label{fdm2}
```{r m2, out.width='0.99\\textwidth', fig.width=12, fig.height=5}
dfn <- function(lambda) {
  outer(x0, x0, function(a,b) {
        r <- sqrt(a^2+b^2)
        lambda*exp(-lambda*r)/(2*pi*r)})
}
d2 <- lapply(lseq, function(l) dfn(l))
par(mfcol = c(1, 3), mar = c(0,0,3,0))
for(i in 1:nl) {
  bkz <- exp(seq(log(min(d2[[i]]))-1e-9, 
                 log(max(d2[[i]]))+1e-9, length = 65))
  surf3D(matrix(x0, nx0, nx0), 
         matrix(rep(x0, each = nx0), nx0), 
         matrix(d2[[i]], nx0), 
         breaks = bkz, colkey = !TRUE,
         xlab = "", 
         ylab  = "", 
         zlab = 'Density',
         box = TRUE, bty = "b", phi = 20, theta = 30)
  text(-0.15, -0.43, expression(xi[1]), adj = 1)
  text(0.28, -0.37, expression(xi[2]), adj = 1)
  legend("topleft", bty = "n", as.expression(llab[[i]]), cex = 2)
}

```
\caption{Bivariate Penalized Complexity prior, around $\bxi_0$,
for some values of $\lambda$.}
\end{figure}

For $m=3$ we have $|S_{d(\bxi)}|$ = $4\pi r^2$, giving
\[
\pi_{\bxi}(\xi_1,\xi_2,\xi_3|\lambda) =
\frac{\lambda\textrm{sin}(\varphi_1)}{4\pi}
\textrm{e}^{-\lambda r} .
\]
The mapping is defined as
\[\begin{array}{ccl|ccl}
r & = & \sqrt{\xi_1^2+\xi_2^2+\xi_3^2} &
\xi_1 & = & r\textrm{cos}(\varphi_1) \\
\varphi_1 & = & 
\textrm{atan2}(\sqrt{\xi_2+\xi_3^2},\xi_1) & 
\xi_2 & = & r\textrm{sin}(\varphi_1)\textrm{cos}(\varphi_2) \\
\varphi_2 & = & 
\textrm{atan2}(\xi_3,\xi_2) & 
\xi_3 & = & r\textrm{sin}(\varphi_1)\textrm{sin}(\varphi_2) .
\end{array}\]
Considering
\[ \begin{array}{rcl}
r^2 & = & \xi_1^2+\xi_2^2+\xi_3^2 \\
 & = & r^2\textrm{cos}^2(\varphi_1) + 
r^2\textrm{sin}^2(\varphi_1)\textrm{cos}^2(\varphi_2) +
r^2\textrm{sin}^2(\varphi_1)\textrm{sin}^2(\varphi_2) \\
& = & r^2 \{ \textrm{cos}^2(\varphi_1) +
r^2\textrm{sin}^2(\varphi_1)[\textrm{cos}^2(\varphi_2)+\textrm{sin}^2(\varphi_2)]\} \\
& = & r^2[\textrm{cos}^2(\varphi_1) +
r^2\textrm{sin}^2(\varphi_1)] = r^2
\end{array}\]
and $|\mathbf{J}|$ = $r^2\textrm{sin}(\varphi_1)$,
we verify that
\[ \begin{array}{rcl}
 \int_\mathbb{R} \int_\mathbb{R} \int_\mathbb{R} 
\frac{\lambda \textrm{e}^{-\lambda r}}{4\pi r^2}
\partial \xi_1 \partial \xi_2 \partial \xi_3 & = & 
\int_0^{2\pi} \int_0^\pi \int_0^\infty 
\frac{\textrm{sin}(\varphi_1)}{4\pi}
\lambda \textrm{e}^{-\lambda r} 
\partial r \partial \varphi_1 \partial \varphi_2 \\
& = &
\int_{0}^{2\pi} \int_{0}^{\pi}
\frac{\textrm{sin}(\varphi_1)}{4\pi}
(-\textrm{e}^{-\lambda r})|_{r=0}^\infty 
\partial \varphi_1 \partial \varphi_2 =
 \int_{0}^{2\pi} \int_{0}^{\pi}
\frac{\textrm{sin}(\varphi_1)}{4\pi} \partial \varphi_1 \partial \varphi_2  \\
 & = & \int_{0}^{2\pi} \frac{(-\textrm{cos}(x))}{4\pi}
 |_{x=0}^{x=\pi} \partial \varphi_2 
\;\; = \;\; \int_0^{2\pi} \frac{1}{2\pi} \partial \varphi_2
\;\; = \;\; \frac{x}{2\pi}{|}_{x=0}^{x=2\pi} = 1
\end{array} \]

For $m>3$ we have the mapping defined as
\[\begin{array}{ccl|ccl}
r & = & \sqrt{\xi_1^2+\xi_2^2+\ldots+\xi_m^2} &
\xi_1 & = & r\textrm{cos}(\varphi_1) \\
\varphi_1 & = & 
\textrm{atan2}(\sqrt{\xi_2^2+\ldots+\xi_m^2},\xi_1) & 
\xi_2 & = & r\textrm{sin}(\varphi_1)\textrm{cos}(\varphi_2) \\
& \vdots & & & \vdots & \\
\varphi_{m-2} & = & 
\textrm{atan2}(\sqrt{\xi_{m-1}^2+\xi_{m}^2},\xi_{m-2}) & 
\xi_{m-1} & = &  r\textrm{sin}(\varphi_1)\ldots\textrm{sin}(\varphi_{m-2})\textrm{cos}(\varphi_{m-1}) \\
\varphi_{m-1} & = & \textrm{atan2}(\xi_{m},\xi_{m-1}) &
\xi_{m} & = & r\textrm{sin}(\varphi_1)\ldots\textrm{sin}(\varphi_{m-2})\textrm{sin}(\varphi_{m-1})
\end{array} .\]


## Example with $m=2$

We consider a three dimentional random vector 
$\mathbf{x}=\{x_1, x_2, x_3\}$ and assume that it follows a
bivariate Gaussian distribution with zero mean vector
and unit marginal variances. 
Therefore, remains three correlations to be modeled:
$\rho_{1,2}$ = Cor($x_1,x_2$),
$\rho_{1,3}$ = Cor($x_1,x_3$) and
$\rho_{2,3}$ = Cor($x_2,x_3$). 
We assume that conditional on $x_1$, 
the other two variables are independent, Cor($x_2,x_3|x_1$)=0. 
This assumption can be taken by modeling the
correlation matrix from the decomposition 
$\mathbf{C}(\brho) = \mathbf{L}(\brho)\mathbf{L}(\brho)^T$, where
\[ \mathbf{L}(\brho) = \left[ \begin{array}{ccc} 
1 & 0 & 0 \\
\rho_{1,2} & \sqrt{1-\rho_{1,2}^2} & 0 \\
\rho_{1,3} & 0 & \sqrt{1-\rho_{1,3}^2}
\end{array} \right] \textrm{ which gives } 
\mathbf{C}(\brho) = \left[ \begin{array}{ccc} 
1 & \rho_{1,2} & \rho_{1,3} \\
\rho_{1,2} & 1 & \rho_{1,2}\rho_{1,3} \\
\rho_{1,3} & \rho_{1,2}\rho_{1,3} & 1
\end{array} \right]  \]
giving a model with two parameters because 
Cor($x_2,x_3$) now became $\rho_{1,2}\rho_{1,3}$.

The next step is to consider the mapping from the model parameter. 
We consider $\rho_1=\rho_{1,2}$, $\rho_2=\rho_{1,3}$ in 
$\brho=\{\rho_1, \rho_2\}$ and 
$\btheta=\{\textrm{atan}(\rho_1),
\theta_2=\textrm{atan}(\rho_2)\}$.
This gives 
\[ \pi_{\brho}(\brho|\lambda) = 
\frac{\pi_{\btheta}(\btheta|\lambda)}
{\sqrt{(1-\rho_1^2)(1-\rho_2^2)}}.\]
In the next step, 
although possible to do it analytically,
the computation of $\mathbf{I}(\btheta_0)$, 
the Hessian for KLD($\btheta,\btheta_0$)
will be performed numerically.

The prior distribution for $\brho$, 
computed for $\brho_0$=$\{-0.3,0.7\}$,
can be seen in Figure 4,
for some values of $\lambda$.

\begin{figure}\label{frho2}
```{r m2ex, out.width='0.99\\textwidth', fig.width=9, fig.height=3}
c3m2r <- function(r) {
  matrix(c(1,r[1],r[2], r[1],1,r[1]*r[2], r[2],r[1]*r[2],1), 3)
}
c3m2 <- function(x) {
  c3m2r(tanh(x))
}
drho2 <- function(rho, rho0, l, H) {
  x <- atanh(rho)
  x0 <- atanh(rho0)
  xi <- H$h.5 %*% (x - x0)
  r <- sqrt(sum(xi^2))
  ld <- log(l/r) + sum(log(H$s)) -0.5*sum(log(1-rho^2)) -l*r -log(2*pi)
  exp(ld) 
}

rs0 <- seq(-0.99, 0.99, 0.03)
nr0 <- length(rs0)
rs0x <- as.matrix(expand.grid(a=rs0, b=rs0))
rho0 <- c(-0.3, 0.7)

C0 <- c3m2r(rho0)
H <- list(
    I0=hessian(function(x) 
        graphpcor:::KLD10(c3m2(x), C0), 
        x = atanh(rho0)))
H$Hd <- svd(H$I0)
H$s <- sqrt(pmax(H$Hd$d, 0.0))
H$h.5 <- t(H$Hd$v %*% (t(H$Hd$u) * H$s))
H$hneg.5 <- t(H$Hd$v %*% (t(H$Hd$u) / H$s))

par(mfrow = c(1, 3), mar = c(0,1,0,2))
for(i in 1:nl) {
  d2 <- sapply(1:nrow(rs0x), function(j) 
    drho2(rs0x[j, ], rho0, lseq[i], H))
  bkz <- exp(seq(log(min(d2))-1e-9, 
                 log(max(d2))+1e-9, length = 65))
   surf3D(matrix(rs0, nr0, nr0), 
          matrix(rep(rs0, each = nr0), nr0), 
          matrix(d2, nr0), 
          breaks = bkz, colkey = !TRUE,
          xlab = '', ylab  = '', zlab = '',
          box = TRUE, bty = "b")
  text(-0.15, -0.43, expression(rho[1]), adj = 1)
  text(0.28, -0.37, expression(rho[2]), adj = 1)
  legend("topleft", bty = "n", as.expression(llab[[i]]), cex = 2)
}

```
\caption{Joint prior for $\brho$, 
  computed around $\brho_0$=$\{-0.3,0.7\}$, 
  for some values of $\lambda$.}
\end{figure}

One way to look at the marginal priors for each 
correlation pairs, including Cor($x_2$,$x_3$), is through sampling. 
We can sample from the MPCP as suggested in @simpsonetal2017.
Then we can evaluate $\mathbf{C}(\brho)$ at each sample
and visualize histograms like in Figure 5.

\begin{figure}\label{fcor3}
```{r rm, out.width='0.99\\textwidth', fig.width=9, fig.height=3}
rsphere <- function(n, m=2) {
    t(sapply(1:n, function(i) {
        z <- rnorm(m)
        z <- z / sqrt(sum(z^2))
    }))
}
rmpcp <- function(n, m, lambda = 1, Hneg.5, theta0) {
  out <- rsphere(n, m) * rexp(n, lambda)
  if(!missing(Hneg.5)) {
    out <- out %*% Hneg.5
  }
  if(!missing(theta0)) {
    out <- sweep(out, 2, -theta0)
  }
  return(out)
}

r2l <- lapply(1:nl, function(i) {
  rx <- rmpcp(1000, 2, lseq[i], H$hneg.5, atanh(rho0))
  r2 <- cbind(tanh(rx[, 1]), tanh(rx[, 2]))
})

h3l <- lapply(r2l, function(x)
  list(h1 = hist(x[, 1], -20:20/20, plot = FALSE), 
       h2 = hist(x[, 2], -20:20/20, plot = FALSE), 
       h3 = hist(x[, 1] * x[, 2], -20:20/20, plot = FALSE)))
yl <- c(0, max(unlist(lapply(h3l, lapply, function(x) max(x$density)))))

lfill <- c(gray(0.3, 0.5), rgb(0:1, 0, 1:0, 0.5))

par(mfrow = c(1, 3), mar = c(3,3,0,0), mgp = c(2,0.5,0))
xlb <- c("Cor(X1,X2)", "Cor(X1,X3)", "Cor(X2,X3)")
for(k in 1:3) {
  plot(h3l[[1]][[k]], freq = FALSE, 
       col = lfill[1], border = lfill[1], ylim = yl,
       xlab = xlb[k], main = '')
  for(i in 2:3)
    plot(h3l[[i]][[k]], freq = FALSE, 
         add = TRUE, col = lfill[i], border = lfill[i])
}
legend('topright', as.expression(llab), 
       fill = lfill, border = lfill)
```
\caption{Histograms for the three correlation pairs computed from 
samples of the MCPC prior, for some values of $\lambda$.}
\end{figure}

We now consider a more natural way to consider the assumption
Cor($X_2$, $X_3$ |, $X_1$)=0.
This assumption can be represented in the precision matrix, 
$\mathbf{Q}(\btheta)$.
which is a symmetric positive definite matrix and is zero 
for the entries $\mathbf{Q}(\btheta)_{2,3}=\mathbf{Q}(\btheta)_{3,2}=0$.
We start having the parameters in the lower trinagle but 
only where  the precision is non-zero. 
The Cholesky may have **fill-in** elements, which are 
function of the others, as the element {3,2} in our example.
The initial precision matrix is 
$\mathbf{Q}_0(\btheta) = \mathbf{L}_0(\btheta)\mathbf{L}_0(\btheta)^{T}$
where 
\[\mathbf{L}_0(\btheta) = \left[
\begin{array}{ccc}
1 & 0 & 0 \\
\theta_1 & 1 & 0 \\
\theta_2 &  \boldsymbol{-\theta_{1}\theta_{2}} & 1 
\end{array} \right] \textrm{ and so }
\mathbf{Q}_0(\btheta) = \left[
\begin{array}{ccc}
1 & \theta_1 & \theta_2 \\
\theta_1 & 1+\theta_1^2 & 0 \\
\theta_2 &  0 & 1+(1+\theta_1^2)\theta_2^2
\end{array} \right]
\]
and scale it in order to have 
a precision matrix whose inverse is a correlation matrix. 
This is done by defining 
$\mathbf{C}(\btheta) = 
\mathbf{S}_{0}^{-1}(\btheta) \Sigma_{0}(\btheta) \mathbf{S}_0^{-1}\btheta$, 
where $\Sigma_0(\btheta) =\mathbf{Q}_0(\btheta)$ and 
$\mathbf{S}_0(\btheta)$ is a diagonal matrix whose diagonal elements 
are the square root of the diagonal elements of $\Sigma_0(\btheta)$.
Notice that the model parameter space, for $\btheta$, is unbounded
and we do not need to to the first mapping.

The choice of a base model is not straightforward. 
The model parameters are elements of the Cholesky factor
of the initial precision matrix, thus not easy to interpret.
However, we can compute $\btheta_0$ from a given 
base correlation matrix. 
But, not all correlation matrices are allowed because of the
conditional independence assumption.
Thus, for a given correlation matrix we check if it fulfill 
the model assumption by checking if its inverse is zero at 
its entries $[2,3]$ and $[3,2]$.
If so, $\btheta_0$ can be computed by doing the reverse of the computations from a given base correlation matrix. 

We can do it with the following script
```{r gpc, echo = TRUE}
rho0 <- c(-0.3, 0.7)
C0 <- matrix(c(1, rho0, 
               rho0[1], 1, prod(rho0), 
               rho0[2], prod(rho0), 1), 3)

library(graphpcor)
gpc <- graphpcor(x1 ~ x2 + x3)
H <- hessian(gpc, x = C0)
(theta0 <- attr(H, "theta")) 
all.equal(C0,
          vcov(gpc, theta = theta0))
```

We evaluate $\pi_{\btheta}(\btheta|\lambda)$ 
with the `cgeneric` query function from the 
\textsf{\textbf{INLAtools}} package. 
For that, we consider a grid of values around $\btheta_0$. 
The marginals can be computed by numerical integration.

```{r cprior, out.width='0.99\\textwidth', fig.width=9, fig.height=3}
## set up a grid of theta around theta0
h <- 0.01
th0s <- seq(-1.5+h/2, 1.5-h/2, h)
nth0 <- length(th0s) 
th0x <- t(expand.grid(th0s + theta0[1], 
                      th0s + theta0[2])) ## each column: theta

dth <- lapply(lseq, function(lambda) {
  cm <- cgeneric(gpc, base = C0, 
                 lambda = lambda)
  matrix(exp(prior(cm, theta = th0x)), nth0)
})

par(mfrow = c(1, 2), mar = c(4,4,1,1), mgp = c(3,1,0))
plot(th0s + theta0[1], rowSums(dth[[nl]])*h, type = 'n', 
     xlab = expression(theta[1]), 
     ylab = expression(pi[theta[1]](theta[1]~"|"~lambda)))
for(i in 1:nl) {
  lines(th0s + theta0[1], 
        rowSums(dth[[i]])*h, lty = i)
}
plot(th0s + theta0[2], colSums(dth[[nl]])*h, type = 'n', 
     xlab = expression(theta[2]), 
     ylab = expression(pi[theta[2]](theta[2]~"|"~lambda)))
for(i in 1:nl) {
  lines(th0s + theta0[2], 
        colSums(dth[[i]])*h, lty = i)
}
legend('topright', as.expression(llab), lty = 1:nl, bty = 'n')
```


```{r cprior2, out.width='0.99\\textwidth', fig.width=9, fig.height=3}
par(mfrow = c(1, 3), mar = c(0,1,0,2))
for(i in 1:nl) {
  bkz <- exp(seq(log(min(dth[[i]]))-1e-9, 
                 log(max(dth[[i]]))+1e-9, length = 65))
   surf3D(matrix(th0x[1, ], nth0),
          matrix(th0x[2, ], nth0), dth[[i]], 
          breaks = bkz, colkey = !TRUE,
          xlab = '', ylab  = '', zlab = '',
          box = TRUE, bty = "b")
  text(-0.15, -0.43, expression(theta[1]), adj = 1)
  text(0.28, -0.37, expression(theta[2]), adj = 1)
  legend("topleft", bty = "n", as.expression(llab[[i]]), cex = 2)
}

```

To evaluate $\pi_{\brho}(\brho|\lambda)$ 
we need to add the Jacobian for the transformation 
from $\btheta$ to $\brho$.
This mapping is not given by two univariate transformation, 
but bivariate. 


# Example with $m=3$

```{r exm3, echo=TRUE, results='markup'}
## the model as 'graphpcor'
gph_model <- graphpcor(
    graph = matrix(1, 3, 3) ## dense
)

## base model (base correlation matrix)
C0 <- matrix(c(1,0.5,0.5,0.5,1,0.4,0.5,0.4,1), nrow = 3)

## Hessian around a base model
Ith0 <- hessian(gph_model, C0)
str(Ith0)

### check: evaluating theta0 gives the base model
th0 <- attr(Ith0, 'theta')
vcov(gph_model, theta = th0)

## H^0.5: "SQRT" of the Hessian of KLD around the base model
h.5 <- attr(Ith0, 'h.5')

## its determinant
(dh.5 <- prod(attr(Ith0, 'decomposition')$values^0.5))


## set up a grid of theta around theta0
h <- 0.01
th0s <- seq(-1.5+h/2, 1.5-h/2, h)
th0x <- t(expand.grid(th0s, th0s, th0s)) ## each column ... 
ncol(th0x)/1e6 ## size of the grid

## eval, in R
R <- sqrt(colSums( (h.5 %*% th0x)^2 ))
Sr <- 2 * (pi^(3/2)) * R^2 / gamma(3/2)

param <- 5
dthR <- param * exp(-param * R) * dh.5 / Sr

## define a `cgeneric`
c_model <- cgeneric(
    model = gph_model,
    lambda = param,
    base = C0)

## check: evaluate at the base model
solve(prec(c_model, theta = th0))

## eval, in C (through 'cgeneric' query method in INLAtools)
dth <- exp(
    prior(
        c_model, 
        theta = rbind(th0x[1,] + th0[1],
                      th0x[2,] + th0[2],
                      th0x[3,] + th0[3])
    )
)

all.equal(dthR, dth)

## organize into an array
d3th <- array(dth, dim = rep(length(th0s), 3))

d1th <- apply(d3th * h*h, 1, sum)
d2th <- apply(d3th * h*h, 2, sum)
d3th <- apply(d3th * h*h, 3, sum)

c(sum(d1th*h), sum(d2th*h), sum(d3th*h))

par(mfrow = c(1, 3), bty = 'n')
plot(th0s + th0[1], d1th); abline(v = th0[1])
plot(th0s + th0[2], d2th); abline(v = th0[2])
plot(th0s + th0[3], d3th); abline(v = th0[3])
```

# Example with $m>3$


# References

