---
title: 'Tree based latent variables model'
author: "Anna Freni-Sterrantino, Denis Rustand, Janet van Niekerk, Elias T Krainski, and H\\aa vard Rue"
date: "Started in November 2023, updated in `r format(Sys.time(), '%B, %Y')`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Tree based latent variables model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
- \usepackage{lscape}
- \usepackage{multicol}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(graphpcor)
```


This material detail the model properties of the 
graphical approach to model correlations proposed in @sterrantino2025graphical.
We first formalize the specific kind of graph used
to define the model and provides an intuitive visualization.
We also formalize simple rules for writing the joint distribution
for all the variables in the model. 
This gives also the precision matrix for the full model and the 
covariance for the variables of interest.
We use some examples as illustration.
We recommend to look at the 'treepcor' vignette
on the implementation details.

## The directed tree definition

A directed acyclic graph is defined from a set of nodes and 
directed edges linking them. 
Because the edges are directed, there is no closed loop on it.
A tree is a graph where there is only one possible path 
between a pair of nodes.
Furthermore, the edges of a tree are directed edges.
This also impose a hierarchy on the nodes.

We consider a particular tree definition 
represent two kind of variables and the
directed edges (arrows) to represent conditional distributions.
We classify the nodes as parent or children.
The nodes classified as children are leafs of the tree, 
which are nodes with an ancestor (parent) 
but with no children.
The nodes classified as parent will always have children nodes,
and some may have parent but will still be classified as parent 
because they have children.

The $k$ variables of interest, 
those for we want to model their correlation,
will be represented by the children nodes and
labeled as $c_i$, $i\in\{1,\ldots,m\}$. 
The parent nodes are labeled as $p_j$, $j\in\{1,\ldots,k\}$.
The parent $p_1$ is the main parent, with no parent, 
which is the root of the tree.
We have some examples in Figure 1 bellow.

```{r graphs, echo = FALSE, fig.width = 9, fig.height = 4, out.width="79%", fig.align = 'center', fig.cap = "Graphs representing different models. A model with one parent, (A), a model with two parents (B), and two different models with three parents (C) and (D)."}
t1 <- treepcor(p1 ~ c1 + c2 + c3)
t2 <- treepcor(p1 ~ p2 + c1 + c2, 
               p2 ~ c3 + c4)
t3a <- treepcor(p1 ~ p2 + p3 + c1 + c2, 
                p2 ~ c3 + c4, 
                p3 ~ c5 + c6)
t3b <- treepcor(p1 ~ p2 + c1 + c2, 
                p2 ~ p3 + c3 + c4, 
                p3 ~ c5 + c6)
par(mfrow = c(2, 2), mar = c(0, 0, 0, 0))
plot(t1)
legend("topleft", "(A)", bty = "n")
plot(t2)
legend("topleft", "(B)", bty = "n")
plot(t3a)
legend("topleft", "(C)", bty = "n")
plot(t3b)
legend("topleft", "(D)", bty = "n")
```

The tree based model induces a
non-zero covariance between children as long as there 
are arrows linking them.
The strength of the correlation depends on how many parents, 
and their hierarchy, there is in the path linking them.
Let us represent the correlation between $a$ and $b$ as C(a,b).
In the model defined with graph (A), 
|C($c_1$,$c_2$)| = |C($c_1$,$c_3$)| = |C($c_2$,$c_3$)|.
From graph (B) 
|C($c_1$,$c_3$)| = |C($c_1,c_4$)| = |C($c_2$,$c_3$)| = 
|C($c_2$,$c_4$)| $\leq$ |C($c_1$,$c_2$)| $\leq$ |C($c_3$,$c_4$)|.

With the models defined from the graphs (C) and (D) we have 
|C($c_1$,$c_2$)|$\leq$|C($c_3$,$c_4$)|, 
|C($c_1$,$c_2$)|$\leq$|C($c_5$,$c_6$)|, and
|C($c_1$,$c_3$)|=|C($c_1$,$c_4$)|=
|C($c_2$,$c_3$)|=|C($c_2$,$c_4$)| and
|C($c_1$,$c_5$)|=|C($c_1$,$c_6$)|=
|C($c_2$,$c_5$)|=|C($c_2$,$c_6$)|.
If we want to not fix the order between |C($c_3$,$c_4$)| 
and |C($c_5$,$c_6$) we can use the model from graph (C), 
whereas model (D) impose |C($c_3$,$c_4$)|$\leq$|C($c_5$,$c_6$)| and
the correlation strength between a children of $p_3$ to a 
children of $p_1$ to be smaller than to a children of $p_2$, e.g.
|C($c_1$,$c_5$)|$\leq$|C($c_3$,$c_5$)|.

## Model definition

We now define a set of rules to propose a model.
We assume a distribution for the main parent, $p_1$,
and conditional distributions for each of the
remaining $k-1$ parents and each of the $m$ children.
The arrows represent each of these conditional distributions.
Since we assume Gaussian distributions, 
we just need to define the (conditional) mean and variance.

* **1st rule**: For the main parent $p_1$, 
the mean is zero and the variance is $\gamma_1^2$. 

* **2nd rule**: The conditional mean for $p_j$, $j=2$,...$k$, 
is its parent and the conditional variance is $\gamma_j^2$.

* **3rd rule**: The conditional mean of $c_i$, $i=1$, ..., $m$, 
is $s_i$ times its parent and unit conditional variance, 
where $s_i$ is assumed known, either "-1" or "+1".

We have $p$ parameters, one for each parent.
These conditional distributions build up a joint one 
that will have a covariance matrix. 
Since we are primarily interested in the correlations, 
we will scale it by the implied variances and obtaining 
a correlation matrix. 
The marginal variances for each children are assigned 
after this scaling.
As we will usually have $k<m$, 
this represents a reduction on the number of 
parameters, from $m(m+1)/2$ to $k+m$.

As an illustration, consider the graph (A) from Figure 1.
The joint distribution for $\{c_1, c_2, c_3, p_1\}$ is 
\[
\pi(c_1, c_2, c_3, p_1) \propto \mathrm{exp} 
  \left\{-\frac{1}{2} \left[ \frac{p_1^2}{\gamma_1^2} + \frac{(c_1-s_1p_1)^2}{s_1^2} + \frac{(c_2-s_2p_1)^2}{s_2^2} + \frac{(c_3-s_3p_1)^2}{s_3^2} \right] \right\}.
\]
From this set of conditional distributions we obtain the precision matrix as 
\[
\left[ \begin {array}{cccc} 
{{\it s_1}}^{-2}&0&0&-{{\it s_1}}^{-1}\\ 
0&{{\it s_2}}^{-2}&0&-{{\it s_2}}^{-1}\\ 
0&0&{{\it s_3}}^{-2}&-{{\it s_3}}^{-1}\\ 
-{{\it s_1}}^{-1}&-{{\it s_2}}^{-1}&-{{\it s_3}}^{-1}&{\gamma_1}^{-2}+3
\end {array} \right].
\]

The joint distribution for $\{c_1, c_2, c_3, c_4, p_1, p_2\}$, 
from the DAG (B) in Figure 1, is proportional to
\[ 
\mathrm{exp} 
  \left\{-\frac{1}{2} \left[ \frac{p_1^2}{\gamma_1^2} + \frac{(p_2-p_1)^2}{\gamma_2^2} +
  \frac{(c_1-s_1p_1)^2}{s_1^2} + \frac{(c_2-s_2p_1)^2}{s_2^2} +
  \frac{(c_3-s_3p_2)^2}{s_3^2} + \frac{(c_4-s_4p_2)^2}{s_4^2} \right] \right\}.
\]
The precision matrix is 
\[
\left[ \begin {array}{cccccc} 
{{\it s_1}}^{-2}&0&0&0&-{{\it s_1}}^{-1}&0\\ 
0&{{\it s_2}}^{-2}&0&0&-{{\it s_2}}^{-1}&0\\ 
0&0&{{\it s_3}}^{-2}&0&0&-{{\it s_3}}^{-1}\\ 
0&0&0&{{\it s_4}}^{-2}&0&-{{\it s_4}}^{-1}\\ 
-{{\it s_1}}^{-1}&-{{\it s_2}}^{-1}&0&0&
{\gamma_1}^{-2}+{\gamma_2}^{-2}+2&-{\gamma_2}^{-2}\\ 
0&0&-{{\it s_3}}^{-1}&-{{\it s_4}}^{-1}&-{\gamma_2}^{-2}&{\gamma_2}^{-2}+2
\end {array} \right] 
\]

The models represented by graphs (C) and (D) in Figure 1 have both the same
number of children and parents. 
The only difference is that parent $p_3$ is children to parent $p_1$ in (C) 
whereas in (D) $p_3$ is children to parent $p_2$.
For the case when $p_1$ is parent to $p_3$ the last three rows of the precision matrix are 
\[
\left[ \begin {array}{ccccccccc} 
%% {{\it s_1}}^{-2}&0&0&0&0&0&-{{\it s_1}}^{-1}&0&0\\ 
%% 0&{{\it s_2}}^{-2}&0&0&0&0&-{{\it s_2}}^{-1}&0&0\\ 
%% 0&0&{{\it s_3}}^{-2}&0&0&0&0&-{{\it s_3}}^{-1}&0\\ 
%% 0&0&0&{{\it s_4}}^{-2}&0&0&0&-{{\it s_4}}^{-1}&0\\ 
%% 0&0&0&0&{{\it s_5}}^{-2}&0&0&0&-{{\it s_5}}^{-1} \\ 
%% 0&0&0&0&0&{{\it s_6}}^{-2}&0&0&-{{\it s_6}}^{-1}\\ 
-{{\it s_1}}^{-1}&-{{\it s_2}}^{-1}&0&0&0&0&
 {\gamma_1}^{-2}+{\gamma_2}^{-2}+{\gamma_3}^{-2} +2 &
-{\gamma_2}^{-2}&-{\gamma_3}^{-2}\\ 
0&0&-{{\it s_3}}^{-1}&-{{\it s_4}}^{-1}&0&0&
 -{\gamma_2}^{-2}&{\gamma_2}^{-2}+2&0\\ 
 0&0&0&0&-{{\it s_5}}^{-1}&-{{\it s_6}}^{-1}&
 -{\gamma_3}^{-2}&0&{\gamma_3}^{-2}+2
 \end {array} \right]
\]
and the case when $p_2$ is parent to $p_3$ the last three rows of the precision matrix are
\[ 
\left[ \begin {array}{ccccccccc} 
-{{\it s_1}}^{-1}&-{{\it s_2}}^{-1}&0&0&0&0&{\gamma_1}^{-2}+{\gamma_2}^{-2}+2&-{\gamma_2}^{-2}
&0\\ 0&0&-{{\it s_3}}^{-1}&-{{\it s_4}}^{-1}&0&0&-{\gamma_2}^{-2}&{\gamma_2}^{-2}+{\gamma_3}^{-2}+2&-{\gamma_3}^{-2}\\ 0&0&0&0&-{{\it s_5}}^{-1}&-{
{\it s_6}}^{-1}&0&-{\gamma_3}^{-2}&{\gamma_3}^{-2}+2\end {array} \right] 
 \]
and the first $6$ rows are equal the ones from the previous case.

## Model properties

The conditional distributions defines a precision matrix 
for all the $m+k$ variables in the model and a joint distribution.
Therefore we can obtain the implied covariance matrix 
to learn the marginal properties, and correlations as well.
Since our primarily interest is in the correlation between the children variables, we consider its first $m$ rows and columns.

The model defined from the DAG (A) in Figure 1 for 
$\{c_1, c_2, c_3, p_1\}$ induces a covariance matrix equals to
\[ 
\left[ \begin {array}{cccc}  
( 1+{\gamma_1}^{2} ) {{\it s_1}}^{2} & {\it s_2}{\it s_1}{\gamma_1}^{2}&
{\it s_3}{\it s_1}{\gamma_1}^{2} & {\it s_1}{\gamma_1}^{2} \\ 
{\it s_2}{\it s_1}{\gamma_1}^{2} & ( 1+{\gamma_1}^{2} ) {{\it s_2}}^{2}& 
{\it s_3}{\it s_2}{\gamma_1}^{2} & {\it s_2}{\gamma_1}^{2} \\ 
{\it s_3}{\it s_1}{\gamma_1}^{2} &{\it s_3}{\it s_2}{\gamma_1}^{2}& 
( 1+{\gamma_1}^{2} ) {{\it s_3}}^{2} & {\it s_3}{\gamma_1}^{2} \\ 
{\it s_1}{\gamma_1}^{2} & {\it s_2}{\gamma_1}^{2} &
{\it s_3}{\gamma_1}^{2}&{\gamma_1}^{2}
\end {array} \right] 
\]
and the C($c_i,c_j$) = sign($s_is_j$)$\gamma_1^2/(1+\gamma_1^2)$, for $i,j\in\{1,2,3\}$ and $i\neq j$.

The model defined from the DAG (B) in Figure 1 gives the covariance matrix 
for $\{c_1, c_2, c_3, c_4, p_1, p_2\}$ as
\[ 
\left[ \begin {array}{cccccc}  \left( 1+{\gamma_1}^{2} \right) {{\it s_1}}^{2}&{\it s_2}\,{\gamma_1}^{2}{\it s_1}&{\it s_3}\,{\gamma_1}^{2}{\it s_1}&{\it s_4}\,{\gamma_1}^{2}{\it 
s_1}&{\gamma_1}^{2}{\it s_1}&{\gamma_1}^{2}{\it s_1}\\ {\it s_2}\,{\gamma_1}^{2}{\it s_1}& \left( 1+{\gamma_1}^{2} \right) {{\it s_2}}^{2}&{\it s_3}\,{\gamma_1
}^{2}{\it s_2}&{\it s_4}\,{\gamma_1}^{2}{\it s_2}&{\gamma_1}^{2}{\it s_2}&{\gamma_1}^{2}{\it s_2}\\ {\it s_3}\,{\gamma_1}^{2}{\it s_1}&{\it s_3}\,{\gamma_1}^{
2}{\it s_2}& \left( 1+{\gamma_2}^{2}+{\gamma_1}^{2} \right) {{\it s_3}}^{2}&{\it s_4}\, \left( {\gamma_2}^{2}+{\gamma_1}^{2} \right) {\it s_3}&{\gamma_1}^{2}{\it s_3}& \left( {{
\it \gamma_2}}^{2}+{\gamma_1}^{2} \right) {\it s_3}\\ {\it s_4}\,{\gamma_1}^{2}{\it s_1}&{\it s_4}\,{\gamma_1}^{2}{\it s_2}&{\it s_4}\, \left( {\gamma_2}^{2}+{{\it 
\gamma_1}}^{2} \right) {\it s_3}& \left( 1+{\gamma_2}^{2}+{\gamma_1}^{2} \right) {{\it s_4}}^{2}&{\gamma_1}^{2}{\it s_4}& \left( {\gamma_2}^{2}+{\gamma_1}^{2} \right) {\it s_4}
\\ {\gamma_1}^{2}{\it s_1}&{\gamma_1}^{2}{\it s_2}&{\gamma_1}^{2}{\it s_3}&{\gamma_1}^{2}{\it s_4}&{\gamma_1}^{2}&{\gamma_1}^{2}\\ {\gamma_1
}^{2}{\it s_1}&{\gamma_1}^{2}{\it s_2}& \left( {\gamma_2}^{2}+{\gamma_1}^{2} \right) {\it s_3}& \left( {\gamma_2}^{2}+{\gamma_1}^{2} \right) {\it s_4}&{\gamma_1}^{2}&{\gamma_2}
^{2}+{\gamma_1}^{2}\end {array} \right] 
\]
from what the marginal properties can be learn. 
The upper triangle of the children correlation matrix is
\[
\left[ \begin {array}{ccc} 
\frac{\textrm{sign}(s_1s_2){\gamma_1}^{2}}{{( {\gamma_1}^{2}+1 )}} & 
\frac{\textrm{sign}(s_1 s_3){\gamma_1}^{2}} {\sqrt{( {\gamma_1}^{2}+1 )} \sqrt { ( 1+{\gamma_2}^{2}+{\gamma_1}^{2} )}} & 
\frac{\textrm{sign}(s_1 s_4){\gamma_1}^{2}} {\sqrt{( {\gamma_1}^{2}+1 )} \sqrt { ( 1+{\gamma_2}^{2}+{\gamma_1}^{2} )}}\\ 
& \frac{\textrm{sign}(s_2 s_3){\gamma_1}^{2}} {\sqrt{( {\gamma_1}^{2}+1 )} \sqrt { ( 1+{\gamma_2}^{2}+{\gamma_1}^{2} )}} & 
\frac{\textrm{sign}(s_2 s_4){\gamma_1}^{2}} {\sqrt{( {\gamma_1}^{2}+1 )} \sqrt { ( 1+{\gamma_2}^{2}+{\gamma_1}^{2} )}}\\
 &  & \frac{{\textrm{sign}(s_3 s_4) ( {\gamma_2}^{2}+{\gamma_1}^{2}) }}{
{( 1+{\gamma_2}^{2}+{\gamma_1}^{2})}}
\end{array} \right] 
\]
where C($s_i,s_j$) = sign($s_is_j$)$\frac{\gamma_1^2}{(\gamma_1^2+1)(1+\gamma_1^2+\gamma_2^2)}$ 
for $i\in\{1,2\}$ and $j\in\{3,4\}$, which defines a block with the same absolute value 
and is the main structure that this model induced for the correlation matrix.

In the model given from graph (C), when $p_1$ is parent to $p_3$,
the covariance matrix of $\{c_1,...,c_6\}$ is 
\[ \scriptsize 
\left[ \begin {array}{cccccc}  \left( 1+{\gamma_1}^{2} \right) {{\it s_1}}^{2}&{\it s_2}\,{\gamma_1}^{2}{\it s_1}&{\it s_3}\,{\gamma_1}^{2}{\it s_1}&{\it s_4}\,{\gamma_1}^{2}{\it 
s_1}&{\it s_5}\,{\gamma_1}^{2}{\it s_1}&{\it s_6}\,{\gamma_1}^{2}{\it s_1}\\ {\it s_2}\,{\gamma_1}^{2}{\it s_1}& \left( 1+{\gamma_1}^{2} \right) {{\it s_2}}^{2}
&{\it s_3}\,{\gamma_1}^{2}{\it s_2}&{\it s_4}\,{\gamma_1}^{2}{\it s_2}&{\it s_5}\,{\gamma_1}^{2}{\it s_2}&{\it s_6}\,{\gamma_1}^{2}{\it s_2}\\ {\it s_3}\,{{
\it \gamma_1}}^{2}{\it s_1}&{\it s_3}\,{\gamma_1}^{2}{\it s_2}& \left( 1+{\gamma_2}^{2}+{\gamma_1}^{2} \right) {{\it s_3}}^{2}&{\it s_4}\, \left( {\gamma_2}^{2}+{\gamma_1}^{2}
 \right) {\it s_3}&{\it s_5}\,{\gamma_1}^{2}{\it s_3}&{\it s_6}\,{\gamma_1}^{2}{\it s_3}\\ {\it s_4}\,{\gamma_1}^{2}{\it s_1}&{\it s_4}\,{\gamma_1}^{2}{\it s_2
}&{\it s_4}\, \left( {\gamma_2}^{2}+{\gamma_1}^{2} \right) {\it s_3}& \left( 1+{\gamma_2}^{2}+{\gamma_1}^{2} \right) {{\it s_4}}^{2}&{\it s_5}\,{\gamma_1}^{2}{\it s_4}&{\it s_6}
\,{\gamma_1}^{2}{\it s_4}\\ {\it s_5}\,{\gamma_1}^{2}{\it s_1}&{\it s_5}\,{\gamma_1}^{2}{\it s_2}&{\it s_5}\,{\gamma_1}^{2}{\it s_3}&{\it s_5}\,{\gamma_1}^{2}
{\it s_4}& \left( 1+{\gamma_3}^{2}+{\gamma_1}^{2} \right) {{\it s_5}}^{2}&{\it s_6}\,{\it s_5}\, \left( {\gamma_3}^{2}+{\gamma_1}^{2} \right) \\ {\it s_6}\,{{\it 
\gamma_1}}^{2}{\it s_1}&{\it s_6}\,{\gamma_1}^{2}{\it s_2}&{\it s_6}\,{\gamma_1}^{2}{\it s_3}&{\it s_6}\,{\gamma_1}^{2}{\it s_4}&{\it s_6}\,{\it s_5}\, \left( {\gamma_3}^{2}+{{\it 
\gamma_1}}^{2} \right) & \left( 1+{\gamma_3}^{2}+{\gamma_1}^{2} \right) {{\it s_6}}^{2}\end {array} \right] 
\]
where the variances and covariance for $c_5$ and $c_6$ does not depends on $\gamma_2^2$.

For the case when $p_2$ is parent to $p_3$ the covariance of $c_1$, ..., $c_6$ is 
\[ \tiny
\left[ \begin {array}{cccccc}  \left( 1+{\gamma_1}^{2} \right) {{\it s_1}}^{2}&{\it s_2}\,{\gamma_1}^{2}{\it s_1}&{\it s_3}\,{\gamma_1}^{2}{\it s_1}&{\it s_4}\,{\gamma_1}^{2}{\it 
s_1}&{\it s_5}\,{\gamma_1}^{2}{\it s_1}&{\it s_6}\,{\gamma_1}^{2}{\it s_1}\\ {\it s_2}\,{\gamma_1}^{2}{\it s_1}& \left( 1+{\gamma_1}^{2} \right) {{\it s_2}}^{2}
&{\it s_3}\,{\gamma_1}^{2}{\it s_2}&{\it s_4}\,{\gamma_1}^{2}{\it s_2}&{\it s_5}\,{\gamma_1}^{2}{\it s_2}&{\it s_6}\,{\gamma_1}^{2}{\it s_2}\\ {\it s_3}\,{{
\it \gamma_1}}^{2}{\it s_1}&{\it s_3}\,{\gamma_1}^{2}{\it s_2}& \left( 1+{\gamma_2}^{2}+{\gamma_1}^{2} \right) {{\it s_3}}^{2}&{\it s_4}\, \left( {\gamma_2}^{2}+{\gamma_1}^{2}
 \right) {\it s_3}&{\it s_5}\, \left( {\gamma_2}^{2}+{\gamma_1}^{2} \right) {\it s_3}&{\it s_6}\, \left( {\gamma_2}^{2}+{\gamma_1}^{2} \right) {\it s_3}\\ {\it 
s_4}\,{\gamma_1}^{2}{\it s_1}&{\it s_4}\,{\gamma_1}^{2}{\it s_2}&{\it s_4}\, \left( {\gamma_2}^{2}+{\gamma_1}^{2} \right) {\it s_3}& \left( 1+{\gamma_2}^{2}+{\gamma_1}^{2}
 \right) {{\it s_4}}^{2}&{\it s_5}\, \left( {\gamma_2}^{2}+{\gamma_1}^{2} \right) {\it s_4}&{\it s_6}\, \left( {\gamma_2}^{2}+{\gamma_1}^{2} \right) {\it s_4}\\ {
\it s_5}\,{\gamma_1}^{2}{\it s_1}&{\it s_5}\,{\gamma_1}^{2}{\it s_2}&{\it s_5}\, \left( {\gamma_2}^{2}+{\gamma_1}^{2} \right) {\it s_3}&{\it s_5}\, \left( {\gamma_2}^{2}+{{\it 
\gamma_1}}^{2} \right) {\it s_4}& \left( 1+{\gamma_3}^{2}+{\gamma_2}^{2}+{\gamma_1}^{2} \right) {{\it s_5}}^{2}&{\it s_6}\,{\it s_5}\, \left( {\gamma_3}^{2}+{\gamma_2}^{2}+{\gamma_1}
^{2} \right) \\ {\it s_6}\,{\gamma_1}^{2}{\it s_1}&{\it s_6}\,{\gamma_1}^{2}{\it s_2}&{\it s_6}\, \left( {\gamma_2}^{2}+{\gamma_1}^{2} \right) {\it s_3}&{\it s_6
}\, \left( {\gamma_2}^{2}+{\gamma_1}^{2} \right) {\it s_4}&{\it s_6}\,{\it s_5}\, \left( {\gamma_3}^{2}+{\gamma_2}^{2}+{\gamma_1}^{2} \right) & \left( 1+{\gamma_3}^{2}+{\gamma_2
}^{2}+{\gamma_1}^{2} \right) {{\it s_6}}^{2}\end {array} \right] 
\]
where the variances and covariance for $c_5$ and $c_6$ now depends on $\gamma_2^2$. 

For the correlations, we now compare C($c_i,c_j$), for $i\neq j$ and $i,j\in\{3,4,5,6\}$, 
with these two different models. 
The upper triangle of the correlation matrix for $c_3$, ..., $c_6$
obtained with the model where $p_1$ is parent to $p_3$ is 
\[
\left[ \begin {array}{ccc} 
\textrm{sign}(s_3s_4)\frac{\gamma_1^2+\gamma_2^2}{1+\gamma_1^2+\gamma_2^2} & 
\textrm{sign}(s_3s_5)\frac{\gamma_1^2}{\sqrt{(1+\gamma_1^2)(1+\gamma_1^2+\gamma_3^2)}} &
\textrm{sign}(s_3s_6)\frac{\gamma_1^2}{\sqrt{(1+\gamma_1^2)(1+\gamma_1^2+\gamma_3^2)}} \\
 & \textrm{sign}(s_4s_5)\frac{\gamma_1^2}{\sqrt{(1+\gamma_1^2)(1+\gamma_1^2+\gamma_3^2)}} &
\textrm{sign}(s_4s_6)\frac{\gamma_1^2}{\sqrt{(1+\gamma_1^2)(1+\gamma_1^2+\gamma_3^2)}} \\
 & & \textrm{sign}(s_5s_6)\frac{\gamma_1^2 + \gamma_3^2}{\sqrt{(1+\gamma_1^2+\gamma_3^2)}} \\
\end{array} \right],
\]
and with the model when $p_2$ is parent to $p_3$ it is
\[
\left[ \begin {array}{ccc} 
\textrm{sign}(s_3s_4)\frac{\gamma_1^2+\gamma_2^2}{1+\gamma_1^2+\gamma_2^2} & 
\textrm{sign}(s_3s_5)\frac{\gamma_1^2+\gamma_2^2}{\sqrt{(1+\gamma_1^2+\gamma_2^2)(1+\gamma_1^2+\gamma_3^2)}} &
\textrm{sign}(s_3s_6)\frac{\gamma_1^2+\gamma_2^2}{\sqrt{(1+\gamma_1^2+\gamma_2^2)(1+\gamma_1^2+\gamma_3^2)}} \\
 & \textrm{sign}(s_4s_5)\frac{\gamma_1^2+\gamma_2^2}{\sqrt{(1+\gamma_1^2+\gamma_2^2)(1+\gamma_1^2+\gamma_3^2)}} &
\textrm{sign}(s_4s_6)\frac{\gamma_1^2+\gamma_2^2}{\sqrt{(1+\gamma_1^2+\gamma_2^2)(1+\gamma_1^2+\gamma_3^2)}} \\
 & & \textrm{sign}(s_5s_6)\frac{\gamma_1^2+\gamma_2^2+\gamma_3^2}{1+\gamma_1^2+\gamma_2^2+\gamma_3^2} \\
\end{array} \right].
\]
Comparing these results, when $p_1$ is parent to $p_3$ there is no contribution 
from $\gamma_2^2$ in C($c_5,c_6$) and neither in C($c_i,c_j$), for $i\in\{3,4\}$,
whereas when $p_2$ is parent to $p_3$ there is.

From these examples we can list a set of properties as follows.

* **Property 1**: The marginal variance of $p_i$ is $\gamma_i^2$ plus the variance of its parent.
   * **Property 1a**: For $p_1$ it is equal its conditional variance, $\gamma_1^2$.
   * **Property 1b**: The marginal variance of $p_i$ is $\gamma_i^2$ plus 
   the conditional variance of its ancestors.
* **Property 2**: The marginal variance of $c_i$ is equal $s_i^2(1+V)$,
where $V$ is the variance of its parent.
* **Property 3**: The covariance between $c_i$ and $c_j$ is equal $s_is_jR$, 
where $R$ is the variance of the top indexed parent in the path linking them. 
   * **Property 3a**: If $c_i$ and $c_j$ have a common parent, 
   $R$ is the variance of their parent.
* **Property 4**: C($c_i,c_j$)=sign($s_is_j$)R/$\sqrt{\gamma_{c_i}\gamma_{c_j}}$, 
where $\gamma_{c_i}$ and $\gamma_{c_j}$ are the marginal variances of $c_i$ and $c_j$, respectively. 
   * **Property 4a**: If $c_i$ and $c_j$ have a common parent, 
   the correlation is sign($s_is_j$)$V/(1+V)$ where $V$ is the variance of their parent.
* **Property 5**: Let $k_0$ be the number of parents with no children variable as children, 
   $k_1$ be the number of parents with one children variable as children,
   and $k_2$ be the number of parents with at least two children variables as children.
   The number of distinct absolute correlation values is $k_2+(k_1-k_0)(k_1-k_0-1)/2$.
   * **Property 5a**: If each parent variable has only one children variable as children,
   this number is $k(k-1)/2$.
   * **Property 5b**: The maximum number of distinct absolute correlation values is 
$k(k+1)/2$, achieved if each parent variable is parent to at least two children variables.

This intuitive modeling approach gives way to build a
structured correlation matrix.
The number of distinct absolute correlation values is $k(k+1)/2$.
The sign assumed for each arrow ending at a children variable
allows the full range of the correlation values, from $-1$ to $1$.
Prior distributions can be considered for the marginal variances 
and for the correlation matrix separately.

## A further example

We now consider more variables and use two models, as shown in Figure 2 bellow.
The difference in on who is the parent to $p_4$. 
We will use it to further illustrate the **Property 2a**.

```{r t4, echo = FALSE,  fig.width = 9, fig.height = 4, out.width="69%", fig.align = 'center', fig.cap = "Two different models for the same number of children and parents."}
t4a <- treepcor(
  p1 ~ p2 + p3 + c1 + c2, 
  p2 ~ p4 + c3 + c4, 
  p3 ~ p5 + c5 + c6, 
  p4 ~ c7 + c8, p5 ~ c9 + c10)
t4b <- treepcor(p1 ~ p2 + p3 + c1 + c2, 
  p2 ~ c3 + c4, p3 ~ p4 + p5 + c5 + c6, 
  p4 ~ c7 + c8, p5 ~ c9 + c10)
par(mfrow = c(1, 2), mar = c(0, 0, 0, 0))
plot(t4a)
legend("topleft", "(E)", bty = "n")
plot(t4b)
legend("topleft", "(F)", bty = "n")
```

The model for the DAG in (E) gives a precision matrix whose last five columns are
\[ \scriptsize
\left[ \begin {array}{ccccc} -{{\it s_1}}^{-1}&0&0&0&0\\ -{{\it s_2}}^{-1}&0&0&0&0\\ 0&-{{\it s_3}}^{-1}&0&0&0\\ 0&-{{\it s_4}}^{-1}&0&0
&0\\ 0&0&-{{\it s_5}}^{-1}&0&0\\ 0&0&-{{\it s_6}}^{-1}&0&0\\ 0&0&0&-{{\it s_7}}^{-1}&0\\ 0&0&0&-{{\it s_8}}^{-1}&0
\\ 0&0&0&0&-{{\it s_9}}^{-1}\\ 0&0&0&0&-{{\it s_{10}}}^{-1}\\ {\gamma_1}^{-2}+2+{\gamma_2}^{-2}+{\gamma_3}^{-2}&-{\gamma_2}^{-2}&-{{
\it \gamma_3}}^{-2}&0&0\\ -{\gamma_2}^{-2}&{\gamma_2}^{-2}+2+{\gamma_4}^{-2}&0&-{\gamma_4}^{-2}&0\\ -{\gamma_3}^{-2}&0&{\gamma_3}^{-2}+2+{\gamma_5}^
{-2}&0&-{\gamma_5}^{-2}\\ 0&-{\gamma_4}^{-2}&0&{\gamma_4}^{-2}+2&0\\ 0&0&-{\gamma_5}^{-2}&0&{\gamma_5}^{-2}+2\end {array} \right] 
\]
while the last five columns for the precision matrix for the model for DAG in (F) are
\[ \scriptsize  
\left[ \begin {array}{ccccc} 
-{{\it s_1}}^{-1}&0&0&0&0\\ -{{\it s_2}}^{-1}&0&0&0&0\\ 0&-{{\it s_3}}^{-1}&0&0&0\\ 0&-{{\it s_4}}^{-1}&0&0
&0\\ 0&0&-{{\it s_5}}^{-1}&0&0\\ 0&0&-{{\it s_6}}^{-1}&0&0\\ 0&0&0&-{{\it s_7}}^{-1}&0\\ 0&0&0&-{{\it s_8}}^{-1}&0
\\ 0&0&0&0&-{{\it s_9}}^{-1}\\ 0&0&0&0&-{{\it s_{10}}}^{-1}\\ {\gamma_1}^{-2}+2+{\gamma_2}^{-2}+{\gamma_3}^{-2}&-{\gamma_2}^{-2}&-{{
\it \gamma_3}}^{-2}&0&0\\ -{\gamma_2}^{-2}&{\gamma_2}^{-2}+2&0&0&0\\ -{\gamma_3}^{-2}&0&{\gamma_3}^{-2}+2+{\gamma_4}^{-2}+{\gamma_5}^{-2}&-{\gamma_4
}^{-2}&-{\gamma_5}^{-2}\\ 0&0&-{\gamma_4}^{-2}&{\gamma_4}^{-2}+2&0\\ 0&0&-{\gamma_5}^{-2}&0&{\gamma_5}^{-2}+2
\end {array} \right]
\]

To illustrate property about the top parent on a path, we consider 
the upper triangle of the correlation matrix for the last four children variables, $c_7,...,c_{10}$.
For the model defined from (E) we have it as
\[ 
\left[ \begin {array}{ccc} 
 {\frac { \textrm{sign}({\it s_7 s_8}) ( {\gamma_1}^{2}+{\gamma_2}^{2}+{\gamma_4}^{2} )}{( 1+{\gamma_1}^{2}+{\gamma_2}^{2}+{\gamma_4}^{2} )}} &
\frac {\textrm{sign}({\it s_7 s_9}) {\gamma_1}^{2}}{\sqrt { ( 1+{\gamma_1}^{2}+{\gamma_2}^{2}+{\gamma_4}^{2} ) ( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_5}^{2} )}} & 
\frac {\textrm{sign}({\it s_7 s_{10}}) {\gamma_1}^{2}}{\sqrt { ( 1+{\gamma_1}^{2}+{\gamma_2}^{2}+{\gamma_4}^{2} ) ( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_5}^{2} )}} \\ 
 & \frac {\textrm{sign}({\it s_8 s_9}) {\gamma_1}^{2}}{\sqrt { ( 1+{\gamma_1}^{2}+{\gamma_2}^{2}+{\gamma_4}^{2} ) ( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_5}^{2} )}} &
\frac {\textrm{sign}({\it s_8 s_{10}}) {\gamma_1}^{2}}{\sqrt { ( 1+{\gamma_1}^{2}+{\gamma_2}^{2}+{\gamma_4}^{2} ) ( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_5}^{2} )}}\\ 
&  &  \frac { \textrm{sign}({\it s_{9} s_{10}}) ( {\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_5}^{2} ) }{\sqrt { ( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_5}^{2} ) }} \\
\end {array} \right] 
\]
where  C($c_i,c_j$), for $i\in\{7,8\}$ and $j\in\{9,10\}$ 
increase depends only on $\gamma_1^2$ and 
depends on all $\gamma_j^2$ for $j=\{1,\ldots,5\}$.
With the model defined from (F) we have 
\[ \
\left[ \begin {array}{ccc} 
\frac {\textrm{sign}({\it s_7s_8}) ( {\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_4}^{2} ) }{( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_4}^{2} ) } & 
\frac { \textrm{sign}({\it s_7s_9}) ( {\gamma_1}^{2}+{\gamma_3}^{2} ) }{\sqrt { ( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_4}^{2} ) ( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_5}^{2} ) }} &
\frac { \textrm{sign}({\it s_7s_{10}}) ( {\gamma_1}^{2}+{\gamma_3}^{2} ) }{\sqrt { ( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_4}^{2} ) ( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_5}^{2} ) }} \\ 
& \frac { \textrm{sign}({\it s_9s_9}) ( {\gamma_1}^{2}+{\gamma_3}^{2} ) }{\sqrt { ( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_4}^{2} ) ( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_5}^{2} ) }}  &
\frac { \textrm{sign}({\it s_8s_{10}}) ( {\gamma_1}^{2}+{\gamma_3}^{2} ) }{\sqrt { ( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_4}^{2} ) ( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_5}^{2} ) }} \\
& & \frac {\textrm{sign}({\it s_9s_{10}}) ( {\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_5}^{2} ) }{ ( 1+{\gamma_1}^{2}+{\gamma_3}^{2}+{\gamma_5}^{2} )}
\end {array} \right] 
\]
where now C($c_i,c_j$), for $i\in\{7,8\}$ and $j\in\{9,10\}$, 
increases if $\gamma_1^2+\gamma_3^2$ (the variance of $p_3$) increases 
and no longer depends on $\gamma_2^2$.



# References

